---
title: "Spruce budworm flight phenology in Maine"
subtitle: "Seasonal dynamics — analysis log"
author: "JN Candau"
date: today
date-format: "D MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk:
    comment: "#>"
---

```{r}
#| label: setup
## Paths are written relative to the project root, not to
## seasonal_dynamics/analysis/.
knitr::opts_knit$set(root.dir = normalizePath(file.path("..", "..")))
```

```{r}
#| label: data
library(dplyr)
library(knitr)
library(ggplot2)

trap_records <- readRDS(file.path("data", "processed", "trap_records.rds"))
season       <- readRDS(file.path("data", "processed", "seasonal",
                                  "season_percentiles.rds"))
```

# Purpose of this document

This is the running log of the *Seasonal dynamics* analysis of spruce budworm
(*Choristoneura fumiferana*) light-trap catches in Maine. Where the *Daily
dynamics* analysis (`daily_dynamics/`) models the catch night by night, this
one treats each locality-year as a single observation — when its flight season
was centred, how long it lasted and how it was shaped — and asks how those
quantities depend on temperature.

The two analyses share one dataset. Its rules are set once, in the data layer
at the project root (`R/01_read_trap_data.R`, `R/02_read_weather_data.R`), and
hold identically here:

- the season of each year runs from the first to the last night on which any
  trap caught a moth that year;
- a `-` cell inside that window is a night the trap ran and caught nothing,
  i.e. a zero;
- an `m` cell is a night whose count was lost, and stays missing;
- one record (Elliotsville Township 1978) is excluded because its only missing
  nights fell at the peak of its season.

The data layer also holds the two development-model runs used below,
BioSIM (`R/03_biosim_phenology.R`) and bayessbw (`R/04_pupal_phenology.R`).
Nothing in this analysis reads an output of `daily_dynamics/`.

# The season-level data {#sec-season}

`seasonal_dynamics/R/01_season_percentiles.R` reduces each record to one row.
The percentiles come from the cumulative catch over the counted nights,
computed with the project's shared helper (`R/functions/phenology.R`), so they
are identical to those used in the daily analysis.

```{r}
#| label: tbl-season-overview
#| tbl-cap: "The season-level response table"
q <- function(x, d = 1) sprintf(paste0("%.", d, "f (%.", d, "f to %.", d, "f)"),
                                stats::median(x), min(x), max(x))
tibble::tibble(
  Quantity = c("Locality-years", "Localities", "Years",
               "Usable for phenology",
               "Median catch date, DOY: median (range)",
               "Central 90% of the season, d",
               "Central 50% of the season, d",
               "Asymmetry, d (positive = longer decline)",
               "Share of the season on the peak night",
               "Records with right-censored nights"),
  Value = c(nrow(season), n_distinct(season$locality), n_distinct(season$year),
            sum(season$usable_phenology),
            q(season$p50), q(season$width90), q(season$width50),
            q(season$asymmetry), q(season$peak_share, 2),
            sum(season$n_cens_right > 0))
) |>
  kable(align = c("l", "r"))
```

```{r}
#| label: fig-season-years
#| fig-cap: "Flight season of every locality-year, by year. Point: median catch date; bar: central 90% of the catch (p05 to p95). Records within a year are ordered by latitude, south to north."
#| fig-width: 9
#| fig-height: 4.5
season |>
  group_by(year) |>
  mutate(offset = (rank(latitude, ties.method = "first") - (n() + 1) / 2) /
           (n() + 1) * 0.8) |>
  ungroup() |>
  ggplot(aes(x = year + offset)) +
  geom_linerange(aes(ymin = p05, ymax = p95), colour = "grey60",
                 linewidth = 0.4) +
  geom_point(aes(y = p50, colour = latitude), size = 1.2) +
  ## plotmath, because the render locale mangles a literal degree sign
  scale_colour_viridis_c(name = expression(Latitude~(degree*N)), option = "D") +
  labs(x = "Year", y = "Day of year") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

The year-to-year spread of the median catch date dwarfs the spread within a
year: the medians of a given year cluster within a few days of each other,
while the year means range over
`r sprintf("%.0f", diff(range(tapply(season$p50, season$year, mean))))` days.
That pattern, already central to the daily analysis, sets the problem for this
one: a seasonal model must first explain the differences between years, and
only then the much smaller differences between localities within a year.

# Predicted pupation dates {#sec-pupation}

```{r}
#| label: pupation-data
pup <- readRDS(file.path("data", "processed", "seasonal", "pupation_dates.rds"))
pup_ok <- filter(pup, biosim_pup_complete)
```

The two development models of the shared data layer predict pupation, not
flight: BioSIM (`R/03_biosim_phenology.R`, mortality switched off) and
bayessbw (`R/04_pupal_phenology.R`). `seasonal_dynamics/R/02_pupation_dates.R`
puts both on the same footing, as a distribution of pupation dates per
locality-year.

For bayessbw this is direct: the model returns the date on which each
simulated individual pupates. BioSIM does not. It reports pupae as a **stock**,
the share of the cohort that is a pupa on a given day, and the median of that
curve is the middle of the pupal period rather than the date of pupation. The
pupation-date distribution is recovered from mass balance. With mortality off,
an individual that has pupated by day *t* is either still a pupa or has
already emerged, so cumulative pupation is the pupae present plus the
cumulative emergence, and its daily increments are the distribution.

The distinction is not cosmetic. The stock median falls a median
`r sprintf("%.1f", median(pup_ok$biosim_stock_p50 - pup_ok$biosim_pup_p50))`
days after the true median pupation date, and with the true date BioSIM
pupates `r sprintf("%.1f", abs(mean(pup_ok$biosim_pup_p50 - pup_ok$bayes_pup_p50)))`
days *earlier* than bayessbw on average, where the stock median had put the two
models within a day of each other. The two remain well correlated across
locality-years (r = `r sprintf("%.2f", cor(pup_ok$biosim_pup_p50, pup_ok$bayes_pup_p50))`).

In `r sum(pup$biosim_pup_lost > 0.5)` locality-years BioSIM's pupae vanish
within a day without emerging, although mortality is switched off. These are
all records whose BioSIM cohort R/03 already flags as incomplete, and they
leave the analysis set below. The pupation dates of the remaining records are
unaffected.

# The median catch date as a function of pupation {#sec-median}

```{r}
#| label: median-data
mvp <- readRDS(file.path("data", "processed", "seasonal", "median_vs_pupation.rds"))
msc <- mvp$scores
mco <- mvp$coefs
mget <- function(sch, dr, mod, col) {
  msc[[col]][msc$scheme == sch & msc$driver == dr & msc$model == mod]
}
cget <- function(dr, mod, term, col = "estimate") {
  mco[[col]][mco$driver == dr & mco$model == mod & mco$term == term]
}
LOYO <- "leave-one-year-out"; LOLO <- "leave-one-locality-out"
```

`seasonal_dynamics/R/03_median_vs_pupation.R` models the observed median
catch date of `r mvp$summary$n_rows` locality-years
(`r mvp$summary$n_years` years, `r mvp$summary$n_locations` localities: usable
for phenology, with a complete BioSIM cohort) from the predicted median
pupation date, separately for each model:

- **lag**: median catch date = pupation + a constant interval;
- **linear**: median catch date = *a* + *b* x pupation;
- **split**: the predicted pupation date is divided into the mean over the
  localities of that year and each locality's departure from it, each with its
  own slope. The year mean uses predicted dates only, so it is available at
  prediction time.

Each is scored twice: leaving out one year at a time, the test for a new year,
and leaving out one locality at a time, the test for a new place. Two
baselines give the scale. *Climatology* predicts the mean date of the training
data. *Same-year* predicts the mean observed date of the other localities that
year; it is not a forecast, since it uses the year being predicted, but it is
the floor that any year-level model faces.

```{r}
#| label: tbl-median-scores
#| tbl-cap: "Skill in predicting the observed median catch date (RMSE and MAE in days)"
msc |>
  filter(!(model == "same-year" & driver == "bayessbw"),
         !(model == "climatology" & driver == "bayessbw")) |>
  mutate(driver = if_else(model %in% c("same-year", "climatology"), "-", driver)) |>
  tidyr::pivot_wider(id_cols = c(driver, model), names_from = scheme,
                     values_from = c(rmse, mae)) |>
  transmute(Driver = driver, Model = model,
            `RMSE, new year` = sprintf("%.2f", .data[[paste0("rmse_", LOYO)]]),
            `MAE, new year` = sprintf("%.2f", .data[[paste0("mae_", LOYO)]]),
            `RMSE, new place` = sprintf("%.2f", .data[[paste0("rmse_", LOLO)]]),
            `MAE, new place` = sprintf("%.2f", .data[[paste0("mae_", LOLO)]])) |>
  arrange(Driver == "-", Driver, `RMSE, new year`) |>
  kable(align = c("l", "l", "r", "r", "r", "r"))
```

**The fixed-lag model is worse than knowing nothing.** Adding a constant
interval to either model's pupation date gives an RMSE of
`r sprintf("%.2f", mget(LOYO, "BioSIM", "lag", "rmse"))` (BioSIM) and
`r sprintf("%.2f", mget(LOYO, "bayessbw", "lag", "rmse"))` days (bayessbw) for a
new year, against `r sprintf("%.2f", mget(LOYO, "BioSIM", "climatology", "rmse"))`
for the climatology.

**The split model shows why.** A year whose predicted pupation is early also
flies early, with a slope of
`r sprintf("%.2f", cget("BioSIM", "split", "pupY"))`
(`r sprintf("%.2f", cget("BioSIM", "split", "pupY", "lo"))` to
`r sprintf("%.2f", cget("BioSIM", "split", "pupY", "hi"))`) for BioSIM and
`r sprintf("%.2f", cget("bayessbw", "split", "pupY"))`
(`r sprintf("%.2f", cget("bayessbw", "split", "pupY", "lo"))` to
`r sprintf("%.2f", cget("bayessbw", "split", "pupY", "hi"))`) for bayessbw, both
compatible with one day of flight per day of pupation. But a locality whose
predicted pupation is early *relative to the others that year* barely flies
earlier at all: the within-year slopes are
`r sprintf("%.2f", cget("BioSIM", "split", "pupW"))`
(`r sprintf("%.2f", cget("BioSIM", "split", "pupW", "lo"))` to
`r sprintf("%.2f", cget("BioSIM", "split", "pupW", "hi"))`) and
`r sprintf("%.2f", cget("bayessbw", "split", "pupW"))`
(`r sprintf("%.2f", cget("bayessbw", "split", "pupW", "lo"))` to
`r sprintf("%.2f", cget("bayessbw", "split", "pupW", "hi"))`). The models
place the year correctly and the localities within it wrongly, so a fixed lag
inherits the full spatial spread of the predictions, which is mostly not there
in the catches. The coefficients come from mixed models with a random year
intercept, because locality-years of one year share its weather.

With the split, both drivers beat the climatology for a new year
(`r sprintf("%.2f", mget(LOYO, "BioSIM", "split", "rmse"))` and
`r sprintf("%.2f", mget(LOYO, "bayessbw", "split", "rmse"))` days) and for a new
place (`r sprintf("%.2f", mget(LOLO, "BioSIM", "split", "rmse"))` and
`r sprintf("%.2f", mget(LOLO, "bayessbw", "split", "rmse"))`), bayessbw by a
small margin. Both stay well above the same-year floor of
`r sprintf("%.2f", mget(LOYO, "BioSIM", "same-year", "rmse"))` days.

```{r}
#| label: fig-median-pupation
#| fig-cap: "Observed median catch date against predicted median pupation date. Grey: locality-years. Black: year means. Solid line: the year-level slope of the split model through the year means. Dashed line: the fixed-lag model."
#| fig-width: 9
#| fig-height: 4.2
d <- mvp$data
ym <- d |> group_by(driver, Year) |>
  summarise(pup = mean(pup), p50 = mean(p50), .groups = "drop")
lines <- mco |> filter(model == "split", term == "pupY") |>
  select(driver, slope = estimate) |>
  left_join(ym |> group_by(driver) |>
              summarise(mx = mean(pup), my = mean(p50), .groups = "drop"),
            by = "driver") |>
  mutate(intercept = my - slope * mx)
lags <- mco |> filter(model == "lag") |> select(driver, lag = estimate)
ggplot(d, aes(pup, p50)) +
  geom_point(colour = "grey70", size = 1) +
  geom_abline(data = lags, aes(intercept = lag, slope = 1), linetype = "dashed",
              colour = "grey30") +
  geom_abline(data = lines, aes(intercept = intercept, slope = slope),
              colour = "black", linewidth = 0.7) +
  geom_point(data = ym, colour = "black", size = 1.8) +
  facet_wrap(~ factor(driver, c("BioSIM", "bayessbw")), scales = "free_x") +
  labs(x = "Predicted median pupation date (day of year)",
       y = "Observed median catch date (day of year)") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

The year-level relationship carries the predictive skill. For a new year the
split model needs only the network mean of the predicted pupation dates; the
locality's own predicted date adds almost nothing. That is the seasonal
counterpart of the daily analysis, where a single regional BioSIM run beat
site-specific runs.

# Season width and the size of the catch {#sec-width}

```{r}
#| label: width-data
wvc <- readRDS(file.path("data", "processed", "seasonal", "width_vs_catch.rds"))
wf <- wvc$fits; wd <- wvc$decomposition; wn <- wvc$null_summary
wget <- function(w, mod, col) wf[[col]][wf$width == w & wf$model == mod]
dget <- function(w, comp, col) wd[[col]][wd$width == w & wd$component == comp]
```

`seasonal_dynamics/R/04_width_vs_catch.R` asks whether the length of the
flight season depends on how many moths the trap caught, over the
`r wvc$summary$n` locality-years usable for phenology.

```{r}
#| label: tbl-width-fits
#| tbl-cap: "Change in season width per tenfold increase in the season total (days)"
wf |>
  transmute(Width = recode(width, width90 = "central 90%", width50 = "central 50%"),
            Model = model, n,
            `d per tenfold` = sprintf("%+.2f", slope),
            `95% CI` = sprintf("%+.2f to %+.2f", lo, hi),
            `R2` = sprintf("%.3f", r2)) |>
  kable(align = c("l", "l", "r", "r", "r", "r"))
```

**Yes: larger catches have shorter seasons.** The central 90% shortens by
`r sprintf("%.1f", abs(wget("width90", "alone", "slope")))` days per tenfold
increase in catch, and the central 50% by
`r sprintf("%.1f", abs(wget("width50", "alone", "slope")))` days. The slope is
unchanged with year fixed effects and without the one season concentrated on
a single night.

**Sampling explains part of it, not all.** A percentile estimated from a
small catch is noisier, and on a spiky seasonal curve that noise need not
average to zero. The rarefaction null measures the effect directly. Each of
the `r wvc$summary$n_known` seasons of at least 3,000 moths is treated as a
known season, `r wvc$summary$null_reps` smaller catches are drawn from it at
each size, and their widths are measured exactly as for the real data.

```{r}
#| label: tbl-width-null
#| tbl-cap: "Width change produced by sampling alone: catches of N moths drawn from seasons of 3,000 or more"
wn |>
  transmute(`N moths` = format(N, big.mark = ","),
            `central 90%, d` = sprintf("%+.2f (sd %.2f)", d_width90, sd_width90),
            `central 50%, d` = sprintf("%+.2f (sd %.2f)", d_width50, sd_width50)) |>
  kable(align = c("r", "r", "r"))
```

Small catches do come out wider, by about a day at 100 moths and by almost
nothing above 1,000. Subtracting from each record the sampling bias expected
at its own catch size leaves a slope of
`r sprintf("%+.2f", wget("width90", "alone, sampling bias removed", "slope"))`
days per tenfold for the central 90%
(`r sprintf("%+.2f", wget("width90", "alone, sampling bias removed", "lo"))` to
`r sprintf("%+.2f", wget("width90", "alone, sampling bias removed", "hi"))`) and
`r sprintf("%+.2f", wget("width50", "alone, sampling bias removed", "slope"))`
for the central 50%
(`r sprintf("%+.2f", wget("width50", "alone, sampling bias removed", "lo"))` to
`r sprintf("%+.2f", wget("width50", "alone, sampling bias removed", "hi"))`).
About a fifth of the central-90% effect is sampling; the rest belongs to the
seasons.

**The effect is within localities.** Over the `r wd$n_localities[1]`
localities with at least three years, a locality's large-catch years are
shorter than its small-catch years (r =
`r sprintf("%+.2f", dget("width90", "within localities", "r"))` for the central
90%, `r sprintf("%+.2f", dget("width50", "within localities", "r"))` for the
central 50%). The relationship between locality means is weaker
(`r sprintf("%+.2f", dget("width90", "between localities", "r"))` and
`r sprintf("%+.2f", dget("width50", "between localities", "r"))`). This is the
opposite of the median-date offset of the daily analysis, which belonged to
the locality and not to the year. Season width tracks how many moths a given
season produced at a given place.

```{r}
#| label: fig-width-catch
#| fig-cap: "Width of the central 90% of the flight season against the season total. Points: locality-years. Solid line: least-squares fit to the observed widths. Dashed line: the fit after subtracting the width that sampling alone adds at each catch size."
#| fig-width: 6.5
#| fig-height: 4.2
wd_dat <- wvc$data
fit_obs <- stats::lm(width90 ~ log_total, wd_dat)
fit_adj <- stats::lm(width90_adj ~ log_total, wd_dat)
grid <- tibble::tibble(log_total = seq(min(wd_dat$log_total), max(wd_dat$log_total),
                                       length.out = 50)) |>
  mutate(total = 10^log_total,
         observed = stats::predict(fit_obs, newdata = pick(everything())),
         adjusted = stats::predict(fit_adj, newdata = pick(everything())))
ggplot(wd_dat, aes(season_total, width90)) +
  geom_point(colour = "grey55", size = 1.2) +
  geom_line(data = grid, aes(total, observed), colour = "black", linewidth = 0.7) +
  geom_line(data = grid, aes(total, adjusted), colour = "black", linewidth = 0.7,
            linetype = "dashed") +
  scale_x_log10(labels = scales::label_number(big.mark = ",")) +
  labs(x = "Season total (moths, log scale)",
       y = "Central 90% of the season (days)") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

What drives it is not established here. Two readings fit a within-locality
effect: a dense population emerging more synchronously, or a season whose
total is inflated by a few warm nights of mass flight, which concentrates the
catch whatever the emergence. The second predicts that the effect weakens once
nightly temperature is accounted for, which the daily analysis is equipped to
test.


# Next steps

1. **Asymmetry** (not yet modelled): see the options discussed with the
   analysis; the first decision is a measure that does not scale with width.
2. **Temperature summaries per locality-year**, as an alternative driver to
   the development models: degree-days and window means, at the trap and
   averaged over the network.
