---
title: "Spruce budworm flight phenology in Maine"
subtitle: "Seasonal dynamics — analysis log"
author: "JN Candau"
date: today
date-format: "D MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk:
    comment: "#>"
---

```{r}
#| label: setup
## Paths are written relative to the project root, not to
## seasonal_dynamics/analysis/.
knitr::opts_knit$set(root.dir = normalizePath(file.path("..", "..")))
```

```{r}
#| label: data
library(dplyr)
library(knitr)
library(ggplot2)

trap_records <- readRDS(file.path("data", "processed", "trap_records.rds"))
season       <- readRDS(file.path("data", "processed", "seasonal",
                                  "season_percentiles.rds"))
```

# Purpose of this document

This is the running log of the *Seasonal dynamics* analysis of spruce budworm
(*Choristoneura fumiferana*) light-trap catches in Maine. Where the *Daily
dynamics* analysis (`daily_dynamics/`) models the catch night by night, this
one treats each locality-year as a single observation — when its flight season
was centred, how long it lasted and how it was shaped — and asks how those
quantities depend on temperature.

The two analyses share one dataset. Its rules are set once, in the data layer
at the project root (`R/01_read_trap_data.R`, `R/02_read_weather_data.R`), and
hold identically here:

- the season of each year runs from the first to the last night on which any
  trap caught a moth that year;
- a `-` cell inside that window is a night the trap ran and caught nothing,
  i.e. a zero;
- an `m` cell is a night whose count was lost, and stays missing;
- one record (Elliotsville Township 1978) is excluded because its only missing
  nights fell at the peak of its season.

Nothing in this analysis reads an output of `daily_dynamics/`.

# The season-level data {#sec-season}

`seasonal_dynamics/R/01_season_percentiles.R` reduces each record to one row.
The percentiles come from the cumulative catch over the counted nights,
computed with the project's shared helper (`R/functions/phenology.R`), so they
are identical to those used in the daily analysis.

```{r}
#| label: tbl-season-overview
#| tbl-cap: "The season-level response table"
q <- function(x, d = 1) sprintf(paste0("%.", d, "f (%.", d, "f to %.", d, "f)"),
                                stats::median(x), min(x), max(x))
tibble::tibble(
  Quantity = c("Locality-years", "Localities", "Years",
               "Usable for phenology",
               "Median catch date, DOY: median (range)",
               "Central 90% of the season, d",
               "Central 50% of the season, d",
               "Asymmetry, d (positive = longer decline)",
               "Share of the season on the peak night",
               "Records with right-censored nights"),
  Value = c(nrow(season), n_distinct(season$locality), n_distinct(season$year),
            sum(season$usable_phenology),
            q(season$p50), q(season$width90), q(season$width50),
            q(season$asymmetry), q(season$peak_share, 2),
            sum(season$n_cens_right > 0))
) |>
  kable(align = c("l", "r"))
```

```{r}
#| label: fig-season-years
#| fig-cap: "Flight season of every locality-year, by year. Point: median catch date; bar: central 90% of the catch (p05 to p95). Records within a year are ordered by latitude, south to north."
#| fig-width: 9
#| fig-height: 4.5
season |>
  group_by(year) |>
  mutate(offset = (rank(latitude, ties.method = "first") - (n() + 1) / 2) /
           (n() + 1) * 0.8) |>
  ungroup() |>
  ggplot(aes(x = year + offset)) +
  geom_linerange(aes(ymin = p05, ymax = p95), colour = "grey60",
                 linewidth = 0.4) +
  geom_point(aes(y = p50, colour = latitude), size = 1.2) +
  ## plotmath, because the render locale mangles a literal degree sign
  scale_colour_viridis_c(name = expression(Latitude~(degree*N)), option = "D") +
  labs(x = "Year", y = "Day of year") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

The year-to-year spread of the median catch date dwarfs the spread within a
year: the medians of a given year cluster within a few days of each other,
while the year means range over
`r sprintf("%.0f", diff(range(tapply(season$p50, season$year, mean))))` days.
That pattern, already central to the daily analysis, sets the problem for this
one: a seasonal model must first explain the differences between years, and
only then the much smaller differences between localities within a year.

# Next steps

1. **Temperature summaries per locality-year**, from the hourly series of the
   shared weather layer: degree-days above a range of base temperatures
   accumulated from a range of start dates, and mean temperatures over fixed
   calendar windows, computed both at the trap and averaged over the network.
2. **Models of p50, then of width and asymmetry**, as functions of those
   summaries, separating the between-year from the within-year component and
   scored by leave-one-year-out and leave-one-locality-out prediction.
