---
title: "Spruce budworm flight phenology in Maine"
subtitle: "Seasonal dynamics — analysis log"
author: "JN Candau"
date: today
date-format: "D MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk:
    comment: "#>"
---

```{r}
#| label: setup
## Paths are written relative to the project root, not to
## seasonal_dynamics/analysis/.
knitr::opts_knit$set(root.dir = normalizePath(file.path("..", "..")))
```

```{r}
#| label: data
library(dplyr)
library(knitr)
library(ggplot2)

trap_records <- readRDS(file.path("data", "processed", "trap_records.rds"))
season       <- readRDS(file.path("data", "processed", "seasonal",
                                  "season_percentiles.rds"))
```

# Purpose of this document

This is the running log of the *Seasonal dynamics* analysis of spruce budworm
(*Choristoneura fumiferana*) light-trap catches in Maine. Where the *Daily
dynamics* analysis (`daily_dynamics/`) models the catch night by night, this
one treats each locality-year as a single observation — when its flight season
was centred, how long it lasted and how it was shaped — and asks how those
quantities depend on temperature.

The two analyses share one dataset. Its rules are set once, in the data layer
at the project root (`R/01_read_trap_data.R`, `R/02_read_weather_data.R`), and
hold identically here:

- the season of each year runs from the first to the last night on which any
  trap caught a moth that year;
- a `-` cell inside that window is a night the trap ran and caught nothing,
  i.e. a zero;
- an `m` cell is a night whose count was lost, and stays missing;
- one record (Elliotsville Township 1978) is excluded because its only missing
  nights fell at the peak of its season.

The data layer also holds the two development-model runs used below,
BioSIM (`R/03_biosim_phenology.R`) and bayessbw (`R/04_pupal_phenology.R`).
Nothing in this analysis reads an output of `daily_dynamics/`.

# The season-level data {#sec-season}

`seasonal_dynamics/R/01_season_percentiles.R` reduces each record to one row.
The percentiles come from the cumulative catch over the counted nights,
computed with the project's shared helper (`R/functions/phenology.R`), so they
are identical to those used in the daily analysis.

```{r}
#| label: tbl-season-overview
#| tbl-cap: "The season-level response table"
q <- function(x, d = 1) sprintf(paste0("%.", d, "f (%.", d, "f to %.", d, "f)"),
                                stats::median(x), min(x), max(x))
tibble::tibble(
  Quantity = c("Locality-years", "Localities", "Years",
               "Usable for phenology",
               "Median catch date, DOY: median (range)",
               "Central 90% of the season, d",
               "Central 50% of the season, d",
               "Asymmetry, d (positive = longer decline)",
               "Share of the season on the peak night",
               "Records with right-censored nights"),
  Value = c(nrow(season), n_distinct(season$locality), n_distinct(season$year),
            sum(season$usable_phenology),
            q(season$p50), q(season$width90), q(season$width50),
            q(season$asymmetry), q(season$peak_share, 2),
            sum(season$n_cens_right > 0))
) |>
  kable(align = c("l", "r"))
```

```{r}
#| label: fig-season-years
#| fig-cap: "Flight season of every locality-year, by year. Point: median catch date; bar: central 90% of the catch (p05 to p95). Records within a year are ordered by latitude, south to north."
#| fig-width: 9
#| fig-height: 4.5
season |>
  group_by(year) |>
  mutate(offset = (rank(latitude, ties.method = "first") - (n() + 1) / 2) /
           (n() + 1) * 0.8) |>
  ungroup() |>
  ggplot(aes(x = year + offset)) +
  geom_linerange(aes(ymin = p05, ymax = p95), colour = "grey60",
                 linewidth = 0.4) +
  geom_point(aes(y = p50, colour = latitude), size = 1.2) +
  ## plotmath, because the render locale mangles a literal degree sign
  scale_colour_viridis_c(name = expression(Latitude~(degree*N)), option = "D") +
  labs(x = "Year", y = "Day of year") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

The year-to-year spread of the median catch date dwarfs the spread within a
year: the medians of a given year cluster within a few days of each other,
while the year means range over
`r sprintf("%.0f", diff(range(tapply(season$p50, season$year, mean))))` days.
That pattern, already central to the daily analysis, sets the problem for this
one: a seasonal model must first explain the differences between years, and
only then the much smaller differences between localities within a year.

# Predicted pupation dates {#sec-pupation}

```{r}
#| label: pupation-data
pup <- readRDS(file.path("data", "processed", "seasonal", "pupation_dates.rds"))
pup_ok <- filter(pup, biosim_pup_complete)
```

The two development models of the shared data layer predict pupation, not
flight: BioSIM (`R/03_biosim_phenology.R`, mortality switched off) and
bayessbw (`R/04_pupal_phenology.R`). `seasonal_dynamics/R/02_pupation_dates.R`
puts both on the same footing, as a distribution of pupation dates per
locality-year.

For bayessbw this is direct: the model returns the date on which each
simulated individual pupates. BioSIM does not. It reports pupae as a **stock**,
the share of the cohort that is a pupa on a given day, and the median of that
curve is the middle of the pupal period rather than the date of pupation. The
pupation-date distribution is recovered from mass balance. With mortality off,
an individual that has pupated by day *t* is either still a pupa or has
already emerged, so cumulative pupation is the pupae present plus the
cumulative emergence, and its daily increments are the distribution.

The distinction is not cosmetic. The stock median falls a median
`r sprintf("%.1f", median(pup_ok$biosim_stock_p50 - pup_ok$biosim_pup_p50))`
days after the true median pupation date, and with the true date BioSIM
pupates `r sprintf("%.1f", abs(mean(pup_ok$biosim_pup_p50 - pup_ok$bayes_pup_p50)))`
days *earlier* than bayessbw on average, where the stock median had put the two
models within a day of each other. The two remain well correlated across
locality-years (r = `r sprintf("%.2f", cor(pup_ok$biosim_pup_p50, pup_ok$bayes_pup_p50))`).

In `r sum(pup$biosim_pup_lost > 0.5)` locality-years BioSIM's pupae vanish
within a day without emerging, although mortality is switched off. These are
all records whose BioSIM cohort R/03 already flags as incomplete, and they
leave the analysis set below. The pupation dates of the remaining records are
unaffected.

# The median catch date as a function of pupation {#sec-median}

```{r}
#| label: median-data
mvp <- readRDS(file.path("data", "processed", "seasonal", "median_vs_pupation.rds"))
msc <- mvp$scores
mco <- mvp$coefs
mget <- function(sch, dr, mod, col) {
  msc[[col]][msc$scheme == sch & msc$driver == dr & msc$model == mod]
}
cget <- function(dr, mod, term, col = "estimate") {
  mco[[col]][mco$driver == dr & mco$model == mod & mco$term == term]
}
LOYO <- "leave-one-year-out"; LOLO <- "leave-one-locality-out"
```

`seasonal_dynamics/R/03_median_vs_pupation.R` models the observed median
catch date of `r mvp$summary$n_rows` locality-years
(`r mvp$summary$n_years` years, `r mvp$summary$n_locations` localities: usable
for phenology, with a complete BioSIM cohort) from the predicted median
pupation date, separately for each model:

- **lag**: median catch date = pupation + a constant interval;
- **linear**: median catch date = *a* + *b* x pupation;
- **split**: the predicted pupation date is divided into the mean over the
  localities of that year and each locality's departure from it, each with its
  own slope. The year mean uses predicted dates only, so it is available at
  prediction time.

Each is scored twice: leaving out one year at a time, the test for a new year,
and leaving out one locality at a time, the test for a new place. Two
baselines give the scale. *Climatology* predicts the mean date of the training
data. *Same-year* predicts the mean observed date of the other localities that
year; it is not a forecast, since it uses the year being predicted, but it is
the floor that any year-level model faces.

```{r}
#| label: tbl-median-scores
#| tbl-cap: "Skill in predicting the observed median catch date (RMSE and MAE in days)"
msc |>
  filter(!(model == "same-year" & driver == "bayessbw"),
         !(model == "climatology" & driver == "bayessbw")) |>
  mutate(driver = if_else(model %in% c("same-year", "climatology"), "-", driver)) |>
  tidyr::pivot_wider(id_cols = c(driver, model), names_from = scheme,
                     values_from = c(rmse, mae)) |>
  transmute(Driver = driver, Model = model,
            `RMSE, new year` = sprintf("%.2f", .data[[paste0("rmse_", LOYO)]]),
            `MAE, new year` = sprintf("%.2f", .data[[paste0("mae_", LOYO)]]),
            `RMSE, new place` = sprintf("%.2f", .data[[paste0("rmse_", LOLO)]]),
            `MAE, new place` = sprintf("%.2f", .data[[paste0("mae_", LOLO)]])) |>
  arrange(Driver == "-", Driver, `RMSE, new year`) |>
  kable(align = c("l", "l", "r", "r", "r", "r"))
```

**The fixed-lag model is worse than knowing nothing.** Adding a constant
interval to either model's pupation date gives an RMSE of
`r sprintf("%.2f", mget(LOYO, "BioSIM", "lag", "rmse"))` (BioSIM) and
`r sprintf("%.2f", mget(LOYO, "bayessbw", "lag", "rmse"))` days (bayessbw) for a
new year, against `r sprintf("%.2f", mget(LOYO, "BioSIM", "climatology", "rmse"))`
for the climatology.

**The split model shows why.** A year whose predicted pupation is early also
flies early, with a slope of
`r sprintf("%.2f", cget("BioSIM", "split", "pupY"))`
(`r sprintf("%.2f", cget("BioSIM", "split", "pupY", "lo"))` to
`r sprintf("%.2f", cget("BioSIM", "split", "pupY", "hi"))`) for BioSIM and
`r sprintf("%.2f", cget("bayessbw", "split", "pupY"))`
(`r sprintf("%.2f", cget("bayessbw", "split", "pupY", "lo"))` to
`r sprintf("%.2f", cget("bayessbw", "split", "pupY", "hi"))`) for bayessbw, both
compatible with one day of flight per day of pupation. But a locality whose
predicted pupation is early *relative to the others that year* barely flies
earlier at all: the within-year slopes are
`r sprintf("%.2f", cget("BioSIM", "split", "pupW"))`
(`r sprintf("%.2f", cget("BioSIM", "split", "pupW", "lo"))` to
`r sprintf("%.2f", cget("BioSIM", "split", "pupW", "hi"))`) and
`r sprintf("%.2f", cget("bayessbw", "split", "pupW"))`
(`r sprintf("%.2f", cget("bayessbw", "split", "pupW", "lo"))` to
`r sprintf("%.2f", cget("bayessbw", "split", "pupW", "hi"))`). The models
place the year correctly and the localities within it wrongly, so a fixed lag
inherits the full spatial spread of the predictions, which is mostly not there
in the catches. The coefficients come from mixed models with a random year
intercept, because locality-years of one year share its weather.

With the split, both drivers beat the climatology for a new year
(`r sprintf("%.2f", mget(LOYO, "BioSIM", "split", "rmse"))` and
`r sprintf("%.2f", mget(LOYO, "bayessbw", "split", "rmse"))` days) and for a new
place (`r sprintf("%.2f", mget(LOLO, "BioSIM", "split", "rmse"))` and
`r sprintf("%.2f", mget(LOLO, "bayessbw", "split", "rmse"))`), bayessbw by a
small margin. Both stay well above the same-year floor of
`r sprintf("%.2f", mget(LOYO, "BioSIM", "same-year", "rmse"))` days.

```{r}
#| label: fig-median-pupation
#| fig-cap: "Observed median catch date against predicted median pupation date. Grey: locality-years. Black: year means. Solid line: the year-level slope of the split model through the year means. Dashed line: the fixed-lag model."
#| fig-width: 9
#| fig-height: 4.2
d <- mvp$data
ym <- d |> group_by(driver, Year) |>
  summarise(pup = mean(pup), p50 = mean(p50), .groups = "drop")
lines <- mco |> filter(model == "split", term == "pupY") |>
  select(driver, slope = estimate) |>
  left_join(ym |> group_by(driver) |>
              summarise(mx = mean(pup), my = mean(p50), .groups = "drop"),
            by = "driver") |>
  mutate(intercept = my - slope * mx)
lags <- mco |> filter(model == "lag") |> select(driver, lag = estimate)
ggplot(d, aes(pup, p50)) +
  geom_point(colour = "grey70", size = 1) +
  geom_abline(data = lags, aes(intercept = lag, slope = 1), linetype = "dashed",
              colour = "grey30") +
  geom_abline(data = lines, aes(intercept = intercept, slope = slope),
              colour = "black", linewidth = 0.7) +
  geom_point(data = ym, colour = "black", size = 1.8) +
  facet_wrap(~ factor(driver, c("BioSIM", "bayessbw")), scales = "free_x") +
  labs(x = "Predicted median pupation date (day of year)",
       y = "Observed median catch date (day of year)") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

The year-level relationship carries the predictive skill. For a new year the
split model needs only the network mean of the predicted pupation dates; the
locality's own predicted date adds almost nothing. That is the seasonal
counterpart of the daily analysis, where a single regional BioSIM run beat
site-specific runs.

# Season width and the size of the catch {#sec-width}

```{r}
#| label: width-data
wvc <- readRDS(file.path("data", "processed", "seasonal", "width_vs_catch.rds"))
wf <- wvc$fits; wd <- wvc$decomposition; wn <- wvc$null_summary
wget <- function(w, mod, col) wf[[col]][wf$width == w & wf$model == mod]
dget <- function(w, comp, col) wd[[col]][wd$width == w & wd$component == comp]
```

`seasonal_dynamics/R/04_width_vs_catch.R` asks whether the length of the
flight season depends on how many moths the trap caught, over the
`r wvc$summary$n` locality-years usable for phenology.

```{r}
#| label: tbl-width-fits
#| tbl-cap: "Change in season width per tenfold increase in the season total (days)"
wf |>
  transmute(Width = recode(width, width90 = "central 90%", width50 = "central 50%"),
            Model = model, n,
            `d per tenfold` = sprintf("%+.2f", slope),
            `95% CI` = sprintf("%+.2f to %+.2f", lo, hi),
            `R2` = sprintf("%.3f", r2)) |>
  kable(align = c("l", "l", "r", "r", "r", "r"))
```

**Yes: larger catches have shorter seasons.** The central 90% shortens by
`r sprintf("%.1f", abs(wget("width90", "alone", "slope")))` days per tenfold
increase in catch, and the central 50% by
`r sprintf("%.1f", abs(wget("width50", "alone", "slope")))` days. The slope is
unchanged with year fixed effects and without the one season concentrated on
a single night.

**Sampling explains part of it, not all.** A percentile estimated from a
small catch is noisier, and on a spiky seasonal curve that noise need not
average to zero. The rarefaction null measures the effect directly. Each of
the `r wvc$summary$n_known` seasons of at least 3,000 moths is treated as a
known season, `r wvc$summary$null_reps` smaller catches are drawn from it at
each size, and their widths are measured exactly as for the real data.

```{r}
#| label: tbl-width-null
#| tbl-cap: "Width change produced by sampling alone: catches of N moths drawn from seasons of 3,000 or more"
wn |>
  transmute(`N moths` = format(N, big.mark = ","),
            `central 90%, d` = sprintf("%+.2f (sd %.2f)", d_width90, sd_width90),
            `central 50%, d` = sprintf("%+.2f (sd %.2f)", d_width50, sd_width50)) |>
  kable(align = c("r", "r", "r"))
```

Small catches do come out wider, by about a day at 100 moths and by almost
nothing above 1,000. Subtracting from each record the sampling bias expected
at its own catch size leaves a slope of
`r sprintf("%+.2f", wget("width90", "alone, sampling bias removed", "slope"))`
days per tenfold for the central 90%
(`r sprintf("%+.2f", wget("width90", "alone, sampling bias removed", "lo"))` to
`r sprintf("%+.2f", wget("width90", "alone, sampling bias removed", "hi"))`) and
`r sprintf("%+.2f", wget("width50", "alone, sampling bias removed", "slope"))`
for the central 50%
(`r sprintf("%+.2f", wget("width50", "alone, sampling bias removed", "lo"))` to
`r sprintf("%+.2f", wget("width50", "alone, sampling bias removed", "hi"))`).
About a fifth of the central-90% effect is sampling; the rest belongs to the
seasons.

**The effect is within localities.** Over the `r wd$n_localities[1]`
localities with at least three years, a locality's large-catch years are
shorter than its small-catch years (r =
`r sprintf("%+.2f", dget("width90", "within localities", "r"))` for the central
90%, `r sprintf("%+.2f", dget("width50", "within localities", "r"))` for the
central 50%). The relationship between locality means is weaker
(`r sprintf("%+.2f", dget("width90", "between localities", "r"))` and
`r sprintf("%+.2f", dget("width50", "between localities", "r"))`). This is the
opposite of the median-date offset of the daily analysis, which belonged to
the locality and not to the year. Season width tracks how many moths a given
season produced at a given place.

```{r}
#| label: fig-width-catch
#| fig-cap: "Width of the central 90% of the flight season against the season total. Points: locality-years. Solid line: least-squares fit to the observed widths. Dashed line: the fit after subtracting the width that sampling alone adds at each catch size."
#| fig-width: 6.5
#| fig-height: 4.2
wd_dat <- wvc$data
fit_obs <- stats::lm(width90 ~ log_total, wd_dat)
fit_adj <- stats::lm(width90_adj ~ log_total, wd_dat)
grid <- tibble::tibble(log_total = seq(min(wd_dat$log_total), max(wd_dat$log_total),
                                       length.out = 50)) |>
  mutate(total = 10^log_total,
         observed = stats::predict(fit_obs, newdata = pick(everything())),
         adjusted = stats::predict(fit_adj, newdata = pick(everything())))
ggplot(wd_dat, aes(season_total, width90)) +
  geom_point(colour = "grey55", size = 1.2) +
  geom_line(data = grid, aes(total, observed), colour = "black", linewidth = 0.7) +
  geom_line(data = grid, aes(total, adjusted), colour = "black", linewidth = 0.7,
            linetype = "dashed") +
  scale_x_log10(labels = scales::label_number(big.mark = ",")) +
  labs(x = "Season total (moths, log scale)",
       y = "Central 90% of the season (days)") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

What drives it is not established here. Two readings fit a within-locality
effect: a dense population emerging more synchronously, or a season whose
total is inflated by a few warm nights of mass flight, which concentrates the
catch whatever the emergence. The second predicts that the effect weakens once
nightly temperature is accounted for, which the daily analysis is equipped to
test.


# Asymmetry of the season {#sec-asym}

```{r}
#| label: asym-data
asy <- readRDS(file.path("data", "processed", "seasonal", "asymmetry.rds"))
ad  <- asy$data
aloc <- asy$location; acomp <- asy$components; arel <- asy$reliability
acor <- asy$correlations
lget <- function(v, col) aloc[[col]][aloc$measure == v]
vget <- function(v, comp) 100 * acomp$share[acomp$measure == v & acomp$component == comp]
rget <- function(v, col) arel[[col]][arel$measure == v]
kget <- function(v, w) acor$r[acor$measure == v & acor$with == w]
## Split of the asymmetry-date relationship into within-year and between-year
## parts, for the text and the figure.
a_within <- ad |> group_by(year) |> filter(n() >= 3) |>
  mutate(k = kelly - mean(kelly), d = p50 - mean(p50)) |> ungroup()
a_years <- ad |> group_by(year) |>
  summarise(k = mean(kelly), d = mean(p50), .groups = "drop")
```

`seasonal_dynamics/R/05_asymmetry.R` covers the first two steps: choosing a
measure and describing it. It does not yet model asymmetry.

## The measure

The asymmetry of @sec-season, (p95 - p50) - (p50 - p05), is in days. Two
measures that do not depend on the length of the season replace it, both
bounded between -1 and 1 and positive when the decline is longer than the rise:

- **Kelly**, (p90 + p10 - 2 p50) / (p90 - p10), over the central 80% of the catch;
- **Bowley**, (p75 + p25 - 2 p50) / (p75 - p25), over the central 50%.

Neither uses p05 or p95, the percentiles most exposed to the right-censored
records and to single stray nights in the tails. Kelly is the primary measure:
Bowley rests on a central 50% that is often only three or four nights wide.

The concern that motivated the change turns out not to bite. The day-based
measure is uncorrelated with season width (r =
`r sprintf("%+.2f", asy$summary$r_days_width)`), because width changes its
magnitude but not its sign, and Kelly tracks it closely (r =
`r sprintf("%.2f", kget("kelly", "asymmetry in days (01)"))`). The scale-free
measure is still the right one to model, because its value means the same thing
in a ten-day season as in a thirty-day one.

## Is the typical season skewed?

No. Over `r asy$summary$n` locality-years (usable for phenology, at least three
positive nights), the median Kelly asymmetry is
`r sprintf("%+.2f", lget("kelly", "median_raw"))` and
`r sprintf("%.0f%%", 100 * lget("kelly", "share_positive"))` of seasons are
right-skewed. A mixed model with crossed random effects of year and locality,
which does not count seasons sharing a year or a place as independent, puts
the mean at `r sprintf("%+.3f", lget("kelly", "intercept"))`
(`r sprintf("%+.3f", lget("kelly", "lo"))` to
`r sprintf("%+.3f", lget("kelly", "hi"))`). Bowley gives the same answer
(`r sprintf("%+.3f", lget("bowley", "intercept"))`,
`r sprintf("%+.3f", lget("bowley", "lo"))` to
`r sprintf("%+.3f", lget("bowley", "hi"))`). The flight season is symmetric on
average, and individual seasons depart from symmetry in both directions.

## Signal or noise?

```{r}
#| label: tbl-asym-variance
#| tbl-cap: "Where the variation in asymmetry lies. Variance shares from the mixed model; sampling variance from resampling every record's own catch"
tibble::tibble(
  Measure = c("Kelly", "Bowley"),
  `Year, %` = sprintf("%.0f", c(vget("kelly", "year"), vget("bowley", "year"))),
  `Locality, %` = sprintf("%.0f", c(vget("kelly", "locality"), vget("bowley", "locality"))),
  `Locality-year, %` = sprintf("%.0f", c(vget("kelly", "Residual"), vget("bowley", "Residual"))),
  `Sampling sd, median` = sprintf("%.3f", c(rget("kelly", "median_se"), rget("bowley", "median_se"))),
  `Reliability` = sprintf("%.2f", c(rget("kelly", "reliability"), rget("bowley", "reliability")))
) |>
  kable(align = c("l", "r", "r", "r", "r", "r"))
```

Each record's asymmetry is estimated from a finite catch. To measure that
error, `r asy$summary$n_boot` catches of the record's own size were drawn from
its own nightly shares and the measure recomputed. The sampling error is small
(median sd `r sprintf("%.3f", rget("kelly", "median_se"))` against a spread of
`r sprintf("%.2f", sd(ad$kelly))` between seasons), so
`r sprintf("%.0f%%", 100 * rget("kelly", "reliability"))` of the observed
variation is not counting noise.

Almost none of it belongs to the year or to the locality:
`r sprintf("%.0f%%", vget("kelly", "Residual"))` of the variance of Kelly is at
the level of the individual locality-year. A season's asymmetry is not
inherited from its year (as the median date largely is) or from its place (as
the model offsets of the daily analysis were).

What the resampling cannot see is the night-to-night variability of flight
itself. It treats moths as independent draws from the season's curve, whereas
the daily analysis found nightly catches far more variable than that (NB2
θ ≈ 0.19). A single warm or cold night, or a single mass flight, moves a
percentile, and that is part of the "real" variation here. The asymmetry of a
season may therefore be as much a property of its weather in particular weeks
as of its emergence.

## What asymmetry goes with

```{r}
#| label: tbl-asym-cor
#| tbl-cap: "Correlation of the Kelly and Bowley asymmetry with other properties of the season"
acor |>
  filter(with != "asymmetry in days (01)") |>
  tidyr::pivot_wider(names_from = measure, values_from = r) |>
  transmute(`Property` = with, Kelly = sprintf("%+.2f", kelly),
            Bowley = sprintf("%+.2f", bowley)) |>
  kable(align = c("l", "r", "r"))
```

Asymmetry is unrelated to the width of the season, to the size of the catch
and to latitude. Its one clear association is with the median catch date
(r = `r sprintf("%+.2f", kget("kelly", "median catch date"))`), and that
association is entirely **within years**. Among localities of the same year,
one that flies later than the others has a left-skewed season, a long rise
and an abrupt end (r = `r sprintf("%+.2f", cor(a_within$k, a_within$d))` over
`r n_distinct(a_within$year)` years). Between year means there is no relation
(r = `r sprintf("%+.2f", cor(a_years$k, a_years$d))`).

This is not an artefact of the season window. The asymmetry is unrelated to
how close a record's p10 or p90 comes to the first or last capture of its year.

```{r}
#| label: fig-asym
#| fig-cap: "Kelly asymmetry of each locality-year. Left: by year; black points are year means. Right: against the locality's departure from the mean median catch date of its year, with a least-squares line. Positive asymmetry: longer decline than rise."
#| fig-width: 9
#| fig-height: 3.8
library(patchwork)
pa <- ggplot(ad, aes(year, kelly)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_point(colour = "grey55", size = 1, position = position_jitter(width = 0.15, seed = 1)) +
  geom_point(data = a_years, aes(year, k), colour = "black", size = 1.8) +
  labs(x = "Year", y = "Kelly asymmetry") +
  theme_minimal(base_size = 10) + theme(panel.grid.minor = element_blank())
pb <- ggplot(a_within, aes(d, kelly)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_point(colour = "grey55", size = 1) +
  geom_smooth(method = "lm", se = FALSE, colour = "black", linewidth = 0.7) +
  labs(x = "Median catch date minus the year's mean (days)", y = NULL) +
  theme_minimal(base_size = 10) + theme(panel.grid.minor = element_blank())
pa + pb
```

Two readings fit a within-year pattern, and step 3 (@sec-asym-mech) tests
them.

- **A common end to the season.** If flight ends at about the same time
  everywhere in a year, for example at the first cool spell, a locality that
  started late has its decline cut short and an early one has room for a long
  tail.
- **Moths from elsewhere.** If part of the catch is moths that emerged
  elsewhere in the region, a late locality receives them before its own moths
  fly, which stretches its rise, and an early locality receives them after,
  which stretches its decline. That reading fits the daily analysis's finding
  that one regional development curve predicts the catches better than local
  ones.

The first predicts that late localities share their end date with the rest of
the year; the second predicts that the early part of a late locality's catch
is synchronous with the peak elsewhere. Both are testable with the data in
hand.


## Step 3: which part of the season is shared? {#sec-asym-mech}

```{r}
#| label: asym-mech-data
amc <- readRDS(file.path("data", "processed", "seasonal", "asymmetry_mechanisms.rds"))
asp <- amc$spread; asl <- amc$slopes; acn <- amc$contrasts
spget <- function(qq, col) asp[[col]][as.character(asp$q) == qq]
slget <- function(qq, col) asl[[col]][as.character(asl$q) == qq]
cnget <- function(qty, col) acn[[col]][acn$quantity == qty]
## Habitual lateness of localities with at least three years, for the text.
loc_late <- amc$data |> filter(q == "p50") |> group_by(locality) |>
  summarise(late = mean(dep), n = n(), .groups = "drop") |> filter(n >= 3) |>
  left_join(season |> distinct(locality, latitude, longitude), by = "locality")
geo_r2 <- summary(lm(late ~ latitude + longitude, loc_late))$r.squared
```

The two readings of @sec-asym, and the null, disagree about **which part of
the season** is shared among the localities of a year.

| Hypothesis | Spread among localities, by percentile | How each percentile follows a locality's lateness |
|---|---|---|
| Pure shift | the same at every percentile | the same at every percentile |
| Common end | smaller in the late tail | weaker in the late tail |
| Regional mixing | smaller in both tails | weaker in both tails |

: What each explanation predicts {#tbl-asym-hyp}

`seasonal_dynamics/R/06_asymmetry_mechanisms.R` measures both profiles over
`r amc$summary$n` locality-years in `r amc$summary$n_years` years. The spread
is the pooled within-year standard deviation of each percentile, corrected for
sampling error with the same resampling as above, since tail percentiles rest
on fewer moths and would otherwise look more variable. A locality's
**lateness** is its mean departure of the median catch date from its year's
mean, taken over its *other* years; that way it shares no sampling error with
the season it is used to explain. Both profiles are scaled to the median, and
their intervals come from a bootstrap over years.

```{r}
#| label: fig-asym-mech
#| fig-cap: "Left: spread of each percentile among the localities of a year, relative to the spread of the median (sampling error removed). Right: how strongly each percentile follows the locality's lateness in its other years, relative to the median. Bars: 95% bootstrap intervals over years. Dashed line: pure shift."
#| fig-width: 9
#| fig-height: 3.6
pq <- function(d, y, lo, hi, ylab) {
  ggplot(d, aes(q, .data[[y]], group = 1)) +
    geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
    geom_linerange(aes(ymin = .data[[lo]], ymax = .data[[hi]]), colour = "grey55") +
    geom_line(colour = "black") + geom_point(colour = "black", size = 1.8) +
    labs(x = "Percentile of the season's catch", y = ylab) +
    theme_minimal(base_size = 10) + theme(panel.grid.minor = element_blank())
}
pq(asp, "rel_corr", "rel_corr_lo", "rel_corr_hi", "Spread relative to the median") +
  pq(asl, "rel_slope", "rel_lo", "rel_hi", "Slope relative to the median") +
  coord_cartesian(ylim = c(0, 2.5))
```

**No part of the season is pinned to a common date.** Every percentile
varies about equally among the localities of a year: relative to the median,
the spread is `r sprintf("%.2f", cnget("spread, early tail / median", "estimate"))`
in the early tail
(`r sprintf("%.2f", cnget("spread, early tail / median", "lo"))` to
`r sprintf("%.2f", cnget("spread, early tail / median", "hi"))`) and
`r sprintf("%.2f", cnget("spread, late tail / median", "estimate"))` in the
late tail (`r sprintf("%.2f", cnget("spread, late tail / median", "lo"))` to
`r sprintf("%.2f", cnget("spread, late tail / median", "hi"))`). A common end
date would have shrunk the late tail; regional mixing would have shrunk both.
Neither happens.

**But lateness is lopsided.** A locality that is habitually late is late in
its median and in its decline by the full amount: the late tail follows its
lateness at `r sprintf("%.2f", cnget("slope, late tail / median", "estimate"))`
times the median's slope. Its early tail follows at only
`r sprintf("%.2f", cnget("slope, early tail / median", "estimate"))` times
(`r sprintf("%.2f", cnget("slope, early tail / median", "lo"))` to
`r sprintf("%.2f", cnget("slope, early tail / median", "hi"))`), and the
difference,
`r sprintf("%+.2f", cnget("slope, late minus early", "estimate"))`
(`r sprintf("%+.2f", cnget("slope, late minus early", "lo"))` to
`r sprintf("%+.2f", cnget("slope, late minus early", "hi"))`), excludes zero.
Habitually late localities start nearly as early as the others, then peak and
finish late. This is a **common start** for the persistent part of lateness,
the mirror image of the common-end hypothesis, and it produces the long rise
of late localities found in @sec-asym.

The two results are compatible. Habitual lateness is only part of a
locality's departure in a given year: the slope of the median on lateness is
`r sprintf("%.2f", slget("p50", "slope"))`
(`r sprintf("%.2f", slget("p50", "slope_lo"))` to
`r sprintf("%.2f", slget("p50", "slope_hi"))`). The remainder, which is
specific to the year, moves the whole season and keeps the spread profile flat.

What this rules out, and what it leaves:

- **A regional end to flight is ruled out.** Late tails are not shared and
  follow each locality's own timing fully.
- **Symmetric regional mixing is ruled out.** The late tails of early
  localities are not pulled toward the regional dates.
- **Left standing: an early fraction of the catch at late localities that
  flies on the regional schedule.** It could be moths arriving from earlier
  areas before the local population emerges, which would be dispersal in one
  direction only. It could equally be local heterogeneity: warm pockets around
  a cool site that emerge with the rest of the region.

Geography does not separate the two here. Habitual lateness is barely related
to latitude and longitude (R^2^ = `r sprintf("%.2f", geo_r2)` over
`r nrow(loc_late)` localities with at least three years). The two most
habitually early localities, Greenbush and Passadumkeag, are neighbours in the
Penobscot valley, and the late ones are scattered across the state. A
dispersal source would have to be identified from the timing of the early
catches themselves, night by night, which is a question for the daily
analysis.


## Decision: the season is treated as symmetric {#sec-asym-decision}

```{r}
#| label: shape-data
shp <- readRDS(file.path("data", "processed", "seasonal", "asymmetry.rds"))$data |>
  mutate(ratio = width50 / width90)
shape_theory <- c(
  normal   = (qnorm(0.75) - qnorm(0.25)) / (qnorm(0.95) - qnorm(0.05)),
  logistic = (qlogis(0.75) - qlogis(0.25)) / (qlogis(0.95) - qlogis(0.05)),
  laplace  = log(2) / log(10))
set.seed(1978)
shape_ci <- stats::quantile(replicate(2000, stats::median(sample(shp$ratio, replace = TRUE))),
                            c(0.025, 0.975), names = FALSE)
```

The typical season is symmetric (@sec-asym), its asymmetry is idiosyncratic
to the locality-year, and the one systematic pattern — the early start of
habitually late localities (@sec-asym-mech) — is a within-year contrast that a
model of the whole network does not need. The seasonal model therefore treats
the flight season as **symmetric about its median**. Asymmetry is not modelled.

A symmetric season still needs a shape, and the ratio of the central 50% to
the central 90% of the catch identifies it without fitting anything. For a
normal curve the ratio is `r sprintf("%.3f", shape_theory[["normal"]])`, for a
logistic `r sprintf("%.3f", shape_theory[["logistic"]])` and for a Laplace
(double exponential) `r sprintf("%.3f", shape_theory[["laplace"]])`. The
observed median is `r sprintf("%.3f", median(shp$ratio))`
(`r sprintf("%.3f", shape_ci[1])` to `r sprintf("%.3f", shape_ci[2])` over
`r nrow(shp)` locality-years), and `r sprintf("%.3f", median(shp$ratio[shp$season_total >= 3000]))`
in the seasons of 3,000 moths or more, where the percentiles are most precise.
The season is more peaked, with heavier tails, than a normal curve. A logistic
is consistent with the data as a whole, and the best-sampled seasons lie
between the logistic and the Laplace.


# A seasonal predictor {#sec-predictor}

```{r}
#| label: predictor-data
sp  <- readRDS(file.path("data", "processed", "seasonal", "seasonal_predictor.rds"))
sps <- sp$scores; spo <- sp$oracle
LOYO <- "leave-one-year-out"; LOLO <- "leave-one-locality-out"
sget <- function(sch, dr, wm, sh, int, col = "log_score") {
  sps[[col]][sps$scheme == sch & sps$driver == dr & sps$width_model == wm &
               sps$shape == sh & sps$integrate == int]
}
oget <- function(dr, o, col = "log_score") spo[[col]][spo$driver == dr & spo$oracle == o]
## Error of the centre for a new year, weighted by the size of the catch: the
## log score per moth gives each season a weight proportional to its catch.
cerr <- readRDS(file.path("data", "processed", "seasonal", "median_vs_pupation.rds"))$cv |>
  filter(model == "split", scheme == LOYO) |>
  left_join(season |> select(flightID, season_total), by = "flightID") |>
  group_by(driver) |>
  summarise(mae = mean(abs(pred - obs)),
            mae_moth = stats::weighted.mean(abs(pred - obs), season_total),
            .groups = "drop")
ceget <- function(dr, col) cerr[[col]][cerr$driver == dr]
```

`seasonal_dynamics/R/07_seasonal_predictor.R` combines the three results above
into a prediction of how a season's catch is spread over its nights:

- **centre**: the median catch date from the split model of @sec-median, the
  network mean of predicted pupation plus the locality's departure from it;
- **width**: the central 90% of the season from the season total
  (@sec-width), fitted to widths with their sampling bias removed, so that the
  curve describes the season and the counting noise is left to the nights;
- **shape**: symmetric about the centre (@sec-asym-decision), with its scale
  set so that its central 90% equals the predicted width.

The share of night *d* is the curve's mass between *d* - 1 and *d*, which puts
its median on the same scale as the observed one. A small uniform floor (a
fraction `r sp$summary$eps` of the season, spread over its nights) keeps every
night above zero. Shares are then normalised over the record's counted nights.
Every component is refitted on the training data of each fold, leaving out one
year at a time and, separately, one locality at a time. The score is the log
score per moth: the summed log share of the night on which each moth was
caught, divided by the number of moths. It is the log-likelihood of the
nightly catch given its total, and higher is better.

## The centre is uncertain, and the curve must say so

```{r}
#| label: tbl-predictor-scores
#| tbl-cap: "Log score per moth of the seasonal predictor (logistic shape, width from the season total)"
tibble::tibble(
  Driver = rep(c("BioSIM", "bayessbw"), each = 2),
  Centre = rep(c("placed as if known", "averaged over its uncertainty"), 2),
  `New year` = c(sget(LOYO, "BioSIM", "catch", "logistic", FALSE),
                 sget(LOYO, "BioSIM", "catch", "logistic", TRUE),
                 sget(LOYO, "bayessbw", "catch", "logistic", FALSE),
                 sget(LOYO, "bayessbw", "catch", "logistic", TRUE)),
  `New locality` = c(sget(LOLO, "BioSIM", "catch", "logistic", FALSE),
                     sget(LOLO, "BioSIM", "catch", "logistic", TRUE),
                     sget(LOLO, "bayessbw", "catch", "logistic", FALSE),
                     sget(LOLO, "bayessbw", "catch", "logistic", TRUE)),
  `Median error, new year, d` = c(rep(sget(LOYO, "BioSIM", "catch", "logistic", FALSE, "mae_median"), 2),
                                  rep(sget(LOYO, "bayessbw", "catch", "logistic", FALSE, "mae_median"), 2))
) |>
  mutate(across(c(`New year`, `New locality`), ~ sprintf("%.3f", .x)),
         `Median error, new year, d` = sprintf("%.2f", `Median error, new year, d`)) |>
  kable(align = c("l", "l", "r", "r", "r"))
```

The predicted centre misses the observed median by about four days on
average. A curve placed as if its centre were exact is then too confident: a
season that peaks four days from the prediction falls on nights the curve has
given almost nothing. The predictive curve is the season curve **averaged over
the uncertainty of its centre**: a normal error whose SD is the residual SD of
the centre model in the training data, about
`r sprintf("%.1f", sget(LOYO, "BioSIM", "catch", "logistic", TRUE, "sigma"))`
days. This raises the log score for a new year from
`r sprintf("%.3f", sget(LOYO, "BioSIM", "catch", "logistic", FALSE))` to
`r sprintf("%.3f", sget(LOYO, "BioSIM", "catch", "logistic", TRUE))` with
BioSIM, and is the largest single gain in the predictor.

With that done, **BioSIM is the better driver** for nightly shares, for a new
year (`r sprintf("%.3f", sget(LOYO, "BioSIM", "catch", "logistic", TRUE))`
against `r sprintf("%.3f", sget(LOYO, "bayessbw", "catch", "logistic", TRUE))`)
and for a new locality
(`r sprintf("%.3f", sget(LOLO, "BioSIM", "catch", "logistic", TRUE))` against
`r sprintf("%.3f", sget(LOLO, "bayessbw", "catch", "logistic", TRUE))`). This
holds although bayessbw has the slightly smaller median error per locality-year
(`r sprintf("%.2f", ceget("bayessbw", "mae"))` against
`r sprintf("%.2f", ceget("BioSIM", "mae"))` days). The log score per moth
weighs each season by its catch, and weighted that way the ranking reverses:
BioSIM's median error is `r sprintf("%.2f", ceget("BioSIM", "mae_moth"))` days
against `r sprintf("%.2f", ceget("bayessbw", "mae_moth"))` for bayessbw. BioSIM
places the large seasons better, and those are where most of the moths are.

## What the components are worth

```{r}
#| label: tbl-predictor-oracle
#| tbl-cap: "Log score per moth for a new year (BioSIM, logistic, centre placed as if known) when one or both components are replaced by the observed value"
spo |>
  filter(driver == "BioSIM") |>
  transmute(`Given the observed` = recode(oracle, none = "nothing (the predictor)",
                                          width = "width", centre = "centre",
                                          both = "centre and width"),
            `Log score per moth` = sprintf("%.3f", log_score)) |>
  kable(align = c("l", "r"))
```

Almost all the room for improvement is in the centre. Knowing the observed
median catch date would raise the log score from
`r sprintf("%.3f", oget("BioSIM", "none"))` to
`r sprintf("%.3f", oget("BioSIM", "centre"))`; knowing the width as well adds
only `r sprintf("%.3f", oget("BioSIM", "both") - oget("BioSIM", "centre"))`.
Knowing the observed width *alone* makes the prediction worse
(`r sprintf("%.3f", oget("BioSIM", "width"))`). When the centre is wrong, a
true, narrow season concentrates the curve on the wrong nights, whereas the
predicted width, drawn toward the average, hedges the miss.

**The width model is unbiased.** Its mean error against the bias-corrected
widths is `r sprintf("%+.2f", sget(LOYO, "BioSIM", "catch", "logistic", TRUE, "width_err_adj"))`
days for a new year. Adding the predicted spread of pupation as a second
covariate changes nothing, so the development models carry no information
about season width beyond what the catch provides.

**Shape matters little once the centre's uncertainty is averaged in.**

```{r}
#| label: tbl-predictor-shape
#| tbl-cap: "Log score per moth by shape, BioSIM, centre averaged over its uncertainty"
sps |>
  filter(driver == "BioSIM", width_model == "catch", integrate) |>
  select(scheme, shape, log_score) |>
  tidyr::pivot_wider(names_from = scheme, values_from = log_score) |>
  transmute(Shape = shape, `New year` = sprintf("%.3f", .data[[LOYO]]),
            `New locality` = sprintf("%.3f", .data[[LOLO]])) |>
  kable(align = c("l", "r", "r"))
```

The three shapes differ by less than 0.01. Placed as if the centre were known,
the logistic clearly beat the normal, as the shape check of
@sec-asym-decision predicted. Averaged over a four-to-five-day uncertainty in
the centre, the differences in the tails no longer matter. The predictor uses
the logistic, which matches the observed shape of the seasons themselves.

## Held-out predictions

```{r}
#| label: fig-predictor-examples
#| fig-cap: "Held-out predictions (BioSIM, new year: each season predicted from a model fitted without its year). Line: expected catch; points: observed catches, plus one on a log scale. The four seasons of at least 1,000 moths shown are those whose error in the median catch date is closest to the 25th, 50th, 75th and 90th percentiles of that error."
#| fig-width: 9
#| fig-height: 5.5
ex <- sp$nights |>
  group_by(flightID) |>
  mutate(total = sum(moths), expected = share * total) |>
  ungroup() |>
  left_join(season |> select(flightID, locality, year, p50), by = "flightID")
centre_err <- sp$nights |> distinct(flightID, mu) |>
  left_join(season |> select(flightID, p50, season_total), by = "flightID") |>
  filter(season_total >= 1000) |>
  mutate(err = abs(mu - p50))
pick <- sapply(stats::quantile(centre_err$err, c(0.25, 0.5, 0.75, 0.9)),
               function(q) centre_err$flightID[which.min(abs(centre_err$err - q))])
ex |>
  filter(flightID %in% pick) |>
  left_join(centre_err |> select(flightID, err), by = "flightID") |>
  mutate(panel = sprintf("%s %d (median off by %.1f d)",
                         sub(", Maine$", "", locality), year, err)) |>
  ggplot(aes(doy)) +
  geom_line(aes(y = expected + 1), colour = "black", linewidth = 0.6) +
  geom_point(aes(y = moths + 1), colour = "grey40", size = 1.2) +
  scale_y_log10(labels = scales::label_number(big.mark = ",")) +
  facet_wrap(~ panel, scales = "free") +
  labs(x = "Day of year", y = "Moths per night + 1") +
  theme_minimal(base_size = 9) +
  theme(panel.grid.minor = element_blank())
```

## Using the predictor

`seasonal_dynamics/R/functions/predict_season.R` holds the model fitted on all
`r sp$summary$n` locality-years:

```r
source("seasonal_dynamics/R/functions/predict_season.R")

predict_season(
  pup          = 176.2,   # predicted median pupation at the locality
  pup_network  = 174.8,   # mean predicted median pupation over the network that year
  season_total = 5000,    # moths caught over the season
  nights_doy   = 180:215, # nights for which shares are wanted
  driver       = "BioSIM"
)
```

It returns each night's share of the season and its expected catch, with the
predicted centre, width and centre uncertainty as attributes. It needs the
network mean of predicted pupation for the year, because that carries most of
the skill. It gives no interval for a single night's count, since the seasonal
model has no term for nightly variability.


# Next steps

1. **A better centre.** The predictor's error is almost entirely in the median
   catch date (@sec-predictor). Candidates: other summaries of the development
   models (BioSIM emergence or flight rather than pupation), and temperature
   summaries per locality-year (degree-days and window means, at the trap and
   averaged over the network).
2. **Nightly variability**, if prediction intervals for single nights are
   wanted: a dispersion term on top of the seasonal curve.
