---
title: "Spruce budworm flight phenology in Maine"
subtitle: "Seasonal dynamics — analysis log"
author: "JN Candau"
date: today
date-format: "D MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk:
    comment: "#>"
---

```{r}
#| label: setup
## Paths are written relative to the project root, not to
## seasonal_dynamics/analysis/.
knitr::opts_knit$set(root.dir = normalizePath(file.path("..", "..")))
```

```{r}
#| label: data
library(dplyr)
library(knitr)
library(ggplot2)

trap_records <- readRDS(file.path("data", "processed", "trap_records.rds"))
season       <- readRDS(file.path("data", "processed", "seasonal",
                                  "season_percentiles.rds"))
```

# Purpose of this document

This is the running log of the *Seasonal dynamics* analysis of spruce budworm
(*Choristoneura fumiferana*) light-trap catches in Maine. Where the *Daily
dynamics* analysis (`daily_dynamics/`) models the catch night by night, this
one treats each locality-year as a single observation — when its flight season
was centred, how long it lasted and how it was shaped — and asks how those
quantities depend on temperature.

The two analyses share one dataset. Its rules are set once, in the data layer
at the project root (`R/01_read_trap_data.R`, `R/02_read_weather_data.R`), and
hold identically here:

- the season of each year runs from the first to the last night on which any
  trap caught a moth that year;
- a `-` cell inside that window is a night the trap ran and caught nothing,
  i.e. a zero;
- an `m` cell is a night whose count was lost, and stays missing;
- one record (Elliotsville Township 1978) is excluded because its only missing
  nights fell at the peak of its season.

The data layer also holds the two development-model runs used below,
BioSIM (`R/03_biosim_phenology.R`) and bayessbw (`R/04_pupal_phenology.R`).
Nothing in this analysis reads an output of `daily_dynamics/`.

# The season-level data {#sec-season}

`seasonal_dynamics/R/01_season_percentiles.R` reduces each record to one row.
The percentiles come from the cumulative catch over the counted nights,
computed with the project's shared helper (`R/functions/phenology.R`), so they
are identical to those used in the daily analysis.

```{r}
#| label: tbl-season-overview
#| tbl-cap: "The season-level response table"
q <- function(x, d = 1) sprintf(paste0("%.", d, "f (%.", d, "f to %.", d, "f)"),
                                stats::median(x), min(x), max(x))
tibble::tibble(
  Quantity = c("Locality-years", "Localities", "Years",
               "Usable for phenology",
               "Median catch date, DOY: median (range)",
               "Central 90% of the season, d",
               "Central 50% of the season, d",
               "Asymmetry, d (positive = longer decline)",
               "Share of the season on the peak night",
               "Records with right-censored nights"),
  Value = c(nrow(season), n_distinct(season$locality), n_distinct(season$year),
            sum(season$usable_phenology),
            q(season$p50), q(season$width90), q(season$width50),
            q(season$asymmetry), q(season$peak_share, 2),
            sum(season$n_cens_right > 0))
) |>
  kable(align = c("l", "r"))
```

```{r}
#| label: fig-season-years
#| fig-cap: "Flight season of every locality-year, by year. Point: median catch date; bar: central 90% of the catch (p05 to p95). Records within a year are ordered by latitude, south to north."
#| fig-width: 9
#| fig-height: 4.5
season |>
  group_by(year) |>
  mutate(offset = (rank(latitude, ties.method = "first") - (n() + 1) / 2) /
           (n() + 1) * 0.8) |>
  ungroup() |>
  ggplot(aes(x = year + offset)) +
  geom_linerange(aes(ymin = p05, ymax = p95), colour = "grey60",
                 linewidth = 0.4) +
  geom_point(aes(y = p50, colour = latitude), size = 1.2) +
  ## plotmath, because the render locale mangles a literal degree sign
  scale_colour_viridis_c(name = expression(Latitude~(degree*N)), option = "D") +
  labs(x = "Year", y = "Day of year") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

The year-to-year spread of the median catch date dwarfs the spread within a
year: the medians of a given year cluster within a few days of each other,
while the year means range over
`r sprintf("%.0f", diff(range(tapply(season$p50, season$year, mean))))` days.
That pattern, already central to the daily analysis, sets the problem for this
one: a seasonal model must first explain the differences between years, and
only then the much smaller differences between localities within a year.

# Predicted pupation dates {#sec-pupation}

```{r}
#| label: pupation-data
pup <- readRDS(file.path("data", "processed", "seasonal", "pupation_dates.rds"))
pup_ok <- filter(pup, biosim_pup_complete)
```

The two development models of the shared data layer predict pupation, not
flight: BioSIM (`R/03_biosim_phenology.R`, mortality switched off) and
bayessbw (`R/04_pupal_phenology.R`). `seasonal_dynamics/R/02_pupation_dates.R`
puts both on the same footing, as a distribution of pupation dates per
locality-year.

For bayessbw this is direct: the model returns the date on which each
simulated individual pupates. BioSIM does not. It reports pupae as a **stock**,
the share of the cohort that is a pupa on a given day, and the median of that
curve is the middle of the pupal period rather than the date of pupation. The
pupation-date distribution is recovered from mass balance. With mortality off,
an individual that has pupated by day *t* is either still a pupa or has
already emerged, so cumulative pupation is the pupae present plus the
cumulative emergence, and its daily increments are the distribution.

The distinction is not cosmetic. The stock median falls a median
`r sprintf("%.1f", median(pup_ok$biosim_stock_p50 - pup_ok$biosim_pup_p50))`
days after the true median pupation date, and with the true date BioSIM
pupates `r sprintf("%.1f", abs(mean(pup_ok$biosim_pup_p50 - pup_ok$bayes_pup_p50)))`
days *earlier* than bayessbw on average, where the stock median had put the two
models within a day of each other. The two remain well correlated across
locality-years (r = `r sprintf("%.2f", cor(pup_ok$biosim_pup_p50, pup_ok$bayes_pup_p50))`).

In `r sum(pup$biosim_pup_lost > 0.5, na.rm = TRUE)` locality-years BioSIM's pupae vanish
within a day without emerging, although mortality is switched off. These are
all records whose BioSIM cohort R/03 already flags as incomplete, and they
leave the analysis set below. The pupation dates of the remaining records are
unaffected.

# The median catch date as a function of pupation {#sec-median}

```{r}
#| label: median-data
mvp <- readRDS(file.path("data", "processed", "seasonal", "median_vs_pupation.rds"))
msc <- mvp$scores
mco <- mvp$coefs
mget <- function(sch, dr, mod, col) {
  msc[[col]][msc$scheme == sch & msc$driver == dr & msc$model == mod]
}
cget <- function(dr, mod, term, col = "estimate") {
  mco[[col]][mco$driver == dr & mco$model == mod & mco$term == term]
}
LOYO <- "leave-one-year-out"; LOLO <- "leave-one-locality-out"
```

`seasonal_dynamics/R/03_median_vs_pupation.R` models the observed median
catch date of `r mvp$summary$n_rows` locality-years
(`r mvp$summary$n_years` years, `r mvp$summary$n_locations` localities: usable
for phenology, with a complete BioSIM cohort) from the predicted median
pupation date, separately for each model:

- **lag**: median catch date = pupation + a constant interval;
- **linear**: median catch date = *a* + *b* x pupation;
- **split**: the predicted pupation date is divided into the mean over the
  localities of that year and each locality's departure from it, each with its
  own slope. The year mean uses predicted dates only, so it is available at
  prediction time.

Each is scored twice: leaving out one year at a time, the test for a new year,
and leaving out one locality at a time, the test for a new place. Two
baselines give the scale. *Climatology* predicts the mean date of the training
data. *Same-year* predicts the mean observed date of the other localities that
year; it is not a forecast, since it uses the year being predicted, but it is
the floor that any year-level model faces.

```{r}
#| label: tbl-median-scores
#| tbl-cap: "Skill in predicting the observed median catch date (RMSE and MAE in days)"
msc |>
  filter(!(model == "same-year" & driver == "bayessbw"),
         !(model == "climatology" & driver == "bayessbw")) |>
  mutate(driver = if_else(model %in% c("same-year", "climatology"), "-", driver)) |>
  tidyr::pivot_wider(id_cols = c(driver, model), names_from = scheme,
                     values_from = c(rmse, mae)) |>
  transmute(Driver = driver, Model = model,
            `RMSE, new year` = sprintf("%.2f", .data[[paste0("rmse_", LOYO)]]),
            `MAE, new year` = sprintf("%.2f", .data[[paste0("mae_", LOYO)]]),
            `RMSE, new place` = sprintf("%.2f", .data[[paste0("rmse_", LOLO)]]),
            `MAE, new place` = sprintf("%.2f", .data[[paste0("mae_", LOLO)]])) |>
  arrange(Driver == "-", Driver, `RMSE, new year`) |>
  kable(align = c("l", "l", "r", "r", "r", "r"))
```

**The fixed-lag model is worse than knowing nothing.** Adding a constant
interval to either model's pupation date gives an RMSE of
`r sprintf("%.2f", mget(LOYO, "BioSIM", "lag", "rmse"))` (BioSIM) and
`r sprintf("%.2f", mget(LOYO, "bayessbw", "lag", "rmse"))` days (bayessbw) for a
new year, against `r sprintf("%.2f", mget(LOYO, "BioSIM", "climatology", "rmse"))`
for the climatology.

**The split model shows why.** A year whose predicted pupation is early also
flies early, with a slope of
`r sprintf("%.2f", cget("BioSIM", "split", "pupY"))`
(`r sprintf("%.2f", cget("BioSIM", "split", "pupY", "lo"))` to
`r sprintf("%.2f", cget("BioSIM", "split", "pupY", "hi"))`) for BioSIM and
`r sprintf("%.2f", cget("bayessbw", "split", "pupY"))`
(`r sprintf("%.2f", cget("bayessbw", "split", "pupY", "lo"))` to
`r sprintf("%.2f", cget("bayessbw", "split", "pupY", "hi"))`) for bayessbw, both
compatible with one day of flight per day of pupation. But a locality whose
predicted pupation is early *relative to the others that year* barely flies
earlier at all: the within-year slopes are
`r sprintf("%.2f", cget("BioSIM", "split", "pupW"))`
(`r sprintf("%.2f", cget("BioSIM", "split", "pupW", "lo"))` to
`r sprintf("%.2f", cget("BioSIM", "split", "pupW", "hi"))`) and
`r sprintf("%.2f", cget("bayessbw", "split", "pupW"))`
(`r sprintf("%.2f", cget("bayessbw", "split", "pupW", "lo"))` to
`r sprintf("%.2f", cget("bayessbw", "split", "pupW", "hi"))`). The models
place the year correctly and the localities within it wrongly, so a fixed lag
inherits the full spatial spread of the predictions, which is mostly not there
in the catches. The coefficients come from mixed models with a random year
intercept, because locality-years of one year share its weather.

With the split, both drivers beat the climatology for a new year
(`r sprintf("%.2f", mget(LOYO, "BioSIM", "split", "rmse"))` and
`r sprintf("%.2f", mget(LOYO, "bayessbw", "split", "rmse"))` days) and for a new
place (`r sprintf("%.2f", mget(LOLO, "BioSIM", "split", "rmse"))` and
`r sprintf("%.2f", mget(LOLO, "bayessbw", "split", "rmse"))`), bayessbw by a
small margin. Both stay well above the same-year floor of
`r sprintf("%.2f", mget(LOYO, "BioSIM", "same-year", "rmse"))` days.

```{r}
#| label: fig-median-pupation
#| fig-cap: "Observed median catch date against predicted median pupation date. Grey: locality-years. Black: year means. Solid line: the year-level slope of the split model through the year means. Dashed line: the fixed-lag model."
#| fig-width: 9
#| fig-height: 4.2
d <- mvp$data
ym <- d |> group_by(driver, Year) |>
  summarise(pup = mean(pup), p50 = mean(p50), .groups = "drop")
lines <- mco |> filter(model == "split", term == "pupY") |>
  select(driver, slope = estimate) |>
  left_join(ym |> group_by(driver) |>
              summarise(mx = mean(pup), my = mean(p50), .groups = "drop"),
            by = "driver") |>
  mutate(intercept = my - slope * mx)
lags <- mco |> filter(model == "lag") |> select(driver, lag = estimate)
ggplot(d, aes(pup, p50)) +
  geom_point(colour = "grey70", size = 1) +
  geom_abline(data = lags, aes(intercept = lag, slope = 1), linetype = "dashed",
              colour = "grey30") +
  geom_abline(data = lines, aes(intercept = intercept, slope = slope),
              colour = "black", linewidth = 0.7) +
  geom_point(data = ym, colour = "black", size = 1.8) +
  facet_wrap(~ factor(driver, c("BioSIM", "bayessbw")), scales = "free_x") +
  labs(x = "Predicted median pupation date (day of year)",
       y = "Observed median catch date (day of year)") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

The year-level relationship carries the predictive skill. For a new year the
split model needs only the network mean of the predicted pupation dates; the
locality's own predicted date adds almost nothing. That is the seasonal
counterpart of the daily analysis, where a single regional BioSIM run beat
site-specific runs.

# Season width and the size of the catch {#sec-width}

```{r}
#| label: width-data
wvc <- readRDS(file.path("data", "processed", "seasonal", "width_vs_catch.rds"))
wf <- wvc$fits; wd <- wvc$decomposition; wn <- wvc$null_summary
wget <- function(w, mod, col) wf[[col]][wf$width == w & wf$model == mod]
dget <- function(w, comp, col) wd[[col]][wd$width == w & wd$component == comp]
```

`seasonal_dynamics/R/04_width_vs_catch.R` asks whether the length of the
flight season depends on how many moths the trap caught, over the
`r wvc$summary$n` locality-years usable for phenology.

```{r}
#| label: tbl-width-fits
#| tbl-cap: "Change in season width per tenfold increase in the season total (days)"
wf |>
  transmute(Width = recode(width, width90 = "central 90%", width50 = "central 50%"),
            Model = model, n,
            `d per tenfold` = sprintf("%+.2f", slope),
            `95% CI` = sprintf("%+.2f to %+.2f", lo, hi),
            `R2` = sprintf("%.3f", r2)) |>
  kable(align = c("l", "l", "r", "r", "r", "r"))
```

**Yes: larger catches have shorter seasons.** The central 90% shortens by
`r sprintf("%.1f", abs(wget("width90", "alone", "slope")))` days per tenfold
increase in catch, and the central 50% by
`r sprintf("%.1f", abs(wget("width50", "alone", "slope")))` days. The slope is
unchanged with year fixed effects and without the one season concentrated on
a single night.

**Sampling explains part of it, not all.** A percentile estimated from a
small catch is noisier, and on a spiky seasonal curve that noise need not
average to zero. The rarefaction null measures the effect directly. Each of
the `r wvc$summary$n_known` seasons of at least 3,000 moths is treated as a
known season, `r wvc$summary$null_reps` smaller catches are drawn from it at
each size, and their widths are measured exactly as for the real data.

```{r}
#| label: tbl-width-null
#| tbl-cap: "Width change produced by sampling alone: catches of N moths drawn from seasons of 3,000 or more"
wn |>
  transmute(`N moths` = format(N, big.mark = ","),
            `central 90%, d` = sprintf("%+.2f (sd %.2f)", d_width90, sd_width90),
            `central 50%, d` = sprintf("%+.2f (sd %.2f)", d_width50, sd_width50)) |>
  kable(align = c("r", "r", "r"))
```

Small catches do come out wider, by about a day at 100 moths and by almost
nothing above 1,000. Subtracting from each record the sampling bias expected
at its own catch size leaves a slope of
`r sprintf("%+.2f", wget("width90", "alone, sampling bias removed", "slope"))`
days per tenfold for the central 90%
(`r sprintf("%+.2f", wget("width90", "alone, sampling bias removed", "lo"))` to
`r sprintf("%+.2f", wget("width90", "alone, sampling bias removed", "hi"))`) and
`r sprintf("%+.2f", wget("width50", "alone, sampling bias removed", "slope"))`
for the central 50%
(`r sprintf("%+.2f", wget("width50", "alone, sampling bias removed", "lo"))` to
`r sprintf("%+.2f", wget("width50", "alone, sampling bias removed", "hi"))`).
About a fifth of the central-90% effect is sampling; the rest belongs to the
seasons.

**The effect is within localities.** Over the `r wd$n_localities[1]`
localities with at least three years, a locality's large-catch years are
shorter than its small-catch years (r =
`r sprintf("%+.2f", dget("width90", "within localities", "r"))` for the central
90%, `r sprintf("%+.2f", dget("width50", "within localities", "r"))` for the
central 50%). The relationship between locality means is weaker
(`r sprintf("%+.2f", dget("width90", "between localities", "r"))` and
`r sprintf("%+.2f", dget("width50", "between localities", "r"))`). This is the
opposite of the median-date offset of the daily analysis, which belonged to
the locality and not to the year. Season width tracks how many moths a given
season produced at a given place.

```{r}
#| label: fig-width-catch
#| fig-cap: "Width of the central 90% of the flight season against the season total. Points: locality-years. Solid line: least-squares fit to the observed widths. Dashed line: the fit after subtracting the width that sampling alone adds at each catch size."
#| fig-width: 6.5
#| fig-height: 4.2
wd_dat <- wvc$data
fit_obs <- stats::lm(width90 ~ log_total, wd_dat)
fit_adj <- stats::lm(width90_adj ~ log_total, wd_dat)
grid <- tibble::tibble(log_total = seq(min(wd_dat$log_total), max(wd_dat$log_total),
                                       length.out = 50)) |>
  mutate(total = 10^log_total,
         observed = stats::predict(fit_obs, newdata = pick(everything())),
         adjusted = stats::predict(fit_adj, newdata = pick(everything())))
ggplot(wd_dat, aes(season_total, width90)) +
  geom_point(colour = "grey55", size = 1.2) +
  geom_line(data = grid, aes(total, observed), colour = "black", linewidth = 0.7) +
  geom_line(data = grid, aes(total, adjusted), colour = "black", linewidth = 0.7,
            linetype = "dashed") +
  scale_x_log10(labels = scales::label_number(big.mark = ",")) +
  labs(x = "Season total (moths, log scale)",
       y = "Central 90% of the season (days)") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

What drives it is not established here. Two readings fit a within-locality
effect: a dense population emerging more synchronously, or a season whose
total is inflated by a few warm nights of mass flight, which concentrates the
catch whatever the emergence. The second predicts that the effect weakens once
nightly temperature is accounted for, which the daily analysis is equipped to
test.


# Asymmetry of the season {#sec-asym}

```{r}
#| label: asym-data
asy <- readRDS(file.path("data", "processed", "seasonal", "asymmetry.rds"))
ad  <- asy$data
aloc <- asy$location; acomp <- asy$components; arel <- asy$reliability
acor <- asy$correlations
lget <- function(v, col) aloc[[col]][aloc$measure == v]
vget <- function(v, comp) 100 * acomp$share[acomp$measure == v & acomp$component == comp]
rget <- function(v, col) arel[[col]][arel$measure == v]
kget <- function(v, w) acor$r[acor$measure == v & acor$with == w]
## Split of the asymmetry-date relationship into within-year and between-year
## parts, for the text and the figure.
a_within <- ad |> group_by(year) |> filter(n() >= 3) |>
  mutate(k = kelly - mean(kelly), d = p50 - mean(p50)) |> ungroup()
a_years <- ad |> group_by(year) |>
  summarise(k = mean(kelly), d = mean(p50), .groups = "drop")
```

`seasonal_dynamics/R/05_asymmetry.R` covers the first two steps: choosing a
measure and describing it. It does not yet model asymmetry.

## The measure

The asymmetry of @sec-season, (p95 - p50) - (p50 - p05), is in days. Two
measures that do not depend on the length of the season replace it, both
bounded between -1 and 1 and positive when the decline is longer than the rise:

- **Kelly**, (p90 + p10 - 2 p50) / (p90 - p10), over the central 80% of the catch;
- **Bowley**, (p75 + p25 - 2 p50) / (p75 - p25), over the central 50%.

Neither uses p05 or p95, the percentiles most exposed to the right-censored
records and to single stray nights in the tails. Kelly is the primary measure:
Bowley rests on a central 50% that is often only three or four nights wide.

The concern that motivated the change turns out not to bite. The day-based
measure is uncorrelated with season width (r =
`r sprintf("%+.2f", asy$summary$r_days_width)`), because width changes its
magnitude but not its sign, and Kelly tracks it closely (r =
`r sprintf("%.2f", kget("kelly", "asymmetry in days (01)"))`). The scale-free
measure is still the right one to model, because its value means the same thing
in a ten-day season as in a thirty-day one.

## Is the typical season skewed?

No. Over `r asy$summary$n` locality-years (usable for phenology, at least three
positive nights), the median Kelly asymmetry is
`r sprintf("%+.2f", lget("kelly", "median_raw"))` and
`r sprintf("%.0f%%", 100 * lget("kelly", "share_positive"))` of seasons are
right-skewed. A mixed model with crossed random effects of year and locality,
which does not count seasons sharing a year or a place as independent, puts
the mean at `r sprintf("%+.3f", lget("kelly", "intercept"))`
(`r sprintf("%+.3f", lget("kelly", "lo"))` to
`r sprintf("%+.3f", lget("kelly", "hi"))`). Bowley gives the same answer
(`r sprintf("%+.3f", lget("bowley", "intercept"))`,
`r sprintf("%+.3f", lget("bowley", "lo"))` to
`r sprintf("%+.3f", lget("bowley", "hi"))`). The flight season is symmetric on
average, and individual seasons depart from symmetry in both directions.

## Signal or noise?

```{r}
#| label: tbl-asym-variance
#| tbl-cap: "Where the variation in asymmetry lies. Variance shares from the mixed model; sampling variance from resampling every record's own catch"
tibble::tibble(
  Measure = c("Kelly", "Bowley"),
  `Year, %` = sprintf("%.0f", c(vget("kelly", "year"), vget("bowley", "year"))),
  `Locality, %` = sprintf("%.0f", c(vget("kelly", "locality"), vget("bowley", "locality"))),
  `Locality-year, %` = sprintf("%.0f", c(vget("kelly", "Residual"), vget("bowley", "Residual"))),
  `Sampling sd, median` = sprintf("%.3f", c(rget("kelly", "median_se"), rget("bowley", "median_se"))),
  `Reliability` = sprintf("%.2f", c(rget("kelly", "reliability"), rget("bowley", "reliability")))
) |>
  kable(align = c("l", "r", "r", "r", "r", "r"))
```

Each record's asymmetry is estimated from a finite catch. To measure that
error, `r asy$summary$n_boot` catches of the record's own size were drawn from
its own nightly shares and the measure recomputed. The sampling error is small
(median sd `r sprintf("%.3f", rget("kelly", "median_se"))` against a spread of
`r sprintf("%.2f", sd(ad$kelly))` between seasons), so
`r sprintf("%.0f%%", 100 * rget("kelly", "reliability"))` of the observed
variation is not counting noise.

Almost none of it belongs to the year or to the locality:
`r sprintf("%.0f%%", vget("kelly", "Residual"))` of the variance of Kelly is at
the level of the individual locality-year. A season's asymmetry is not
inherited from its year (as the median date largely is) or from its place (as
the model offsets of the daily analysis were).

What the resampling cannot see is the night-to-night variability of flight
itself. It treats moths as independent draws from the season's curve, whereas
the daily analysis found nightly catches far more variable than that (NB2
θ ≈ 0.19). A single warm or cold night, or a single mass flight, moves a
percentile, and that is part of the "real" variation here. The asymmetry of a
season may therefore be as much a property of its weather in particular weeks
as of its emergence.

## What asymmetry goes with

```{r}
#| label: tbl-asym-cor
#| tbl-cap: "Correlation of the Kelly and Bowley asymmetry with other properties of the season"
acor |>
  filter(with != "asymmetry in days (01)") |>
  tidyr::pivot_wider(names_from = measure, values_from = r) |>
  transmute(`Property` = with, Kelly = sprintf("%+.2f", kelly),
            Bowley = sprintf("%+.2f", bowley)) |>
  kable(align = c("l", "r", "r"))
```

Asymmetry is unrelated to the width of the season, to the size of the catch
and to latitude. Its one clear association is with the median catch date
(r = `r sprintf("%+.2f", kget("kelly", "median catch date"))`), and that
association is entirely **within years**. Among localities of the same year,
one that flies later than the others has a left-skewed season, a long rise
and an abrupt end (r = `r sprintf("%+.2f", cor(a_within$k, a_within$d))` over
`r n_distinct(a_within$year)` years). Between year means there is no relation
(r = `r sprintf("%+.2f", cor(a_years$k, a_years$d))`).

This is not an artefact of the season window. The asymmetry is unrelated to
how close a record's p10 or p90 comes to the first or last capture of its year.

```{r}
#| label: fig-asym
#| fig-cap: "Kelly asymmetry of each locality-year. Left: by year; black points are year means. Right: against the locality's departure from the mean median catch date of its year, with a least-squares line. Positive asymmetry: longer decline than rise."
#| fig-width: 9
#| fig-height: 3.8
library(patchwork)
pa <- ggplot(ad, aes(year, kelly)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_point(colour = "grey55", size = 1, position = position_jitter(width = 0.15, seed = 1)) +
  geom_point(data = a_years, aes(year, k), colour = "black", size = 1.8) +
  labs(x = "Year", y = "Kelly asymmetry") +
  theme_minimal(base_size = 10) + theme(panel.grid.minor = element_blank())
pb <- ggplot(a_within, aes(d, kelly)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_point(colour = "grey55", size = 1) +
  geom_smooth(method = "lm", se = FALSE, colour = "black", linewidth = 0.7) +
  labs(x = "Median catch date minus the year's mean (days)", y = NULL) +
  theme_minimal(base_size = 10) + theme(panel.grid.minor = element_blank())
pa + pb
```

Two readings fit a within-year pattern, and step 3 (@sec-asym-mech) tests
them.

- **A common end to the season.** If flight ends at about the same time
  everywhere in a year, for example at the first cool spell, a locality that
  started late has its decline cut short and an early one has room for a long
  tail.
- **Moths from elsewhere.** If part of the catch is moths that emerged
  elsewhere in the region, a late locality receives them before its own moths
  fly, which stretches its rise, and an early locality receives them after,
  which stretches its decline. That reading fits the daily analysis's finding
  that one regional development curve predicts the catches better than local
  ones.

The first predicts that late localities share their end date with the rest of
the year; the second predicts that the early part of a late locality's catch
is synchronous with the peak elsewhere. Both are testable with the data in
hand.


## Step 3: which part of the season is shared? {#sec-asym-mech}

```{r}
#| label: asym-mech-data
amc <- readRDS(file.path("data", "processed", "seasonal", "asymmetry_mechanisms.rds"))
asp <- amc$spread; asl <- amc$slopes; acn <- amc$contrasts
spget <- function(qq, col) asp[[col]][as.character(asp$q) == qq]
slget <- function(qq, col) asl[[col]][as.character(asl$q) == qq]
cnget <- function(qty, col) acn[[col]][acn$quantity == qty]
## Habitual lateness of localities with at least three years, for the text.
loc_late <- amc$data |> filter(q == "p50") |> group_by(locality) |>
  summarise(late = mean(dep), n = n(), .groups = "drop") |> filter(n >= 3) |>
  left_join(season |> distinct(locality, latitude, longitude), by = "locality")
geo_r2 <- summary(lm(late ~ latitude + longitude, loc_late))$r.squared
```

The two readings of @sec-asym, and the null, disagree about **which part of
the season** is shared among the localities of a year.

| Hypothesis | Spread among localities, by percentile | How each percentile follows a locality's lateness |
|---|---|---|
| Pure shift | the same at every percentile | the same at every percentile |
| Common end | smaller in the late tail | weaker in the late tail |
| Regional mixing | smaller in both tails | weaker in both tails |

: What each explanation predicts {#tbl-asym-hyp}

`seasonal_dynamics/R/06_asymmetry_mechanisms.R` measures both profiles over
`r amc$summary$n` locality-years in `r amc$summary$n_years` years. The spread
is the pooled within-year standard deviation of each percentile, corrected for
sampling error with the same resampling as above, since tail percentiles rest
on fewer moths and would otherwise look more variable. A locality's
**lateness** is its mean departure of the median catch date from its year's
mean, taken over its *other* years; that way it shares no sampling error with
the season it is used to explain. Both profiles are scaled to the median, and
their intervals come from a bootstrap over years.

```{r}
#| label: fig-asym-mech
#| fig-cap: "Left: spread of each percentile among the localities of a year, relative to the spread of the median (sampling error removed). Right: how strongly each percentile follows the locality's lateness in its other years, relative to the median. Bars: 95% bootstrap intervals over years. Dashed line: pure shift."
#| fig-width: 9
#| fig-height: 3.6
pq <- function(d, y, lo, hi, ylab) {
  ggplot(d, aes(q, .data[[y]], group = 1)) +
    geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
    geom_linerange(aes(ymin = .data[[lo]], ymax = .data[[hi]]), colour = "grey55") +
    geom_line(colour = "black") + geom_point(colour = "black", size = 1.8) +
    labs(x = "Percentile of the season's catch", y = ylab) +
    theme_minimal(base_size = 10) + theme(panel.grid.minor = element_blank())
}
pq(asp, "rel_corr", "rel_corr_lo", "rel_corr_hi", "Spread relative to the median") +
  pq(asl, "rel_slope", "rel_lo", "rel_hi", "Slope relative to the median") +
  coord_cartesian(ylim = c(0, 2.5))
```

**No part of the season is pinned to a common date.** Every percentile
varies about equally among the localities of a year: relative to the median,
the spread is `r sprintf("%.2f", cnget("spread, early tail / median", "estimate"))`
in the early tail
(`r sprintf("%.2f", cnget("spread, early tail / median", "lo"))` to
`r sprintf("%.2f", cnget("spread, early tail / median", "hi"))`) and
`r sprintf("%.2f", cnget("spread, late tail / median", "estimate"))` in the
late tail (`r sprintf("%.2f", cnget("spread, late tail / median", "lo"))` to
`r sprintf("%.2f", cnget("spread, late tail / median", "hi"))`). A common end
date would have shrunk the late tail; regional mixing would have shrunk both.
Neither happens.

**But lateness is lopsided.** A locality that is habitually late is late in
its median and in its decline by the full amount: the late tail follows its
lateness at `r sprintf("%.2f", cnget("slope, late tail / median", "estimate"))`
times the median's slope. Its early tail follows at only
`r sprintf("%.2f", cnget("slope, early tail / median", "estimate"))` times
(`r sprintf("%.2f", cnget("slope, early tail / median", "lo"))` to
`r sprintf("%.2f", cnget("slope, early tail / median", "hi"))`), and the
difference,
`r sprintf("%+.2f", cnget("slope, late minus early", "estimate"))`
(`r sprintf("%+.2f", cnget("slope, late minus early", "lo"))` to
`r sprintf("%+.2f", cnget("slope, late minus early", "hi"))`), excludes zero.
Habitually late localities start nearly as early as the others, then peak and
finish late. This is a **common start** for the persistent part of lateness,
the mirror image of the common-end hypothesis, and it produces the long rise
of late localities found in @sec-asym.

The two results are compatible. Habitual lateness is only part of a
locality's departure in a given year: the slope of the median on lateness is
`r sprintf("%.2f", slget("p50", "slope"))`
(`r sprintf("%.2f", slget("p50", "slope_lo"))` to
`r sprintf("%.2f", slget("p50", "slope_hi"))`). The remainder, which is
specific to the year, moves the whole season and keeps the spread profile flat.

What this rules out, and what it leaves:

- **A regional end to flight is ruled out.** Late tails are not shared and
  follow each locality's own timing fully.
- **Symmetric regional mixing is ruled out.** The late tails of early
  localities are not pulled toward the regional dates.
- **Left standing: an early fraction of the catch at late localities that
  flies on the regional schedule.** It could be moths arriving from earlier
  areas before the local population emerges, which would be dispersal in one
  direction only. It could equally be local heterogeneity: warm pockets around
  a cool site that emerge with the rest of the region.

Geography does not separate the two here. Habitual lateness is barely related
to latitude and longitude (R^2^ = `r sprintf("%.2f", geo_r2)` over
`r nrow(loc_late)` localities with at least three years). The two most
habitually early localities, Greenbush and Passadumkeag, are neighbours in the
Penobscot valley, and the late ones are scattered across the state. A
dispersal source would have to be identified from the timing of the early
catches themselves, night by night, which is a question for the daily
analysis.


## Decision: the season is treated as symmetric {#sec-asym-decision}

```{r}
#| label: shape-data
shp <- readRDS(file.path("data", "processed", "seasonal", "asymmetry.rds"))$data |>
  mutate(ratio = width50 / width90)
shape_theory <- c(
  normal   = (qnorm(0.75) - qnorm(0.25)) / (qnorm(0.95) - qnorm(0.05)),
  logistic = (qlogis(0.75) - qlogis(0.25)) / (qlogis(0.95) - qlogis(0.05)),
  laplace  = log(2) / log(10))
set.seed(1978)
shape_ci <- stats::quantile(replicate(2000, stats::median(sample(shp$ratio, replace = TRUE))),
                            c(0.025, 0.975), names = FALSE)
```

The typical season is symmetric (@sec-asym), its asymmetry is idiosyncratic
to the locality-year, and the one systematic pattern — the early start of
habitually late localities (@sec-asym-mech) — is a within-year contrast that a
model of the whole network does not need. The seasonal model therefore treats
the flight season as **symmetric about its median**. Asymmetry is not modelled.

A symmetric season still needs a shape, and the ratio of the central 50% to
the central 90% of the catch identifies it without fitting anything. For a
normal curve the ratio is `r sprintf("%.3f", shape_theory[["normal"]])`, for a
logistic `r sprintf("%.3f", shape_theory[["logistic"]])` and for a Laplace
(double exponential) `r sprintf("%.3f", shape_theory[["laplace"]])`. The
observed median is `r sprintf("%.3f", median(shp$ratio))`
(`r sprintf("%.3f", shape_ci[1])` to `r sprintf("%.3f", shape_ci[2])` over
`r nrow(shp)` locality-years), and `r sprintf("%.3f", median(shp$ratio[shp$season_total >= 3000]))`
in the seasons of 3,000 moths or more, where the percentiles are most precise.
The season is more peaked, with heavier tails, than a normal curve. A logistic
is consistent with the data as a whole, and the best-sampled seasons lie
between the logistic and the Laplace.


# A seasonal predictor {#sec-predictor}

```{r}
#| label: predictor-data
sp  <- readRDS(file.path("data", "processed", "seasonal", "seasonal_predictor.rds"))
sps <- sp$scores; spo <- sp$oracle
LOYO <- "leave-one-year-out"; LOLO <- "leave-one-locality-out"
sget <- function(sch, dr, wm, sh, int, col = "log_score") {
  sps[[col]][sps$scheme == sch & sps$driver == dr & sps$width_model == wm &
               sps$shape == sh & sps$integrate == int]
}
oget <- function(dr, o, col = "log_score") spo[[col]][spo$driver == dr & spo$oracle == o]
## Error of the centre for a new year, weighted by the size of the catch: the
## log score per moth gives each season a weight proportional to its catch.
cerr <- readRDS(file.path("data", "processed", "seasonal", "median_vs_pupation.rds"))$cv |>
  filter(model == "split", scheme == LOYO) |>
  left_join(season |> select(flightID, season_total), by = "flightID") |>
  group_by(driver) |>
  summarise(mae = mean(abs(pred - obs)),
            mae_moth = stats::weighted.mean(abs(pred - obs), season_total),
            .groups = "drop")
ceget <- function(dr, col) cerr[[col]][cerr$driver == dr]
```

`seasonal_dynamics/R/07_seasonal_predictor.R` combines the three results above
into a prediction of how a season's catch is spread over its nights. This
section develops it with the centre of @sec-median, predicted pupation;
@sec-centre then replaces that centre with BioSIM flight, which is the
version adopted (@sec-adopted). The components are:

- **centre**: the median catch date from the split model of @sec-median, the
  network mean of predicted pupation plus the locality's departure from it;
- **width**: the central 90% of the season from the season total
  (@sec-width), fitted to widths with their sampling bias removed, so that the
  curve describes the season and the counting noise is left to the nights;
- **shape**: symmetric about the centre (@sec-asym-decision), with its scale
  set so that its central 90% equals the predicted width.

The share of night *d* is the curve's mass between *d* - 1 and *d*, which puts
its median on the same scale as the observed one. A small uniform floor (a
fraction `r sp$summary$eps` of the season, spread over its nights) keeps every
night above zero. Shares are then normalised over the record's counted nights.
Every component is refitted on the training data of each fold, leaving out one
year at a time and, separately, one locality at a time. The score is the log
score per moth: the summed log share of the night on which each moth was
caught, divided by the number of moths. It is the log-likelihood of the
nightly catch given its total, and higher is better.

## The centre is uncertain, and the curve must say so

```{r}
#| label: tbl-predictor-scores
#| tbl-cap: "Log score per moth of the seasonal predictor (logistic shape, width from the season total)"
tibble::tibble(
  Driver = rep(c("BioSIM pupation", "bayessbw pupation"), each = 2),
  Centre = rep(c("placed as if known", "averaged over its uncertainty"), 2),
  `New year` = c(sget(LOYO, "BioSIM pupation", "catch", "logistic", FALSE),
                 sget(LOYO, "BioSIM pupation", "catch", "logistic", TRUE),
                 sget(LOYO, "bayessbw pupation", "catch", "logistic", FALSE),
                 sget(LOYO, "bayessbw pupation", "catch", "logistic", TRUE)),
  `New locality` = c(sget(LOLO, "BioSIM pupation", "catch", "logistic", FALSE),
                     sget(LOLO, "BioSIM pupation", "catch", "logistic", TRUE),
                     sget(LOLO, "bayessbw pupation", "catch", "logistic", FALSE),
                     sget(LOLO, "bayessbw pupation", "catch", "logistic", TRUE)),
  `Median error, new year, d` = c(rep(sget(LOYO, "BioSIM pupation", "catch", "logistic", FALSE, "mae_median"), 2),
                                  rep(sget(LOYO, "bayessbw pupation", "catch", "logistic", FALSE, "mae_median"), 2))
) |>
  mutate(across(c(`New year`, `New locality`), ~ sprintf("%.3f", .x)),
         `Median error, new year, d` = sprintf("%.2f", `Median error, new year, d`)) |>
  kable(align = c("l", "l", "r", "r", "r"))
```

The predicted centre misses the observed median by about four days on
average. A curve placed as if its centre were exact is then too confident: a
season that peaks four days from the prediction falls on nights the curve has
given almost nothing. The predictive curve is the season curve **averaged over
the uncertainty of its centre**: a normal error whose SD is the residual SD of
the centre model in the training data, about
`r sprintf("%.1f", sget(LOYO, "BioSIM pupation", "catch", "logistic", TRUE, "sigma"))`
days. This raises the log score for a new year from
`r sprintf("%.3f", sget(LOYO, "BioSIM pupation", "catch", "logistic", FALSE))` to
`r sprintf("%.3f", sget(LOYO, "BioSIM pupation", "catch", "logistic", TRUE))` with
BioSIM, and is the largest single gain in the predictor.

With that done, **BioSIM is the better driver** for nightly shares, for a new
year (`r sprintf("%.3f", sget(LOYO, "BioSIM pupation", "catch", "logistic", TRUE))`
against `r sprintf("%.3f", sget(LOYO, "bayessbw pupation", "catch", "logistic", TRUE))`)
and for a new locality
(`r sprintf("%.3f", sget(LOLO, "BioSIM pupation", "catch", "logistic", TRUE))` against
`r sprintf("%.3f", sget(LOLO, "bayessbw pupation", "catch", "logistic", TRUE))`). This
holds although bayessbw has the slightly smaller median error per locality-year
(`r sprintf("%.2f", ceget("bayessbw", "mae"))` against
`r sprintf("%.2f", ceget("BioSIM", "mae"))` days). The log score per moth
weighs each season by its catch, and weighted that way the ranking reverses:
BioSIM's median error is `r sprintf("%.2f", ceget("BioSIM", "mae_moth"))` days
against `r sprintf("%.2f", ceget("bayessbw", "mae_moth"))` for bayessbw. BioSIM
places the large seasons better, and those are where most of the moths are.

## What the components are worth

```{r}
#| label: tbl-predictor-oracle
#| tbl-cap: "Log score per moth for a new year (BioSIM pupation, logistic, centre placed as if known) when one or both components are replaced by the observed value"
spo |>
  filter(driver == "BioSIM pupation") |>
  transmute(`Given the observed` = recode(oracle, none = "nothing (the predictor)",
                                          width = "width", centre = "centre",
                                          both = "centre and width"),
            `Log score per moth` = sprintf("%.3f", log_score)) |>
  kable(align = c("l", "r"))
```

Almost all the room for improvement is in the centre. Knowing the observed
median catch date would raise the log score from
`r sprintf("%.3f", oget("BioSIM pupation", "none"))` to
`r sprintf("%.3f", oget("BioSIM pupation", "centre"))`; knowing the width as well adds
only `r sprintf("%.3f", oget("BioSIM pupation", "both") - oget("BioSIM pupation", "centre"))`.
Knowing the observed width *alone* makes the prediction worse
(`r sprintf("%.3f", oget("BioSIM pupation", "width"))`). When the centre is wrong, a
true, narrow season concentrates the curve on the wrong nights, whereas the
predicted width, drawn toward the average, hedges the miss.

**The width model is unbiased.** Its mean error against the bias-corrected
widths is `r sprintf("%+.2f", sget(LOYO, "BioSIM pupation", "catch", "logistic", TRUE, "width_err_adj"))`
days for a new year. Adding the predicted spread of pupation as a second
covariate changes nothing, so the development models carry no information
about season width beyond what the catch provides.

**Shape matters little once the centre's uncertainty is averaged in.**

```{r}
#| label: tbl-predictor-shape
#| tbl-cap: "Log score per moth by shape, BioSIM pupation, centre averaged over its uncertainty"
sps |>
  filter(driver == "BioSIM pupation", width_model == "catch", integrate) |>
  select(scheme, shape, log_score) |>
  tidyr::pivot_wider(names_from = scheme, values_from = log_score) |>
  transmute(Shape = shape, `New year` = sprintf("%.3f", .data[[LOYO]]),
            `New locality` = sprintf("%.3f", .data[[LOLO]])) |>
  kable(align = c("l", "r", "r"))
```

The three shapes differ by less than 0.01. Placed as if the centre were known,
the logistic clearly beat the normal, as the shape check of
@sec-asym-decision predicted. Averaged over a four-to-five-day uncertainty in
the centre, the differences in the tails no longer matter. The predictor uses
the logistic, which matches the observed shape of the seasons themselves.

## A better centre {#sec-centre}

```{r}
#| label: centre-data
co  <- readRDS(file.path("data", "processed", "seasonal", "centre_options.rds"))
cos <- co$scores; coc <- co$coefs
csget <- function(sp, sch, col) cos[[col]][cos$spec == sp & cos$scheme == sch]
ccget <- function(sp, tm, col = "estimate") coc[[col]][coc$spec == sp & coc$term == tm]
```

Nearly all the predictor's error is in its centre (@sec-predictor).
`seasonal_dynamics/R/08_centre_options.R` tries three alternatives to BioSIM
pupation, each in the same split form (the network mean of the driver that
year plus the locality's departure from it) and scored on the same
`r co$summary$n` locality-years, both alone and inside the full predictor:

- **BioSIM emergence**: the median date of adult emergence (`Emergence_mf`,
  added to `R/03`);
- **BioSIM flight**: the median of BioSIM's flight series;
- **bayessbw pupation plus temperature**: the mean daily temperature at the
  trap over the 14 or 21 days after the locality's predicted median pupation,
  split like the driver into a network-mean and a local-departure term. This
  is the temperature of the pupal and pre-flight period, which a pupation
  date cannot know.

```{r}
#| label: tbl-centre-options
#| tbl-cap: "Centre options. Log score per moth: the full predictor (width from the season total, logistic curve averaged over the centre's uncertainty). MAE: error of the predicted median catch date (days)"
cos |>
  tidyr::pivot_wider(id_cols = spec, names_from = scheme,
                     values_from = c(log_score, mae, rmse)) |>
  arrange(desc(`log_score_leave-one-year-out`)) |>
  transmute(Centre = spec,
            `Log score, new year` = sprintf("%.3f", `log_score_leave-one-year-out`),
            `MAE, new year` = sprintf("%.2f", `mae_leave-one-year-out`),
            `RMSE, new year` = sprintf("%.2f", `rmse_leave-one-year-out`),
            `Log score, new locality` = sprintf("%.3f", `log_score_leave-one-locality-out`),
            `MAE, new locality` = sprintf("%.2f", `mae_leave-one-locality-out`)) |>
  kable(align = c("l", "r", "r", "r", "r", "r"))
```

**BioSIM flight is the best centre** by every measure. For a new year it cuts
the median error from
`r sprintf("%.2f", csget("BioSIM pupation", LOYO, "mae"))` to
`r sprintf("%.2f", csget("BioSIM flight", LOYO, "mae"))` days and raises the
log score from `r sprintf("%.3f", csget("BioSIM pupation", LOYO, "log_score"))`
to `r sprintf("%.3f", csget("BioSIM flight", LOYO, "log_score"))`. For a new
locality the figures are
`r sprintf("%.2f", csget("BioSIM flight", LOLO, "mae"))` days and
`r sprintf("%.3f", csget("BioSIM flight", LOLO, "log_score"))`. Emergence
comes second, between flight and pupation.

The ranking follows how close each driver is to the moment of flight. The
year-level slope rises toward one along the same sequence: pupation
`r sprintf("%.2f", ccget("BioSIM pupation", "biosim_pup_Y"))`, emergence
`r sprintf("%.2f", ccget("BioSIM emergence", "biosim_emerg_Y"))`, flight
`r sprintf("%.2f", ccget("BioSIM flight", "biosim_flight_Y"))`
(`r sprintf("%.2f", ccget("BioSIM flight", "biosim_flight_Y", "lo"))` to
`r sprintf("%.2f", ccget("BioSIM flight", "biosim_flight_Y", "hi"))`). Each
step adds the part of BioSIM's biology, from pupation through the pupal stage
and adult maturation to flight, that the catches confirm. The within-year
slopes stay near 0.1 for every driver
(`r sprintf("%.2f", ccget("BioSIM flight", "biosim_flight_W"))` for flight):
none of them recovers the differences between localities of the same year.

**Temperature after pupation helps bayessbw, but not enough.** Adding the
post-pupation temperature to bayessbw raises its log score for a new year
from `r sprintf("%.3f", csget("bayessbw pupation", LOYO, "log_score"))` to
`r sprintf("%.3f", csget("bayessbw pupation + T14", LOYO, "log_score"))`
(14 days) and `r sprintf("%.3f", csget("bayessbw pupation + T21", LOYO, "log_score"))`
(21 days). That still falls short of BioSIM emergence, let alone flight. The
temperature works through the year: a year warmer by 1 °C after pupation flies
`r sprintf("%.1f", abs(ccget("bayessbw pupation + T14", "T14_Y")))` days
earlier (`r sprintf("%.1f", ccget("bayessbw pupation + T14", "T14_Y", "lo"))`
to `r sprintf("%.1f", ccget("bayessbw pupation + T14", "T14_Y", "hi"))`, 14-day
window). A locality warmer than the others that year flies only
`r sprintf("%.2f", abs(ccget("bayessbw pupation + T14", "T14_W")))` days
earlier per degree
(`r sprintf("%.2f", ccget("bayessbw pupation + T14", "T14_W", "lo"))` to
`r sprintf("%.2f", ccget("bayessbw pupation + T14", "T14_W", "hi"))`). The
local response is again the missing piece. BioSIM, which simulates the pupal
stage and adult maturation explicitly, already does better with that
temperature than an empirical correction added to bayessbw.

```{r}
#| label: fig-centre-options
#| fig-cap: "Centre options for the seasonal predictor: log score per moth (left, higher is better) and error of the predicted median catch date (right), for a new year and a new locality."
#| fig-width: 9
#| fig-height: 3.2
lvl <- cos |> filter(scheme == LOYO) |> arrange(log_score) |> pull(spec)
cd <- cos |> mutate(spec = factor(spec, lvl),
                    scheme = recode(scheme, `leave-one-year-out` = "new year",
                                    `leave-one-locality-out` = "new locality"))
p1 <- ggplot(cd, aes(log_score, spec, shape = scheme)) +
  geom_point(size = 2) + scale_shape_manual(values = c(16, 1), name = NULL) +
  labs(x = "Log score per moth", y = NULL) +
  theme_minimal(base_size = 10) + theme(panel.grid.minor = element_blank(),
                                        legend.position = "none")
p2 <- ggplot(cd, aes(mae, spec, shape = scheme)) +
  geom_point(size = 2) + scale_shape_manual(values = c(16, 1), name = NULL) +
  labs(x = "Error of the median catch date (days)", y = NULL) +
  theme_minimal(base_size = 10) + theme(panel.grid.minor = element_blank(),
                                        axis.text.y = element_blank())
p1 + p2
```

## The adopted predictor {#sec-adopted}

```{r}
#| label: adopted-data
fl_fit <- sp$fits[["BioSIM flight"]]
```

The predictor now uses BioSIM flight as its centre, and the rest is
unchanged: width from the season total, a logistic curve averaged over the
uncertainty of its centre.

```{r}
#| label: tbl-adopted
#| tbl-cap: "The adopted predictor (BioSIM flight centre, width from the season total, logistic curve), against the pupation centre it replaces"
tibble::tibble(
  Centre = c("BioSIM flight", "BioSIM pupation"),
  `Log score, new year` = c(sget(LOYO, "BioSIM flight", "catch", "logistic", TRUE),
                            sget(LOYO, "BioSIM pupation", "catch", "logistic", TRUE)),
  `Log score, new locality` = c(sget(LOLO, "BioSIM flight", "catch", "logistic", TRUE),
                                sget(LOLO, "BioSIM pupation", "catch", "logistic", TRUE)),
  `Median error, new year, d` = c(sget(LOYO, "BioSIM flight", "catch", "logistic", TRUE, "mae_median"),
                                  sget(LOYO, "BioSIM pupation", "catch", "logistic", TRUE, "mae_median")),
  `Centre uncertainty, d` = c(sget(LOYO, "BioSIM flight", "catch", "logistic", TRUE, "sigma"),
                              sget(LOYO, "BioSIM pupation", "catch", "logistic", TRUE, "sigma"))
) |>
  mutate(across(starts_with("Log"), ~ sprintf("%.3f", .x)),
         across(c(`Median error, new year, d`, `Centre uncertainty, d`), ~ sprintf("%.2f", .x))) |>
  kable(align = c("l", "r", "r", "r", "r"))
```

Fitted on all `r fl_fit$n` locality-years, the centre is

  median catch date = `r sprintf("%.1f", fl_fit$centre[["(Intercept)"]])` +
  `r sprintf("%.3f", fl_fit$centre[["xY"]])` x (network mean of BioSIM's median flight date) +
  `r sprintf("%.3f", fl_fit$centre[["xW"]])` x (the locality's departure from it),

with an uncertainty of `r sprintf("%.1f", fl_fit$sigma)` days, and the central
90% of the season is

  width = `r sprintf("%.1f", fl_fit$width[["(Intercept)"]])`
  `r sprintf("%+.2f", fl_fit$width[["log_total"]])` x log10(season total) days. The lessons of @sec-predictor carry over to the new centre.
Averaging over the centre's uncertainty is still worth most: the score is
`r sprintf("%.3f", sget(LOYO, "BioSIM flight", "catch", "logistic", FALSE))`
for a new year with the centre placed as if known. Supplying the observed
width alone still makes it worse
(`r sprintf("%.3f", oget("BioSIM flight", "width"))`).

## Held-out predictions

```{r}
#| label: fig-predictor-examples
#| fig-cap: "Held-out predictions of the adopted predictor (BioSIM flight centre, new year: each season predicted from a model fitted without its year). Line: expected catch; points: observed catches, plus one on a log scale. The four seasons of at least 1,000 moths shown are those whose error in the median catch date is closest to the 25th, 50th, 75th and 90th percentiles of that error."
#| fig-width: 9
#| fig-height: 5.5
ex <- sp$nights |>
  group_by(flightID) |>
  mutate(total = sum(moths), expected = share * total) |>
  ungroup() |>
  left_join(season |> select(flightID, locality, year, p50), by = "flightID")
centre_err <- sp$nights |> distinct(flightID, mu) |>
  left_join(season |> select(flightID, p50, season_total), by = "flightID") |>
  filter(season_total >= 1000) |>
  mutate(err = abs(mu - p50))
pick <- sapply(stats::quantile(centre_err$err, c(0.25, 0.5, 0.75, 0.9)),
               function(q) centre_err$flightID[which.min(abs(centre_err$err - q))])
ex |>
  filter(flightID %in% pick) |>
  left_join(centre_err |> select(flightID, err), by = "flightID") |>
  mutate(panel = sprintf("%s %d (median off by %.1f d)",
                         sub(", Maine$", "", locality), year, err)) |>
  ggplot(aes(doy)) +
  geom_line(aes(y = expected + 1), colour = "black", linewidth = 0.6) +
  geom_point(aes(y = moths + 1), colour = "grey40", size = 1.2) +
  scale_y_log10(labels = scales::label_number(big.mark = ",")) +
  facet_wrap(~ panel, scales = "free") +
  labs(x = "Day of year", y = "Moths per night + 1") +
  theme_minimal(base_size = 9) +
  theme(panel.grid.minor = element_blank())
```

## Using the predictor

`seasonal_dynamics/R/functions/predict_season.R` holds the models fitted on
all `r sp$summary$n` locality-years, with BioSIM flight as the default
driver:

```r
source("seasonal_dynamics/R/functions/predict_season.R")

predict_season(
  driver_p50         = 190.4,   # BioSIM median flight date at the locality
  driver_p50_network = 189.0,   # its mean over the trap network that year
  season_total       = 5000,    # moths caught over the season
  nights_doy         = 180:215  # nights for which shares are wanted
)
```

It returns each night's share of the season and its expected catch, with the
predicted centre, width and centre uncertainty as attributes. It needs the
network mean of the driver for the year, because that carries most of the
skill. With `interval = "nightly"` it also returns a prediction interval for a
single night's count, from the dispersion term of @sec-dispersion. The
pupation drivers remain available through `driver = "BioSIM pupation"` or
`"bayessbw pupation"`.

# The local component {#sec-local}

```{r}
#| label: local-data
lc  <- readRDS(file.path("data", "processed", "seasonal", "local_component.rds"))
lcs <- lc$summary; lcg <- lcs$gradients
lvc <- lc$variance; lsc <- lc$scores; lco <- lc$coefs
vget <- function(m, comp, col = "sd") lvc[[col]][lvc$model == m & lvc$component == comp]
lget <- function(sp, sch, col) lsc[[col]][lsc$spec == sp & lsc$scheme == sch]
cget <- function(sp, tm, col = "estimate") lco[[col]][lco$spec == sp & lco$term == tm]
f1 <- function(x) sprintf("%.1f", x); f2 <- function(x) sprintf("%.2f", x)
f3 <- function(x) sprintf("%.3f", x)
```

The adopted centre follows the differences between years closely and the
differences between localities of the same year hardly at all (slope
`r f3(fl_fit$centre[["xW"]])` on BioSIM's local departure). This section asks
what the remaining local part of the median catch date is made of, and
whether anything available at prediction time explains it
(`seasonal_dynamics/R/09_local_component.R`, the same
`r lcs$n` locality-years, `r lcs$n_loc` localities, `r lcs$n_loc_multi` of
them trapped in two years or more).

## What the residual is made of

The residuals of the centre model were split by a mixed model into a year
part, a persistent locality part (habitual lateness) and a locality-year
remainder. The sampling variance of each observed median was measured by
resampling the record's own catch `r lcs$n_boot` times; subtracting it from
the remainder leaves the transient part, a real departure of that locality in
that year.

```{r}
#| label: tbl-local-variance
#| tbl-cap: "Variance components of the median catch date, before and after the adopted centre (SD in days and share of the total). Interval: parametric bootstrap of the residual model, as SD"
sd_ci <- function(comp) {
  lo <- vget("centre residual", comp, "lo"); hi <- vget("centre residual", comp, "hi")
  if (is.na(lo)) "" else sprintf("%.2f to %.2f", sqrt(max(lo, 0)), sqrt(hi))
}
comps <- c("year", "locality", "transient", "sampling")
tibble::tibble(
  Component = c("Year", "Locality (persistent)", "Locality-year (transient)", "Sampling"),
  `Observed median, SD` = f2(sapply(comps, function(k) vget("observed median", k))),
  `Share` = f2(sapply(comps, function(k) vget("observed median", k, "share"))),
  `Centre residual, SD` = f2(sapply(comps, function(k) vget("centre residual", k))),
  `95% interval` = sapply(comps, sd_ci),
  ` Share` = f2(sapply(comps, function(k) vget("centre residual", k, "share")))
) |>
  kable(align = c("l", "r", "r", "r", "r", "r"))
```

Three things stand out.

- **Sampling error is negligible.** The median of a season is measured to
  within `r f2(lcs$samp_sd)` days (SD), so the residual is not count noise.
- **The driver removes the year and nothing else.** The year SD falls from
  `r f2(vget("observed median", "year"))` to
  `r f2(vget("centre residual", "year"))` days. The locality and transient
  components are the same before and after it.
- **Most of what is left is transient.** A locality departs from its year by
  `r f2(vget("centre residual", "transient"))` days (SD) in a way it does not
  repeat in other years. Habitual lateness is real but smaller
  (`r f2(vget("centre residual", "locality"))` days). It is the same
  property that @sec-asym-mech found shifting a late locality's peak and end
  but not its start.

## Is the local part spatially structured?

Within a year, the residuals of two localities close together agree, and
those of two localities far apart disagree (@fig-local, centre). The
correlation is `r f2(lc$correlogram$r[1])` under 25 km (only
`r lc$correlogram$n_pairs[1]` pairs), near
`r f2(mean(lc$correlogram$r[2:3]))` between 25 and 100 km, and
`r f2(tail(lc$correlogram$r, 1))` beyond 200 km. Centring within a year alone
would give `r f2(mean(lc$correlogram$null))`. The same locality in two
different years correlates at only `r f2(lcs$same_loc_r)`. The local
component is therefore a **regional pattern that changes from year to year**,
not a fixed property of the sites.

A plane in latitude and longitude, fitted separately to each of the
`r lcg$n_years` years with at least six localities, explains
`r sprintf("%.0f%%", 100 * lcg$share)` of the within-year residual variance,
against `r sprintf("%.0f%%", 100 * lcg$null_mean)` when the residuals are
permuted among the localities of a year (p = `r f3(lcg$p)`). Most of that
pattern is a north-south gradient whose strength varies between years:

- **Latitude.** The observed gradient averages
  `r sprintf("%+.2f", lcg$obs_lat_mean)` days per degree north and varies
  between years with an SD of `r f2(lcg$obs_lat_sd_true)` days per degree
  once the sampling error of each year's slope is removed. BioSIM's gradient
  averages `r sprintf("%+.2f", lcg$biosim_lat_mean)`, about
  `r sprintf("%.0f", lcg$biosim_lat_mean / lcg$obs_lat_mean)` times steeper, and
  its year-to-year changes are unrelated to the observed ones
  (r = `r f2(lcg$r_lat)`, @fig-local, right).
- **Longitude.** The observed gradient
  (`r sprintf("%+.2f", lcg$obs_lon_mean)` days per degree) barely changes
  between years (SD `r f2(lcg$obs_lon_sd_true)` net of sampling). BioSIM
  follows its year-to-year changes partly (r = `r f2(lcg$r_lon)`).

BioSIM driven by CaSR temperatures thus predicts a strong, stable
north-south progression of flight. The catches show a weak one that changes
from year to year. That is why the fitted slope on BioSIM's local departure
is close to zero: the model's local signal is mostly a gradient the catches
do not have.

## Can it be predicted?

Each candidate term was added to the adopted centre and scored like the
centre options (@sec-centre). Every covariate entered as a departure from its
year mean, so it could only act on the local component. The candidates were:

- **geography**: latitude, longitude and elevation;
- **temperature**: the local anomaly of mean CaSR temperature over
  1 May to 30 June (larval development) or 15 June to 15 July (pupal and
  pre-flight period), two windows fixed in advance;
- **catch**: the local anomaly of the season total, already an input of the
  predictor;
- **habitual lateness**: a random locality intercept, the locality's shrunken
  mean residual in its training years. It is known only for localities
  trapped before, so it can help a new year but not a new locality. A
  year-adjusted version estimates it net of the year errors.

```{r}
#| label: tbl-local-scores
#| tbl-cap: "Candidate local terms added to the adopted centre. Gain: change in log score per moth of the full predictor relative to the adopted centre (positive is better). MAE: error of the median catch date; per moth: the same weighted by the season total. r within: correlation of predicted and observed departures from the year mean"
lsc |>
  tidyr::pivot_wider(id_cols = spec, names_from = scheme,
                     values_from = c(gain, mae, mae_moth, r_within)) |>
  arrange(desc(`gain_leave-one-year-out`)) |>
  transmute(`Centre` = spec,
            `Gain, new year` = sprintf("%+.3f", `gain_leave-one-year-out`),
            `MAE` = f2(`mae_leave-one-year-out`),
            `MAE per moth` = f2(`mae_moth_leave-one-year-out`),
            `r within` = sprintf("%+.2f", `r_within_leave-one-year-out`),
            `Gain, new locality` = sprintf("%+.3f", `gain_leave-one-locality-out`),
            ` MAE` = f2(`mae_leave-one-locality-out`)) |>
  kable(align = c("l", "r", "r", "r", "r", "r", "r"))
```

**No local term improves the predictor.** None raises the log score by more
than `r f3(max(lsc$gain))`, and most lower it.

- **Habitual lateness** is the only term with a visible signal. For a new
  year it lowers the median error from
  `r f2(lget("adopted (xY + xW)", LOYO, "mae"))` to
  `r f2(lget("+ habitual lateness", LOYO, "mae"))` days and doubles the
  within-year correlation
  (`r f2(lget("adopted (xY + xW)", LOYO, "r_within"))` to
  `r f2(lget("+ habitual lateness", LOYO, "r_within"))`). But the gain is in
  small seasons: weighted by the catch, the error rises from
  `r f2(lget("adopted (xY + xW)", LOYO, "mae_moth"))` to
  `r f2(lget("+ habitual lateness", LOYO, "mae_moth"))` days, and the log
  score, which counts every moth, falls. Separating the year errors from the
  locality intercept makes it worse, not better. Habitual lateness is not
  geographic either: latitude, longitude and elevation explain
  `r sprintf("%.0f%%", 100 * lcs$hab_geo_r2)` of it over the
  `r lcs$n_hab3` localities trapped in three years or more.
- **Local temperature** has no detectable effect: a locality warmer than its
  year by 1 °C over 15 June to 15 July flies
  `r sprintf("%+.2f", cget("+ temperature Jun-Jul", "T_JunJul_W"))` days
  later (`r sprintf("%+.2f", cget("+ temperature Jun-Jul", "T_JunJul_W", "lo"))`
  to `r sprintf("%+.2f", cget("+ temperature Jun-Jul", "T_JunJul_W", "hi"))`).
  The interval includes zero, and the sign is the opposite of development.
  The local temperature anomalies are small (SD
  `r f2(lcs$temp_W_sd[["T_JunJul_W"]])` °C) and carry the same CaSR
  information that BioSIM's local departure already does not use.
- **Catch** is the only term whose coefficient excludes zero: a locality
  catching ten times more than its year flies
  `r f2(abs(cget("+ catch", "log_total_W")))` days earlier
  (`r sprintf("%+.2f", cget("+ catch", "log_total_W", "lo"))` to
  `r sprintf("%+.2f", cget("+ catch", "log_total_W", "hi"))`). It does not
  predict out of sample.
- **Geography** adds parameters and nothing else.

```{r}
#| label: fig-local
#| fig-cap: "The local component of the median catch date. Left: variance components before and after the adopted centre. Centre: correlation of the within-year residuals of two localities trapped in the same year, by distance (bars: 95% intervals from resampling years; dashed line: the correlation produced by centring within a year alone). Right: each year's north-south gradient of the median catch date, observed against BioSIM flight (years with six localities or more; dashed line: equality)."
#| fig-width: 9
#| fig-height: 3.4
library(patchwork)
vd <- lvc |>
  mutate(model = factor(model, c("centre residual", "observed median")),
         component = factor(component, c("year", "locality", "transient", "sampling")))
pa <- ggplot(vd, aes(pmax(var, 0), model, fill = component)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = c(year = "grey20", locality = "grey50",
                               transient = "grey75", sampling = "grey92"), name = NULL) +
  labs(x = expression(Variance~(days^2)), y = NULL) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank(),
        legend.position = "bottom", legend.key.size = grid::unit(0.3, "cm"))
pb <- ggplot(lc$correlogram, aes(km, r)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_hline(yintercept = mean(lc$correlogram$null), linetype = "dashed") +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.04) +
  geom_point(size = 2) +
  scale_x_log10() +
  labs(x = "Distance between localities (km)", y = "Correlation of residuals") +
  theme_minimal(base_size = 10) + theme(panel.grid.minor = element_blank())
pc <- ggplot(lc$gradients, aes(biosim_lat, obs_lat)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_abline(linetype = "dashed") +
  geom_point(size = 2) +
  coord_equal() +
  labs(x = expression(BioSIM~flight~(days~per~degree*N)),
       y = expression(Observed~(days~per~degree*N))) +
  theme_minimal(base_size = 10) + theme(panel.grid.minor = element_blank())
pa + pb + pc + plot_layout(widths = c(1.1, 1, 1))
```

## Conclusion

The adopted predictor stays as it is. Its centre uncertainty
(`r f1(fl_fit$sigma)` days) is mostly this local component, and that is an
honest measure of what can be said about a locality in a given year from a
development model, its temperatures and its catch.

What the local component is remains open. Two explanations are consistent
with the results so far:

- **Temperature at the wrong scale.** CaSR temperature at about 10 km may not
  resolve the conditions that set development at a trap. A test would need
  station records near the traps.
- **Moth movement.** Adult dispersal across the region would flatten the
  north-south gradient of flight relative to local development, as observed
  (`r sprintf("%+.2f", lcg$obs_lat_mean)` against
  `r sprintf("%+.2f", lcg$biosim_lat_mean)` days per degree). It would also
  make the pattern change with each year's winds rather than with
  temperature, and would let late localities start their season as early as
  the rest (@sec-asym-mech). A test would need the nightly series: whether a
  locality's early catches track the regional flight rather than its own
  development.

# Nightly dispersion {#sec-dispersion}

```{r}
#| label: dispersion-data
nd  <- readRDS(file.path("data", "processed", "seasonal", "nightly_dispersion.rds"))
nds <- nd$summary
```

Everywhere above that a season's nights were resampled (@sec-width,
@sec-asym, @sec-asym-mech), moths were drawn as if independent of one
another, i.e. a night's count as Binomial(season total, predicted share). The
*Daily dynamics* analysis found actual nights far more variable than that
(an NB2 fit around its own nightly mean curve, `daily_dynamics/R/11_count_distribution.R`).
This section asks the same question of the predictor of @sec-predictor and
@sec-centre: around the *predicted* share of a night, not its observed one.

`seasonal_dynamics/R/10_nightly_dispersion.R` treats a night's count as
Beta-Binomial(season total, predicted share, θ), the natural two-parameter
family whose variance is

$$\mathrm{Var}(Y_d) = n\,p_d(1-p_d)\,\frac{n+\theta}{1+\theta}$$

the multinomial variance as θ &rarr; &infin;, inflated as it falls. It is the
Dirichlet-multinomial seen one night at a time, so a season's nights still
sum to its observed total. θ is fitted by maximum likelihood, as the only
free parameter, against the predicted share already scored in @sec-predictor:
the leave-one-year-out prediction of the adopted predictor (BioSIM flight
centre, @sec-centre), i.e. the share a user of `predict_season()` would
actually be holding, not an in-sample fit.

```{r}
#| label: tbl-dispersion-fit
#| tbl-cap: "A night's count around its predicted share: multinomial (independence) against beta-binomial, `r nds$n_nights` nights over `r nds$n_seasons` locality-years"
tibble::tibble(
  Model = c("Multinomial (independence)", "Beta-binomial"),
  theta = c("Inf (fixed)", sprintf("%.2f", nds$theta)),
  `RMS standardised residual` = sprintf("%.2f", c(nds$rmse_z_mult_pooled, nds$rmse_z_bb_pooled))
) |>
  kable(align = c("l", "r", "r"), col.names = c("Model", "\u03B8", "RMS standardised residual"))
```

The fitted dispersion is θ = `r sprintf("%.2f", nds$theta)`. A
likelihood-ratio test against the multinomial (θ fixed at infinity) is
decisive (`r sprintf("%.0f", nds$lrt_stat)` on 1 df,
p `r if (nds$lrt_p < 1e-10) "< 1e-10" else sprintf("= %.2g", nds$lrt_p)`), and
the standardised residual of a night's count -- which should average to 1 if
the model is scaled correctly -- falls from
`r sprintf("%.1f", nds$rmse_z_mult_pooled)` under the multinomial to
`r sprintf("%.2f", nds$rmse_z_bb_pooled)` under the beta-binomial. Treating
moths as independent draws understates a night's variability by roughly this
factor.

```{r}
#| label: tbl-dispersion-vif
#| tbl-cap: "Variance inflation of a night's count relative to the multinomial, (n + theta) / (1 + theta), at a few season sizes"
nds$vif_table |>
  transmute(`Season total` = scales::label_comma()(season_total),
            `SD inflation (x multinomial)` = sprintf("%.1f", sd_inflation)) |>
  kable(align = c("r", "r"))
```

The inflation grows with the season total, not against it: a bigger season
is not counted more precisely night to night, because the extra moths are
not independent of each other. At a season of 1,000 moths a night's realistic
SD is about `r sprintf("%.0f", nds$vif_table$sd_inflation[nds$vif_table$season_total == 1000])`
times the multinomial's.

```{r}
#| label: fig-dispersion-example
#| fig-cap: "The adopted predictor's curve (black) with a 90% nightly interval (beta-binomial, theta as fitted above, Monte Carlo quantiles), against observed catches (grey points), for the same held-out seasons as fig-predictor-examples"
#| fig-width: 9
#| fig-height: 5.5
set.seed(2026)
bb_interval <- function(n, p, theta, level = 0.90, nsim = 4000) {
  alo <- (1 - level) / 2
  m <- vapply(p, function(pp) {
    draws <- stats::rbinom(nsim, size = n,
                           prob = stats::rbeta(nsim, theta * pp, theta * (1 - pp)))
    stats::quantile(draws, c(alo, 1 - alo), names = FALSE)
  }, numeric(2))
  list(lwr = m[1, ], upr = m[2, ])
}
disp_ex <- ex |>
  filter(flightID %in% pick) |>
  left_join(centre_err |> select(flightID, err), by = "flightID") |>
  mutate(panel = sprintf("%s %d (median off by %.1f d)",
                         sub(", Maine$", "", locality), year, err))
bnd <- disp_ex |>
  group_by(flightID) |>
  reframe(doy = doy, bb_interval(first(total), share, nd$theta) |> as.data.frame())
disp_ex <- disp_ex |> left_join(bnd, by = c("flightID", "doy"))
ggplot(disp_ex, aes(doy)) +
  geom_ribbon(aes(ymin = lwr + 1, ymax = upr + 1), alpha = 0.15) +
  geom_line(aes(y = expected + 1), colour = "black", linewidth = 0.6) +
  geom_point(aes(y = moths + 1), colour = "grey40", size = 1.2) +
  scale_y_log10(labels = scales::label_number(big.mark = ",")) +
  facet_wrap(~ panel, scales = "free") +
  labs(x = "Day of year", y = "Moths per night + 1") +
  theme_minimal(base_size = 9) + theme(panel.grid.minor = element_blank())
```

The band is wide, and mostly for a mundane reason: it is built from the
*predicted* centre, so a season whose observed median misses the prediction
(the usual case, @sec-adopted) sits off-centre inside it rather than outside
it. θ buys back some of the honesty that assuming independence would have
borrowed: holding `predict_season()` to the narrower, multinomial width would
understate how far a single night's count can reasonably land around the
curve's expectation.

`predict_season()` (@sec-adopted) now takes `interval = "nightly"`, returning
`sd`, `lwr` and `upr` for a single night from this model. Nights are **not**
independent of each other under it -- they are one draw from a
Dirichlet-multinomial season -- so summing the per-night intervals overstates
the uncertainty of a multi-night total; nothing here gives that joint
interval.

This leaves the asymmetry question of @sec-asym-mech open rather than
settled: the resampling there also treated moths as independent, and with θ
this small, a run of nights sharing a warm spell could produce apparent skew
that a wider null would absorb. Revisiting @sec-asym-mech with this
dispersion, rather than the multinomial, is unfinished.

# Population levels {#sec-poplevel}

```{r}
#| label: pop-data
pc  <- readRDS(file.path("data", "processed", "seasonal", "population_classes.rds"))
pcs <- pc$summary; pct <- pc$tests; pcf <- pc$fit2; pcv <- pc$variance
pd  <- pc$data
tget <- function(set, col) pct[[col]][pct$set == set]
S_POS <- "positive totals, log10"; S_ALL <- "all totals, log10(total + 1)"
S_EFF <- "daily series >= 20 nights, log10"
bic_pos <- pc$bic |> filter(set == S_POS)
dbic <- max(bic_pos$BIC[bic_pos$G == 2]) - max(bic_pos$BIC[bic_pos$G == 1])
fget <- function(cl, col) pcf[[col]][pcf$class == cl]
eff <- pd |> filter(has_daily, season_total > 0) |> group_by(class) |>
  summarise(nights = median(n_nights), .groups = "drop")
pvar <- function(k, col = "share") pcv[[col]][pcv$component == k]
## Depth of the trough of the fitted mixture: its density at the lowest point
## between the two component means, relative to the lower of its two peaks.
mix_dens <- function(x) pcf$prop[1] * stats::dnorm(x, pcf$mean_log[1], pcf$sd_log[1]) +
  pcf$prop[2] * stats::dnorm(x, pcf$mean_log[2], pcf$sd_log[2])
xg <- seq(pcf$mean_log[1], pcf$mean_log[2], by = 0.005)
trough_x <- xg[which.min(mix_dens(xg))]
peak_lo <- min(stats::optimize(mix_dens, c(min(xg), trough_x), maximum = TRUE)$objective,
               stats::optimize(mix_dens, c(trough_x, max(xg)), maximum = TRUE)$objective)
trough_rel <- mix_dens(trough_x) / peak_lo
```

Can the season totals be split into two classes, low and high population
level? The question has to be asked of **every** record, not only of the
analysis set, because the 99-moth threshold of `R/01` selects on the season
total itself. `R/01` now also writes `trap_season_totals.rds`: all
`r pcs$n` Maine light-trap records of 1968–1989 before the threshold
(`r pcs$n_daily` with a daily series, `r pcs$n_total_only` with a season
total only), `r pcs$n_retained` of which are retained for the analyses
above. `r pcs$n_zero` records caught no moth at all. They form a point mass
that no continuous distribution describes, so they are set aside and
reported with the year table
(`seasonal_dynamics/R/10_population_classes.R`).

Three tests were applied to the log10 season totals:

- Hartigan's **dip test**, which asks whether the distribution has more than
  one mode;
- **Gaussian mixtures** of one to four components, compared by BIC;
- a **bootstrap likelihood-ratio test** (`r pcs$n_boot` resamples) of one
  component against two, and two against three.

```{r}
#| label: tbl-pop-tests
#| tbl-cap: "Tests for two population classes in the log10 season total. Dip: Hartigan's test of unimodality. Mixture: number of Gaussian components preferred by BIC (E: equal, V: unequal variances). LRT: bootstrap likelihood-ratio test"
pct |>
  transmute(Records = set, n,
            `Dip p` = sprintf("%.3f", dip_p),
            `BIC components` = sprintf("%d (%s)", best_G, best_model),
            `LRT 1 vs 2, p` = sprintf("%.3f", lrt_1v2_p),
            `LRT 2 vs 3, p` = ifelse(is.na(lrt_2v3_p), "-", sprintf("%.3f", lrt_2v3_p))) |>
  kable(align = c("l", "r", "r", "r", "r", "r"))
```

**The mixture models favour two components, but there are not two modes.**
On the `r tget(S_POS, "n")` positive totals, BIC prefers two components to
one by `r sprintf("%.0f", dbic)` units, and the bootstrap rejects one
component (p = `r sprintf("%.3f", tget(S_POS, "lrt_1v2_p"))`) but not two
(p = `r sprintf("%.2f", tget(S_POS, "lrt_2v3_p"))` for a third). The fitted mixture has two peaks and a trough between them, at about
`r sprintf("%.0f", round(10^trough_x, -1))` moths (@fig-pop, left). But the
trough is shallow: the density there is
`r sprintf("%.0f%%", 100 * trough_rel)` of the lower peak. The dip test, the
direct test of two modes in the data rather than in a fitted model, does not
reject a single mode (p = `r sprintf("%.3f", tget(S_POS, "dip_p"))`). The
components are:

- **low**: median `r sprintf("%.0f", fget("low", "median_moths"))` moths,
  SD `r sprintf("%.2f", fget("low", "sd_log"))` log units,
  `r sprintf("%.0f%%", 100 * fget("low", "prop"))` of seasons;
- **high**: median `r sprintf("%.0f", fget("high", "median_moths"))` moths,
  SD `r sprintf("%.2f", fget("high", "sd_log"))` log units,
  `r sprintf("%.0f%%", 100 * fget("high", "prop"))` of seasons.

One component is broad and the other narrow. Such a pair also describes a
single skewed distribution, a long tail of small catches below a
concentration of large ones. A mixture can prefer two components without the
data holding two populations, so the evidence for two classes is suggestive,
not conclusive.

**Part of the low tail is trapping effort.** Among records with a daily
series, the log season total correlates
`r sprintf("%.2f", pcs$cor_nights)` with the log number of nights trapped.
Low-class records were trapped for a median of
`r eff$nights[eff$class == "low"]` nights, against
`r eff$nights[eff$class == "high"]` for high-class records. Restricted to the
`r tget(S_EFF, "n")` records trapped `r pcs$min_nights` nights or more, the
distribution is plainly unimodal (dip p = `r sprintf("%.2f", tget(S_EFF, "dip_p"))`).
Including the zero seasons makes the dip test significant
(p `r ifelse(tget(S_ALL, "dip_p") < 0.001, "< 0.001", sprintf("= %.3f", tget(S_ALL, "dip_p")))`),
but only because of the spike of `r pcs$n_zero` zeros. It is not evidence of
a second population.

**The population level is mostly a property of the year.** Of the variance in
log10(total + 1), the year accounts for
`r sprintf("%.0f%%", 100 * pvar("year"))`, the locality for
`r sprintf("%.0f%%", 100 * pvar("locality"))`, and the locality-year
remainder for `r sprintf("%.0f%%", 100 * pvar("locality-year"))`
(@fig-pop, right). The years trace the outbreak cycle continuously rather
than in two states:

- 1968–1970: medians of 12–19 moths, no high seasons;
- 1971–1973: rising;
- 1974–1975 and 1978–1983: most traps in the high class;
- 1984–1986: falling;
- 1987–1989: most traps catching nothing.

```{r}
#| label: fig-pop
#| fig-cap: "Season totals of all Maine light-trap records, 1968-1989, before the 99-moth threshold. Left: distribution of the log10 total of seasons with at least one moth, with the two-component mixture (dashed: components; solid: their sum); vertical lines: the 99-moth threshold (dotted) and the boundary where a season is equally likely to belong to either component (dashed). Centre: the same for records trapped at least 20 nights. Right: every record by year (zero seasons plotted at 0), with the year median."
#| fig-width: 9
#| fig-height: 3.4
library(patchwork)
mix_curve <- function(x) {
  comp <- lapply(seq_len(nrow(pcf)), function(k)
    pcf$prop[k] * stats::dnorm(x, pcf$mean_log[k], pcf$sd_log[k]))
  tibble::tibble(x = x, low = comp[[1]], high = comp[[2]], total = comp[[1]] + comp[[2]])
}
mc <- mix_curve(seq(-0.2, 5.2, by = 0.02))
pos <- pd |> filter(season_total > 0)
vlines <- list(geom_vline(xintercept = log10(100), linetype = "dotted"),
               geom_vline(xintercept = pcs$boundary_log, linetype = "dashed"))
xlab <- expression(log[10]~season~total)
pa <- ggplot(pos, aes(log_total)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 0.25, boundary = 0,
                 fill = "grey80", colour = "white") +
  geom_line(data = mc, aes(x, low), linetype = "dashed") +
  geom_line(data = mc, aes(x, high), linetype = "dashed") +
  geom_line(data = mc, aes(x, total)) + vlines +
  labs(x = xlab, y = "Density", title = sprintf("All seasons with moths (n = %d)", nrow(pos))) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(), plot.title = element_text(size = 9))
pb <- ggplot(filter(pos, has_daily, n_nights >= pcs$min_nights), aes(log_total)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 0.25, boundary = 0,
                 fill = "grey80", colour = "white") + vlines +
  labs(x = xlab, y = NULL,
       title = sprintf("Trapped >= %d nights (n = %d)", pcs$min_nights,
                       sum(pos$has_daily & pos$n_nights >= pcs$min_nights))) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(), plot.title = element_text(size = 9))
pcy <- ggplot(pd, aes(year, log_total1)) +
  geom_jitter(width = 0.2, height = 0, size = 0.8, colour = "grey50") +
  geom_line(data = pc$by_year, aes(year, log10(median_total + 1))) +
  geom_hline(yintercept = log10(100), linetype = "dotted") +
  labs(x = NULL, y = expression(log[10]~(season~total+1)), title = "By year") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(), plot.title = element_text(size = 9))
pa + pb + pcy + plot_layout(widths = c(1, 1, 1.2))
```

**Conclusion.** The evidence that the season totals fall into two natural
classes is weak. There is a shallow trough near
`r sprintf("%.0f", round(10^trough_x, -1))` moths, but it does not
pass the test for two modes, and it disappears when trapping effort is
controlled. The totals read better as a single skewed continuum, most of it
ordered by the year of the outbreak cycle. A two-class split can still be imposed as a convenience. The
mixture boundary, at about `r sprintf("%.0f", round(pcs$boundary_moths, -1))`
moths per season, is the defensible place for it:
`r sprintf("%.0f%%", 100 * pcs$certain_share)` of seasons fall on one side
with posterior probability of at least 0.9. That boundary sits well above the
99-moth threshold of `R/01`: `r pcs$low_above_99` of the
`r pcs$n_retained` retained seasons fall in the low class, and all
`r pcs$high_retained` high seasons are retained. Because a season's level
is largely set by its year, a classification by outbreak phase would carry
most of the same information.

# Next steps

1. **The origin of the local component** (@sec-local), if it matters for the
   use of the predictor: station temperatures near the traps, or the nightly
   series, to tell temperature at the wrong scale from moth movement.
2. ~~**Nightly variability**, if prediction intervals for single nights are
   wanted: a dispersion term on top of the seasonal curve.~~ Done
   (@sec-dispersion): a beta-binomial term, fitted on the predictor's own
   held-out predictions. The resampling used earlier in this analysis
   (@sec-width, @sec-asym, @sec-asym-mech) treats moths as independent draws;
   this section shows that nights are considerably more variable than that,
   consistent with the Daily dynamics analysis's NB2 fit
   (`daily_dynamics/R/11_count_distribution.R`).
