---
title: "Spruce budworm flight phenology in Maine"
subtitle: "Analysis log"
author: "JN Candau"
date: today
date-format: "D MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk:
    comment: "#>"
---

```{r}
#| label: setup
## Paths are written relative to the project root, not to analysis/.
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r}
#| label: data
library(dplyr)
library(knitr)

trap_records <- readRDS(file.path("data", "processed", "trap_records.rds"))
trap_daily   <- readRDS(file.path("data", "processed", "trap_daily.rds"))
```

# Purpose of this document

This is the running log of the analyses of spruce budworm (*Choristoneura
fumiferana*) flight phenology from daily light-trap catches in Maine. Each
section records what was done, the choices that were made and why, and the
state of the data or the model that resulted. It is updated as the work
proceeds; the numbers below are recomputed from the processed data every time
the document is rendered, so they cannot drift away from the files on disk.

Code lives in `R/`, source data in `data/raw/`, processed tables in
`data/processed/`. The project is version-controlled at
<https://github.com/jncandau/Maine_Oct_2026>.

# Trap data

## Source

The captures come from the eastern spruce budworm phenology dataset of
Fournier, Candau et al. (2027), held in the private repository
`jncandau/SBWFieldPhenoDatabase` and copied unmodified into
`data/raw/SBWFieldPhenoDatabase/`. The workbook
`Eastern Spruce Budworm Phenology V1.ods` holds three tables — `Phenology`,
`Flight` and `References` — joined by `referenceID`. Only `Flight` is used
here.

`Flight` is a wide table: one row per trapping record, with the captures of
each night in the columns `mothsDay160` … `mothsDay231` (day of year 160 to
231). A record is a site × year × sampling method × sex combination, and
carries its coordinates, the reported season total `mothCountYear`, and a
pointer to the source publication. Across North America it holds 763 records;
Maine contributes 643 light-trap records and 36 yard-light records.

## Reading and decoding

`R/01_read_trap_data.R` reads the sheet with every column forced to character,
because the day columns mix numbers with two text codes:

| Cell | Meaning | Handling |
|------|---------|----------|
| `-` | missing, no count available | dropped; these cells fill the whole out-of-season range |
| `m` | missing, no count available | kept as a row with `moths = NA` and `status = "missing"` |
| number | nightly capture | kept, `status = "counted"` |

: Day-cell codes in the `Flight` sheet {#tbl-codes tbl-colwidths="[10,40,50]"}

Both codes mean missing data (confirmed 2026-09-26). They are treated
differently only because of where they occur: `-` pads every record out to
DOY 231, so keeping those cells would invent trap-nights, whereas `m` only
ever appears inside a trapping season. Interior `-` cells — those falling
between a record's first and last counted night — are real gaps, and are
counted in `n_gap_nights` rather than lost.

The wide table is then pivoted to one row per record-night, and a calendar
date is derived from the year and the day of year.

## Selection

Four criteria define the working dataset, all set as constants at the top of
`R/01_read_trap_data.R`:

| Criterion | Value | Reason |
|-----------|-------|--------|
| `keep_states` | `ME` | the study area |
| `keep_methods` | `Light trap` | excludes the 36 opportunistic yard-light records |
| `keep_years` | 1968–1989 | overlap with the CaSR v3.2 reanalysis in `data/raw/Weather_Maine` (1968–1991) |
| `min_season_total` | > 99 moths | a smaller catch cannot describe a flight curve |

: Selection criteria {#tbl-selection}

The year window drops 1961–1967 and 2016. The catch threshold removes a
further 230 in-window records; it also removes, without needing a separate
rule, every record that reports only a season total and no daily series, and
all of 1963–1967 and 2016, where no trap anywhere in Maine exceeded 99 moths.

## Pooling of sex-split records

Two 1980 localities report the male and the female catch of one trap as
separate records. Since the trap is the sampling unit, the pairs are summed
night by night before the catch threshold is applied, and the components are
removed. Both pairs sampled identical nights with no missing values, so the
sums are exact.

```{r}
#| label: tbl-pooled
#| tbl-cap: "Pooled sex-split records"
trap_records |>
  filter(!is.na(pooled_from)) |>
  transmute(flightID, `components` = pooled_from, year, locality,
            `nights` = n_nights, `season total` = season_total) |>
  kable()
```

Pooling matters beyond tidiness: at Bare Island neither sex reached the
threshold alone (83 and 78 moths) but the trap caught 161, so the locality
re-enters the dataset.

## The resulting dataset

Two tables are written to `data/processed/`:

- `trap_records.rds` — one row per trapping record, with sampling effort,
  season total and provenance;
- `trap_daily.rds` — one row per record-night that was inside a trapping
  season, holding the count and its `status`.

```{r}
#| label: tbl-overview
#| tbl-cap: "The working dataset at a glance"
tibble::tibble(
  Quantity = c("Trapping records", "Localities", "Years", "Trap-nights with a count",
               "Nights coded `m` (missing)", "Interior gap nights (`-`)",
               "Moths counted", "Season total, median (range)",
               "Nights per record, median (range)"),
  Value = c(
    format(nrow(trap_records), big.mark = ","),
    format(n_distinct(trap_records$locality), big.mark = ","),
    sprintf("%d (%d\u2013%d)", n_distinct(trap_records$year),
            min(trap_records$year), max(trap_records$year)),
    format(sum(trap_daily$status == "counted"), big.mark = ","),
    format(sum(trap_daily$status == "missing"), big.mark = ","),
    format(sum(trap_records$n_gap_nights), big.mark = ","),
    format(sum(trap_daily$moths, na.rm = TRUE), big.mark = ","),
    sprintf("%s (%s\u2013%s)",
            format(median(trap_records$season_total), big.mark = ","),
            format(min(trap_records$season_total), big.mark = ","),
            format(max(trap_records$season_total), big.mark = ",")),
    sprintf("%d (%d\u2013%d)", median(trap_records$n_nights),
            min(trap_records$n_nights), max(trap_records$n_nights))
  )
) |>
  kable(align = c("l", "r"))
```

```{r}
#| label: tbl-per-year
#| tbl-cap: "Sampling effort and catch by year"
trap_daily |>
  group_by(year) |>
  summarise(
    traps       = n_distinct(flightID),
    localities  = n_distinct(locality),
    nights      = sum(status == "counted"),
    `nights/trap` = round(sum(status == "counted") / n_distinct(flightID), 1),
    `first DOY` = min(doy),
    `last DOY`  = max(doy),
    moths       = sum(moths, na.rm = TRUE),
    .groups     = "drop"
  ) |>
  mutate(year = as.character(year)) |>   # a year is a label, not a count
  kable(format.args = list(big.mark = ","))
```

The network is a rotating panel rather than a fixed set of stations: most
localities contribute a handful of years, and the annual number of traps
varies with the outbreak and with the Maine Forest Service's programme.

```{r}
#| label: tbl-site-years
#| tbl-cap: "Localities by the number of years they contribute"
trap_records |>
  group_by(locality) |>
  summarise(years = n_distinct(year), .groups = "drop") |>
  count(years, name = "localities") |>
  arrange(years) |>
  kable()
```

### Data quality

The following checks are run by `R/01_read_trap_data.R` at every read. They
report problems; they never repair them.

```{r}
#| label: tbl-quality
#| tbl-cap: "Quality checks on the retained records"
gaps <- filter(trap_records, n_gap_nights > 0)
miss <- trap_daily |> filter(status == "missing") |> distinct(flightID)
tibble::tibble(
  Check = c("Season total disagrees with the sum of the daily counts",
            "Locality-year with more than one record",
            "Records with nights coded `m`",
            "Records with interior gaps (`-` inside the season)"),
  Records = c(0L, 0L, nrow(miss), nrow(gaps))
) |>
  kable(align = c("l", "r"))
```

```{r}
#| label: tbl-gaps
#| tbl-cap: "Records with unrecorded nights inside their trapping season"
trap_records |>
  filter(n_gap_nights > 0) |>
  transmute(flightID, year = as.character(year), locality,
            season = sprintf("%d\u2013%d", first_doy, last_doy),
            `nights counted` = n_nights, `gap nights` = n_gap_nights,
            `season total` = season_total) |>
  arrange(desc(`gap nights`)) |>
  kable(format.args = list(big.mark = ","))
```

Georgetown Township 1969 is the one record where the gaps dominate: 23
unrecorded nights against 21 counted, over a nominal season of 44 days. The
1980 pooled records were operated on an irregular schedule rather than
interrupted. All other retained records are continuous between their first and
last night.

## Minimum effort for phenology

A season total above 99 moths says nothing about whether the trap was running
long enough, or at the right time, to describe a flight curve. The rule below
matches the requirement to the statistic: for a percentile to be identifiable,
the sampling window has to bracket it. If more than 5% of a season's catch
falls on the first night, the 5th percentile sits at or before that night and
cannot be located; the same argument applies to p95 and the last night.

| Criterion | Threshold | What it protects against |
|-----------|-----------|--------------------------|
| `min_nights` | ≥ 15 counted nights | percentiles interpolated from too few points |
| `min_coverage` | counted nights / season span ≥ 0.7 | holes inside the season displacing the cumulative curve |
| `max_tail_share` | ≤ 0.05 of the catch on the first *and* on the last night | a flight period truncated at either end |

: Effort criteria, set as constants in `R/01_read_trap_data.R` {#tbl-effort-rule}

The criteria are recorded, not enforced: `trap_records` carries `coverage`,
`tail_first`, `tail_last` and the logical `usable_phenology`, so an analysis
that needs only the median can relax them.

```{r}
#| label: tbl-usable
#| tbl-cap: "Records excluded from percentile-based phenology, by reason"
trap_records |>
  mutate(reason = case_when(
    usable_phenology                       ~ "usable",
    n_nights < 15                          ~ "too few nights (< 15)",
    coverage < 0.7                         ~ "too gappy (coverage < 0.7)",
    TRUE                                   ~ "flight period truncated (tail > 5%)"
  )) |>
  count(reason, name = "records") |>
  arrange(desc(records)) |>
  kable(align = c("l", "r"))
```

Effort itself is rarely the problem: coverage is exactly 1 for most records
and the median series runs 25 nights. Truncation is what binds, and it is not
a marginal call in the worst cases — East Allagash 1986 took 77% of its season
on the final night, Round Mountain 1972 71% on the first.

```{r}
#| label: tbl-usable-year
#| tbl-cap: "Usable records per year"
trap_records |>
  group_by(year) |>
  summarise(records = n(), usable = sum(usable_phenology),
            localities = n_distinct(locality[usable_phenology]),
            nights = sum(n_nights[usable_phenology]), .groups = "drop") |>
  mutate(year = as.character(year)) |>
  kable(format.args = list(big.mark = ","))
```

The cost is concentrated in the sparse years. Sensitivity of the tail
threshold, holding the other two criteria fixed:

```{r}
#| label: tbl-tail-sensitivity
#| tbl-cap: "Effect of the tail threshold"
bind_rows(lapply(c(0.05, 0.10, 0.15), function(tt) {
  k <- with(trap_records, n_nights >= 15 & coverage >= 0.7 &
              tail_first <= tt & tail_last <= tt)
  tibble::tibble(`tail threshold` = tt,
                 records    = sum(k),
                 localities = n_distinct(trap_records$locality[k]),
                 years      = n_distinct(trap_records$year[k]),
                 `years lost` = paste(setdiff(unique(trap_records$year),
                                              unique(trap_records$year[k])),
                                      collapse = ", "))
})) |>
  kable(align = c("r", "r", "r", "r", "l"))
```

1987 and 1988 have a single trap each, so one failing record erases the year.
Whether that is acceptable depends on the analysis: a pooled site-year model
loses little, a year-by-year trend loses the two terminal years of the
outbreak collapse.

## Open questions

- **1987 and 1988 have no usable record.** Under the current tail threshold
  both years disappear from the phenology set. Relaxing the threshold to 0.10
  restores 1987 at the cost of admitting series whose flight period is
  probably truncated; the decision belongs with the analysis, not the reader.
- **Weather resolution.** The CaSR reanalysis covers every retained locality
  (all within 6.3 km of a grid cell) and every retained year, but air
  temperature is stored 3-hourly, not hourly, unlike the other four variables.
- **Records excluded by the year window.** 1961–1967 and 2016 are outside the
  reanalysis, and are set aside rather than lost: widening `keep_years` and
  setting `min_season_total` to 0 in `R/01_read_trap_data.R` restores the full
  643-record Maine light-trap set.
