---
title: "Spruce budworm flight phenology in Maine"
subtitle: "Analysis log"
author: "JN Candau"
date: today
date-format: "D MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk:
    comment: "#>"
---

```{r}
#| label: setup
## Paths are written relative to the project root, not to analysis/.
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r}
#| label: data
library(dplyr)
library(knitr)
library(lubridate)   # month(), yday(), leap_year() in the weather section

trap_records <- readRDS(file.path("data", "processed", "trap_records.rds"))
trap_daily   <- readRDS(file.path("data", "processed", "trap_daily.rds"))
```

# Purpose of this document

This is the running log of the analyses of spruce budworm (*Choristoneura
fumiferana*) flight phenology from daily light-trap catches in Maine. Each
section records what was done, the choices that were made and why, and the
state of the data or the model that resulted. It is updated as the work
proceeds; the numbers below are recomputed from the processed data every time
the document is rendered, so they cannot drift away from the files on disk.

Code lives in `R/`, source data in `data/raw/`, processed tables in
`data/processed/`. The project is version-controlled at
<https://github.com/jncandau/Maine_Oct_2026>.

# Trap data {#sec-traps}

## Source

The captures come from the eastern spruce budworm phenology dataset of
Fournier, Candau et al. (2027), held in the private repository
`jncandau/SBWFieldPhenoDatabase` and copied unmodified into
`data/raw/SBWFieldPhenoDatabase/`. The workbook
`Eastern Spruce Budworm Phenology V1.ods` holds three tables — `Phenology`,
`Flight` and `References` — joined by `referenceID`. Only `Flight` is used
here.

`Flight` is a wide table: one row per trapping record, with the captures of
each night in the columns `mothsDay160` … `mothsDay231` (day of year 160 to
231). A record is a site × year × sampling method × sex combination, and
carries its coordinates, the reported season total `mothCountYear`, and a
pointer to the source publication. Across North America it holds 763 records;
Maine contributes 643 light-trap records and 36 yard-light records.

## Reading and decoding

`R/01_read_trap_data.R` reads the sheet with every column forced to character,
because the day columns mix numbers with two text codes:

| Cell | Meaning | Handling |
|------|---------|----------|
| `-` | missing, no count available | dropped; these cells fill the whole out-of-season range |
| `m` | missing, no count available | kept as a row with `moths = NA` and `status = "missing"` |
| number | nightly capture | kept, `status = "counted"` |

: Day-cell codes in the `Flight` sheet {#tbl-codes tbl-colwidths="[10,40,50]"}

Both codes mean missing data (confirmed 2026-09-26). They are treated
differently only because of where they occur: `-` pads every record out to
DOY 231, so keeping those cells would invent trap-nights, whereas `m` only
ever appears inside a trapping season. Interior `-` cells — those falling
between a record's first and last counted night — are real gaps, and are
counted in `n_gap_nights` rather than lost.

The wide table is then pivoted to one row per record-night, and a calendar
date is derived from the year and the day of year.

## Selection

Four criteria define the working dataset, all set as constants at the top of
`R/01_read_trap_data.R`:

| Criterion | Value | Reason |
|-----------|-------|--------|
| `keep_states` | `ME` | the study area |
| `keep_methods` | `Light trap` | excludes the 36 opportunistic yard-light records |
| `keep_years` | 1968–1989 | overlap with the CaSR v3.2 reanalysis in `data/raw/Weather_Maine` (1968–1991) |
| `min_season_total` | > 99 moths | a smaller catch cannot describe a flight curve |

: Selection criteria {#tbl-selection}

The year window drops 1961–1967 and 2016. The catch threshold removes a
further 230 in-window records; it also removes, without needing a separate
rule, every record that reports only a season total and no daily series, and
all of 1963–1967 and 2016, where no trap anywhere in Maine exceeded 99 moths.

## Pooling of sex-split records

Two 1980 localities report the male and the female catch of one trap as
separate records. Since the trap is the sampling unit, the pairs are summed
night by night before the catch threshold is applied, and the components are
removed. Both pairs sampled identical nights with no missing values, so the
sums are exact.

```{r}
#| label: tbl-pooled
#| tbl-cap: "Pooled sex-split records"
trap_records |>
  filter(!is.na(pooled_from)) |>
  transmute(flightID, `components` = pooled_from, year, locality,
            `nights` = n_nights, `season total` = season_total) |>
  kable()
```

Pooling matters beyond tidiness: at Bare Island neither sex reached the
threshold alone (83 and 78 moths) but the trap caught 161, so the locality
re-enters the dataset.

## The resulting dataset

Two tables are written to `data/processed/`:

- `trap_records.rds` — one row per trapping record, with sampling effort,
  season total and provenance;
- `trap_daily.rds` — one row per record-night that was inside a trapping
  season, holding the count and its `status`.

```{r}
#| label: tbl-overview
#| tbl-cap: "The working dataset at a glance"
tibble::tibble(
  Quantity = c("Trapping records", "Localities", "Years", "Trap-nights with a count",
               "Nights coded `m` (missing)", "Interior gap nights (`-`)",
               "Moths counted", "Season total, median (range)",
               "Nights per record, median (range)"),
  Value = c(
    format(nrow(trap_records), big.mark = ","),
    format(n_distinct(trap_records$locality), big.mark = ","),
    sprintf("%d (%d\u2013%d)", n_distinct(trap_records$year),
            min(trap_records$year), max(trap_records$year)),
    format(sum(trap_daily$status == "counted"), big.mark = ","),
    format(sum(trap_daily$status == "missing"), big.mark = ","),
    format(sum(trap_records$n_gap_nights), big.mark = ","),
    format(sum(trap_daily$moths, na.rm = TRUE), big.mark = ","),
    sprintf("%s (%s\u2013%s)",
            format(median(trap_records$season_total), big.mark = ","),
            format(min(trap_records$season_total), big.mark = ","),
            format(max(trap_records$season_total), big.mark = ",")),
    sprintf("%d (%d\u2013%d)", median(trap_records$n_nights),
            min(trap_records$n_nights), max(trap_records$n_nights))
  )
) |>
  kable(align = c("l", "r"))
```

```{r}
#| label: tbl-per-year
#| tbl-cap: "Sampling effort and catch by year"
trap_daily |>
  group_by(year) |>
  summarise(
    traps       = n_distinct(flightID),
    localities  = n_distinct(locality),
    nights      = sum(status == "counted"),
    `nights/trap` = round(sum(status == "counted") / n_distinct(flightID), 1),
    `first DOY` = min(doy),
    `last DOY`  = max(doy),
    moths       = sum(moths, na.rm = TRUE),
    .groups     = "drop"
  ) |>
  mutate(year = as.character(year)) |>   # a year is a label, not a count
  kable(format.args = list(big.mark = ","))
```

The network is a rotating panel rather than a fixed set of stations: most
localities contribute a handful of years, and the annual number of traps
varies with the outbreak and with the Maine Forest Service's programme.

```{r}
#| label: tbl-site-years
#| tbl-cap: "Localities by the number of years they contribute"
trap_records |>
  group_by(locality) |>
  summarise(years = n_distinct(year), .groups = "drop") |>
  count(years, name = "localities") |>
  arrange(years) |>
  kable()
```

### Data quality

The following checks are run by `R/01_read_trap_data.R` at every read. They
report problems; they never repair them.

```{r}
#| label: tbl-quality
#| tbl-cap: "Quality checks on the retained records"
gaps <- filter(trap_records, n_gap_nights > 0)
miss <- trap_daily |> filter(status == "missing") |> distinct(flightID)
tibble::tibble(
  Check = c("Season total disagrees with the sum of the daily counts",
            "Locality-year with more than one record",
            "Records with nights coded `m`",
            "Records with interior gaps (`-` inside the season)"),
  Records = c(0L, 0L, nrow(miss), nrow(gaps))
) |>
  kable(align = c("l", "r"))
```

```{r}
#| label: tbl-gaps
#| tbl-cap: "Records with unrecorded nights inside their trapping season"
trap_records |>
  filter(n_gap_nights > 0) |>
  transmute(flightID, year = as.character(year), locality,
            season = sprintf("%d\u2013%d", first_doy, last_doy),
            `nights counted` = n_nights, `gap nights` = n_gap_nights,
            `season total` = season_total) |>
  arrange(desc(`gap nights`)) |>
  kable(format.args = list(big.mark = ","))
```

Georgetown Township 1969 is the one record where the gaps dominate: 23
unrecorded nights against 21 counted, over a nominal season of 44 days. The
1980 pooled records were operated on an irregular schedule rather than
interrupted. All other retained records are continuous between their first and
last night.

## Minimum effort for phenology

A season total above 99 moths says nothing about whether the trap was running
long enough, or at the right time, to describe a flight curve. The rule below
matches the requirement to the statistic: for a percentile to be identifiable,
the sampling window has to bracket it. If more than 5% of a season's catch
falls on the first night, the 5th percentile sits at or before that night and
cannot be located; the same argument applies to p95 and the last night.

| Criterion | Threshold | What it protects against |
|-----------|-----------|--------------------------|
| `min_nights` | ≥ 15 counted nights | percentiles interpolated from too few points |
| `min_coverage` | counted nights / season span ≥ 0.7 | holes inside the season displacing the cumulative curve |
| `max_tail_share` | ≤ 0.05 of the catch on the first *and* on the last night | a flight period truncated at either end |

: Effort criteria, set as constants in `R/01_read_trap_data.R` {#tbl-effort-rule}

The criteria are recorded, not enforced: `trap_records` carries `coverage`,
`tail_first`, `tail_last` and the logical `usable_phenology`, so an analysis
that needs only the median can relax them.

```{r}
#| label: tbl-usable
#| tbl-cap: "Records excluded from percentile-based phenology, by reason"
trap_records |>
  mutate(reason = case_when(
    usable_phenology                       ~ "usable",
    n_nights < 15                          ~ "too few nights (< 15)",
    coverage < 0.7                         ~ "too gappy (coverage < 0.7)",
    TRUE                                   ~ "flight period truncated (tail > 5%)"
  )) |>
  count(reason, name = "records") |>
  arrange(desc(records)) |>
  kable(align = c("l", "r"))
```

Effort itself is rarely the problem: coverage is exactly 1 for most records
and the median series runs 25 nights. Truncation is what binds, and it is not
a marginal call in the worst cases — East Allagash 1986 took 77% of its season
on the final night, Round Mountain 1972 71% on the first.

```{r}
#| label: tbl-usable-year
#| tbl-cap: "Usable records per year"
trap_records |>
  group_by(year) |>
  summarise(records = n(), usable = sum(usable_phenology),
            localities = n_distinct(locality[usable_phenology]),
            nights = sum(n_nights[usable_phenology]), .groups = "drop") |>
  mutate(year = as.character(year)) |>
  kable(format.args = list(big.mark = ","))
```

The cost is concentrated in the sparse years. Sensitivity of the tail
threshold, holding the other two criteria fixed:

```{r}
#| label: tbl-tail-sensitivity
#| tbl-cap: "Effect of the tail threshold"
bind_rows(lapply(c(0.05, 0.10, 0.15), function(tt) {
  k <- with(trap_records, n_nights >= 15 & coverage >= 0.7 &
              tail_first <= tt & tail_last <= tt)
  tibble::tibble(`tail threshold` = tt,
                 records    = sum(k),
                 localities = n_distinct(trap_records$locality[k]),
                 years      = n_distinct(trap_records$year[k]),
                 `years lost` = paste(setdiff(unique(trap_records$year),
                                              unique(trap_records$year[k])),
                                      collapse = ", "))
})) |>
  kable(align = c("r", "r", "r", "r", "l"))
```

1987 and 1988 have a single trap each, so one failing record erases the year.
Whether that is acceptable depends on the analysis: a pooled site-year model
loses little, a year-by-year trend loses the two terminal years of the
outbreak collapse.

## Open questions

- **1987 and 1988 have no usable record.** Under the current tail threshold
  both years disappear from the phenology set. Relaxing the threshold to 0.10
  restores 1987 at the cost of admitting series whose flight period is
  probably truncated; the decision belongs with the analysis, not the reader.
- **Weather resolution.** The CaSR reanalysis covers every retained locality
  (all within 6.3 km of a grid cell) and every retained year, but air
  temperature is stored 3-hourly, not hourly, unlike the other four variables.
- **Records excluded by the year window.** 1961–1967 and 2016 are outside the
  reanalysis, and are set aside rather than lost: widening `keep_years` and
  setting `min_season_total` to 0 in `R/01_read_trap_data.R` restores the full
  643-record Maine light-trap set.

# Weather data {#sec-weather}

```{r}
#| label: weather-data
weather_hourly <- readRDS(file.path("data", "processed", "weather_hourly.rds"))

## One summary pass over the nested hourly tables; every weather figure below
## is taken from it.
wsum <- weather_hourly |>
  rowwise() |>
  summarise(
    Location, Year, Latitude, tile,
    hours       = nrow(Weather),
    na_temp     = sum(is.na(Weather$Temp)),
    tt_measured = sum(!is.na(Weather$TT)),
    max_dev     = max(abs(Weather$Temp - Weather$TT), na.rm = TRUE),
    tmin        = min(Weather$Temp),
    tmax        = max(Weather$Temp),
    july_mean   = mean(Weather$Temp[month(Weather$datetime) == 7]),
    season_mean = mean(Weather$Temp[yday(Weather$datetime) %in% 160:230]),
    rain_mm     = sum(Weather$PR0) * 1000,
    wind_mean   = mean(Weather$UVC),
    .groups     = "drop"
  )
```

## Source

Weather comes from the ECCC Canadian Surface Reanalysis (**CaSR v3.2**), a
~10 km rotated-pole reanalysis. The two tiles that cover Maine were downloaded
from the [CaSR download
page](https://hpfx.collab.science.gc.ca/~scar700/rcas-casr/download_CaSR_regions_var_period_fr.html)
into `data/raw/Weather_Maine/` — 60 NetCDF files, 4.1 GB, five variables in
six four-year periods spanning 1968–1991.

| Variable | Kind | Meaning | Units | Stored |
|----------|------|---------|-------|--------|
| `TT` | analysis | air temperature at 1.5 m | °C | every 3 h |
| `PR0` | analysis | quantity of precipitation | m | hourly |
| `FB` | forecast | downward solar flux at surface | W m⁻² | hourly |
| `P0` | forecast | surface pressure | mb | hourly |
| `UVC` | forecast | wind modulus at 10 m | kts | hourly |

: CaSR variables extracted {#tbl-casr-vars}

The reanalysis window is what fixes `keep_years` in the trap selection: 1968 is
the first year with weather, and no trap in 1990 or 1991 was sampled.

## Extraction

`R/02_read_weather_data.R` takes the locality-years from
`data/processed/trap_records.rds`, projects the trap coordinates onto the
rotated-pole grid, assigns each locality to the tile that contains it, and
extracts the grid cell holding the trap for every hour of the calendar year.

```{r}
#| label: tbl-weather-grid
#| tbl-cap: "Trap localities on the CaSR grid"
tibble::tibble(
  Quantity = c("Locality-years extracted", "Localities", "Years",
               "Hourly rows", "Southern tile / northern tile",
               "Distance to the grid cell used (km)"),
  Value = c(
    format(nrow(wsum), big.mark = ","),
    n_distinct(wsum$Location),
    sprintf("%d (%d\u2013%d)", n_distinct(wsum$Year), min(wsum$Year), max(wsum$Year)),
    format(sum(wsum$hours), big.mark = ","),
    ## tiles are named by rlat band: 351-385 is the southern one, 386-420 the northern
    paste(as.integer(table(wsum$tile[!duplicated(wsum$Location)])), collapse = " / "),
    "0.7 to 6.3, median 3.9"
  )
) |>
  kable(align = c("l", "r"))
```

Two details of the file layout needed care.

*The period files are not aligned on calendar years.* Each spans 31 December
13:00 UTC of the year before its period to 31 December 12:00 UTC of its last
year, so the final 11 hours of 1971, 1975, 1979, 1983 and 1987 sit in the
*next* period file. Selecting layers by year within a single file silently
returns 8,749-hour years; the script reads both files for those years.

*The tiles are read separately.* Each locality is extracted from the one tile
that covers it, rather than merging two 35,064-layer rasters per variable and
period.

## The resulting dataset

`data/processed/weather_hourly.rds` (51 MB) holds one row per locality-year
with a nested `Weather` tibble: `datetime` (UTC), `datetime_est` (local
standard time, UTC−5 all year), the interpolated `Temp`, the raw `TT`, and
`PR0`, `FB`, `P0`, `UVC`.

```{r}
#| label: tbl-weather-complete
#| tbl-cap: "Completeness of the extracted series"
tibble::tibble(
  Check = c("Locality-years with a complete calendar year of hours",
            "Hours with no interpolated temperature",
            "Hours with a measured (3-hourly) temperature",
            "Hours missing any of PR0, FB, P0, UVC"),
  Value = c(
    sprintf("%d of %d", sum(wsum$hours %in% c(8760L, 8784L)), nrow(wsum)),
    format(sum(wsum$na_temp), big.mark = ","),
    sprintf("%s (%.0f%% of all hours)", format(sum(wsum$tt_measured), big.mark = ","),
            100 * sum(wsum$tt_measured) / sum(wsum$hours)),
    "0"
  )
) |>
  kable(align = c("l", "r"))
```

### Interpolation of air temperature

Temperature is stored every three hours, so two hours in three are filled with
`chillR::interpolate_gaps_hourly()`, which fits an idealised daily temperature
curve to the daily minima and maxima — hence the latitude argument — and
interpolates the residuals. The measured values are left untouched, which is
worth verifying rather than assuming:

```{r}
#| label: tbl-weather-interp
#| tbl-cap: "Interpolation audit"
tibble::tibble(
  Quantity = c("Largest |interpolated - measured| at a measured hour (°C)",
               "Share of hours interpolated",
               "Temperature range over all locality-years (°C)"),
  Value = c(
    format(max(wsum$max_dev), digits = 3, scientific = TRUE),
    sprintf("%.0f%%", 100 * (1 - sum(wsum$tt_measured) / sum(wsum$hours))),
    sprintf("%.1f to %.1f", min(wsum$tmin), max(wsum$tmax))
  )
) |>
  kable(align = c("l", "r"))
```

The deviation at measured hours is floating-point noise, so `Temp` reproduces
`TT` exactly where `TT` exists and only fills the gaps between.

### Does the weather look like Maine?

```{r}
#| label: tbl-weather-climate
#| tbl-cap: "Climate of the trap localities, as extracted"
tibble::tibble(
  Quantity = c("Mean July temperature (°C)",
               "Mean temperature over the flight window, DOY 160-230 (°C)",
               "Correlation of mean July temperature with latitude",
               "Annual precipitation (mm)",
               "Mean wind speed at 10 m (kts)"),
  Value = c(
    sprintf("%.1f to %.1f", min(wsum$july_mean), max(wsum$july_mean)),
    sprintf("%.1f to %.1f", min(wsum$season_mean), max(wsum$season_mean)),
    sprintf("%.2f", cor(wsum$july_mean, wsum$Latitude)),
    sprintf("%d to %d, median %d", round(min(wsum$rain_mm)), round(max(wsum$rain_mm)),
            round(median(wsum$rain_mm))),
    sprintf("%.1f to %.1f", min(wsum$wind_mean), max(wsum$wind_mean))
  )
) |>
  kable(align = c("l", "r"))
```

Summer temperature falls towards the north, but latitude alone orders the
localities poorly. The coolest Julys belong to coastal and island sites —
Bare Island 17.2 °C at 44.6° N, Georgetown 17.4 °C at 43.8° N, Steuben
17.6 °C at 44.5° N — all cooler than Estcourt, 21.0 °C on the Québec border at
47.4° N. Any temperature-driven model of flight timing therefore has to carry
the maritime gradient, not just latitude.

## Open questions for the weather

- **Temperature is interpolated, not measured, for two hours in three.** Any
  night-level analysis that depends on a specific hour — the temperature at
  dusk, or hours above a flight threshold — inherits the idealised diurnal
  curve, not the reanalysis. Aggregates over a whole night are much less
  exposed than instantaneous values.
- **Local time is a constant offset.** `datetime_est` is UTC−5 all year, so it
  is one hour behind civil time in summer. A series nested by UTC year starts
  at 19:00 on 31 December of the preceding year in local time.
- **Reanalysis, not stations.** CaSR is a model product on a 10 km grid: the
  cell holding a trap is 0.7–6.3 km from it and represents ~100 km², so the
  extracted series describes the neighbourhood, not the trap clearing.

# Modelled phenology from BioSIM {#sec-biosim}

```{r}
#| label: biosim-data
biosim_phenology <- readRDS(file.path("data", "processed", "biosim_phenology.rds"))
biosim_flight    <- readRDS(file.path("data", "processed", "biosim_flight.rds"))
```

Alongside the observed catches, spruce budworm seasonal biology was simulated
at every trap locality-year with **BioSIM**, so that observed and modelled
flight phenology can be compared on the same dates and the same statistic.

## Model and weather

`R/03_biosim_phenology.R` calls the BioSIM Web API through the `BioSIM` R
package, which drives a Java client. For each year one request covers the
localities sampled that year: BioSIM assembles the daily weather for those
coordinates from its own historical station network and runs
`Spruce_Budworm_Biology` on it — Régnière's process-oriented model of spruce
budworm seasonal biology (Régnière 1982, 1983, 1984, 1987, 1990; Régnière &
You 1991), driven by daily minimum, mean and maximum temperature.

This is a second, independent weather source: the CaSR reanalysis of the
previous section is not what BioSIM uses. Every simulated day in this run came
back marked as resting on observed rather than generated weather.

## Parameters

The model takes eight parameters. Seven are left at the server defaults; the
two mortality switches are turned off.

| Parameter | Default | Used here | Meaning |
|-----------|---------|-----------|---------|
| `ApplyMortality` | 1 | **0** | apply natural mortality over the population |
| `ApplyAdultMortality` | 1 | **0** | (undocumented in the model help) |
| `CreateEggs` | 0 | 0 | allow more than one generation |
| `TreeKind` | 0 | 0 | host tree on which the insect develops |
| `LastSampleDate` | *(empty)* | *(unset)* | date of the last field observation |
| `ObservedAI` | 0 | 0 | last observed average instar |
| `Defoliation` | 0 | 0 | (undocumented) |
| `AdultLongevity` | 0 | 0 | (undocumented) |

: BioSIM parameters for `Spruce_Budworm_Biology` {#tbl-biosim-parms}

Mortality is off because the question here is *when* the stages occur, not how
many survive. With mortality on, only a few percent of the cohort reaches the
adult stage, so the flight curve would be estimated from a handful of
individuals: checked at one Maine locality-year, the season totals are
`MaleEmergence + FemaleEmergence` = 100% and `MaleFlight + FemaleFlight` = 97%
with mortality off, against 10% and 6% under the defaults.

`LastSampleDate` and `ObservedAI` are left unset, so the simulations are driven
by weather alone and are not anchored on any field observation of development.

## What the model returns, and what is kept

The output is daily and everything is a **percentage of the simulated cohort**,
quantised in steps of 0.25%. Besides the eight stages (`L2o` … `Adults`) it
reports emergence, flight, oviposition and death, separately for each sex.
Three quantities are kept, each the sum of the two sexes:

| Series | Definition | Kind |
|--------|------------|------|
| `Pupae_mf` | `MalePupae` + `FemalePupae` | stock: percentage present that day |
| `Adult_mf` | `MaleAdult` + `FemaleAdult` | stock: percentage present that day |
| `Flight_mf` | `MaleFlight` + `FemaleFlight` | activity: flight that day |

: Series retained from the model output {#tbl-biosim-series}

Only flight is a daily flux, so only its cumulative curve is a genuine
distribution over the season — and it is the series to compare with trap
catches. The two stocks are weighted by how long individuals remain in the
stage and are kept for context. The sex-specific columns reproduce the
both-sexes columns the model also reports exactly, which is checked at every
run.

The simulation is repeated ten times per locality-year and the runs are
averaged, which smooths the 0.25% quantisation of a single run; percentiles are
unaffected by the choice of averaging over summing.

## The resulting dataset

```{r}
#| label: tbl-biosim-overview
#| tbl-cap: "The BioSIM run at a glance"
tibble::tibble(
  Quantity = c("Locality-years simulated", "Localities", "Years",
               "Runs per locality-year", "Rows of daily output",
               "Days resting on observed weather"),
  Value = c(
    format(nrow(biosim_flight), big.mark = ","),
    n_distinct(biosim_flight$Location),
    sprintf("%d (%d\u2013%d)", n_distinct(biosim_flight$Year),
            min(biosim_flight$Year), max(biosim_flight$Year)),
    sprintf("%d", max(biosim_flight$n_runs)),
    "829,060 raw / 82,906 averaged",
    "all"
  )
) |>
  kable(align = c("l", "r"))
```

```{r}
#| label: tbl-biosim-series-summary
#| tbl-cap: "Modelled phenology, mean over the locality-years"
biosim_phenology |>
  mutate(series = factor(series, levels = c("Pupae_mf", "Adult_mf", "Flight_mf"))) |>
  group_by(series) |>
  ## The range must be taken before `p50` is replaced by its mean: summarise()
  ## evaluates its arguments in order, in the same data mask.
  summarise(`p50 range` = sprintf("%.0f\u2013%.0f", min(p50), max(p50)),
            `p05` = mean(p05), `p25` = mean(p25), `p50` = mean(p50),
            `p75` = mean(p75), `p95` = mean(p95),
            `duration (d)` = mean(duration),
            .groups = "drop") |>
  relocate(`p50 range`, .after = `duration (d)`) |>
  mutate(across(where(is.numeric), ~ round(.x, 1))) |>
  kable(align = c("l", rep("r", 7)))
```

Modelled flight is centred on day of year 190 on average — 9 July — and the
central 90% of it spans a fortnight. Pupation runs about ten days ahead of
flight, and the adult stock peaks slightly after flight because adults persist
once mortality is switched off.

The locality-year medians are spread over six weeks (DOY 169 to 210) and all
227 values differ, so the simulations respond to site and year rather than
returning a regional average: 1974 alone spans DOY 182 to 210 across its 21
localities.

## Cohorts that do not finish

With mortality off, a locality-year should end with essentially the whole
cohort having flown. Not all of them do: part of the cohort fails to complete
development and is dropped, which shows up as a flight season total well below
100%. `flight_complete` marks the locality-years that reach at least 90%.

```{r}
#| label: tbl-biosim-complete
#| tbl-cap: "Completeness of the simulated cohorts"
biosim_flight |>
  mutate(band = case_when(season_pct >= 90 ~ "90% or more (flight_complete)",
                          season_pct >= 50 ~ "50 to 90%",
                          TRUE             ~ "below 50%")) |>
  group_by(band) |>
  summarise(`locality-years` = n(),
            `mean p50` = round(mean(p50), 1),
            `season total, %` = sprintf("%.1f\u2013%.1f", min(season_pct), max(season_pct)),
            .groups = "drop") |>
  arrange(desc(`locality-years`)) |>
  kable(align = c("l", "r", "r", "r"))
```

```{r}
#| label: tbl-biosim-incomplete
#| tbl-cap: "Locality-years whose cohort did not complete flight"
biosim_flight |>
  filter(!flight_complete) |>
  arrange(season_pct) |>
  transmute(Location, Year = as.character(Year),
            `season total, %` = round(season_pct, 1),
            p05 = round(p05, 1), p50 = round(p50, 1), p95 = round(p95, 1)) |>
  kable(align = c("l", "l", "r", "r", "r", "r"))
```

Two features of that list matter. It is concentrated at particular sites rather
than scattered — Clayton Lake is flagged in five separate years — which points
to a site effect rather than to bad years. And the flagged locality-years are
*later*, not earlier, than the complete ones, which is the opposite of what one
would expect if the surviving fraction were the fast developers; it is
consistent instead with incompleteness arising in cold site-years where
development runs late. The flag is carried on all three series and nothing is
dropped: what to do with these locality-years is a decision for the analysis.

## Open questions for the modelled phenology

- **The comparison with the catches is still to be done.** On the three
  locality-years used to test the script, modelled flight ran one to two weeks
  ahead of the observed catch window; whether that holds over the whole run is
  the next question, and the two tables join on locality-year.
- **Three parameters are undocumented.** `ApplyAdultMortality`, `Defoliation`
  and `AdultLongevity` have no description in the model help and no stated
  units. `AdultLongevity = 0` in particular may be what makes the adult stock
  persist to the end of the year.
- **Nothing anchors the simulation to observed development.** The model offers
  `LastSampleDate` and `ObservedAI` for that, and the `Phenology` sheet of the
  source workbook holds average-instar observations — none of them in Maine,
  but the option exists for a validation elsewhere.


# Modelled pupation from bayessbw

```{r}
#| label: pupal-data
pupal_phenology <- readRDS(file.path("data", "processed", "pupal_phenology.rds"))
```

A second, independent model of development was run over the same
locality-years: the Bayesian development model of Studens et al.
([kdis19/bayessbw](https://github.com/kdis19/bayessbw)), driven by the CaSR
hourly temperatures of @sec-weather rather than by BioSIM's own weather.

## What the model is, and what "pupal phenology" means here

`bayessbw` fits Sharpe-DeMichele / Schoolfield development-rate curves to
laboratory rearing data, in a Bayesian hierarchical form: every posterior draw
carries a rate curve for each larval stage plus a lognormal individual-variation
term. The fitted posterior is distributed as the R package
[`sbwFieldPheno`](https://github.com/studensk/sbwFieldPheno), whose `dev_days()`
accumulates development hour by hour through a temperature series and returns
the day of year on which each simulated individual reaches a given stage.

The laboratory data stop at L6, so the model describes the five larval stages
**L2 to L6 only**. Pupation is the end of L6 — the last transition the model
knows about — and the pupal stage itself is not modelled. *Pupal phenology*
here therefore means the **date of pupation**, not the duration of the pupal
stage. This is also how `bayessbw`'s own weather simulations use the package.

`R/04_pupal_phenology.R` runs it with the settings below.

| Setting | Value | Why |
|---------|-------|-----|
| colony | `NB` | six are available (`AB`, `IPQL`, `NB`, `NWT`, `ON`, `QC`); New Brunswick is the nearest field colony to Maine |
| window | 1 March – 31 August | the window used in `bayessbw`'s own simulations |
| posterior draws | 100 | every draw the package holds |
| individuals per draw | 100 | the simulated population |

: Settings for the pupation simulation {#tbl-pupal-settings}

Two things about that table matter more than the rest. `dev_days()` starts
accumulating development on the **first row of weather it is given**, so 1 March
is an assumption about the date of post-diapause L2 emergence, and every
predicted date moves with it. And because all 100 posterior draws are used, the
posterior spread reported below is the model's own uncertainty, not a sampling
error.

## The resulting dataset

```{r}
#| label: tbl-pupal-overview
#| tbl-cap: "Predicted pupation across the 227 locality-years"
tibble::tibble(
  Quantity = c("Locality-years", "Population median (p50), mean",
               "Population median, range",
               "Within-population spread (p05\u2013p95), mean",
               "Posterior interval on the median, mean width"),
  Value = c(
    format(nrow(pupal_phenology), big.mark = ","),
    sprintf("DOY %.1f", mean(pupal_phenology$p50)),
    sprintf("DOY %.0f\u2013%.0f", min(pupal_phenology$p50), max(pupal_phenology$p50)),
    sprintf("%.1f d", mean(pupal_phenology$spread)),
    sprintf("%.1f d", mean(pupal_phenology$post_hi - pupal_phenology$post_lo))
  )
) |>
  kable(align = c("l", "r"))
```

Each simulation returns a matrix of individuals by posterior draws. Sorting a
draw's column turns it into that draw's population distribution, and taking the
median rank by rank across draws gives the **median simulated population** —
the summary `bayessbw` itself uses, written out in full as
`pupal_population.rds`. Percentiles in the table above are percentiles of that
population; the posterior interval is built the other way round, from the
median of each draw, so the two columns separate individual variation from
parameter uncertainty.

They differ by a factor of four: the fitted curves place the population median
to within about two days, while the population itself takes nine days to pupate
from p05 to p95. Individual variation, not parameter uncertainty, dominates.

```{r}
#| label: tbl-pupal-vs-biosim
#| tbl-cap: "Predicted pupation, bayessbw against BioSIM"
biosim_pupae <- biosim_phenology |>
  filter(series == "Pupae_mf") |>
  select(Location, Year, biosim_p50 = p50)

pupal_phenology |>
  select(Location, Year, bayes_p50 = p50) |>
  inner_join(biosim_pupae, by = c("Location", "Year")) |>
  summarise(
    `locality-years` = n(),
    `bayessbw p50, mean` = round(mean(bayes_p50), 1),
    `BioSIM p50, mean`   = round(mean(biosim_p50), 1),
    `difference, median` = round(median(bayes_p50 - biosim_p50), 1),
    `difference, range`  = sprintf("%+.1f to %+.1f", min(bayes_p50 - biosim_p50),
                                   max(bayes_p50 - biosim_p50)),
    `correlation`        = round(cor(bayes_p50, biosim_p50), 2)
  ) |>
  kable(align = c("r", "r", "r", "r", "r", "r"))
```

The two models were built from different data, fitted by different methods and
driven by different weather, and they place pupation within a day of each
other on average. Their disagreement at individual locality-years reaches two
weeks in both directions, which is the more useful number: it is the scale at
which either model's site-level prediction should be trusted.

## Open questions for the pupation model

- **The 1 March start is an assumption, not a measurement.** Anchoring the
  simulation on BioSIM's own L2 emergence date instead would make the two
  models comparable on equal terms and remove one arbitrary choice.
- **One colony of six.** `NB` is used throughout. The other five are one
  configuration change away, and comparing them would show how much the source
  population matters against site and year.
- **The pupal stage is not modelled.** Getting from pupation to flight needs
  either BioSIM or a separate pupal development function; the section below
  measures that interval empirically instead.


# Model versus observed {#sec-compare}

```{r}
#| label: compare-data
model_vs_observed <- readRDS(file.path("data", "processed", "model_vs_observed.rds"))
cmp <- filter(model_vs_observed, comparable)
```

`R/05_compare_model_observed.R` brings the three sides together, one row per
locality-year: the percentiles of the nightly catch curve, the percentiles of
BioSIM's daily flight curve, and the predicted date of pupation from both
models. All of them are percentiles of a curve over the season computed by the
same helper (`R/functions/phenology.R`), so `p50` always means *half of the
season's total has happened by this day*.

## What can be compared

A caught moth is not the same thing as a flying moth: the trap samples adults
that are active, close enough and attracted to light, whereas BioSIM reports the
percentage of the cohort in the flight state. The comparison below is about
**timing**, never about numbers.

Two flags decide which rows carry a meaningful comparison, and both are kept in
the output rather than filtered away.

```{r}
#| label: tbl-comparable
#| tbl-cap: "Locality-years available for comparison"
model_vs_observed |>
  mutate(group = case_when(
    comparable ~ "Comparable: usable season and complete cohort",
    usable_phenology & !flight_complete ~ "Usable season, incomplete modelled cohort",
    !usable_phenology & flight_complete ~ "Truncated trapping season, complete cohort",
    TRUE ~ "Neither"
  )) |>
  group_by(group) |>
  summarise(`locality-years` = n(),
            `median offset, d` = round(median(d_p50), 1),
            .groups = "drop") |>
  arrange(desc(`locality-years`)) |>
  kable(align = c("l", "r", "r"))
```

Only the first group is used for the statistics that follow. The truncated
seasons are shown in the table because their offset is the diagnostic one would
expect: a season that starts too late or ends too early pushes the observed
median towards the middle of the sampled window, and those rows sit further
from the model than the comparable ones.

## Flight

```{r}
#| label: tbl-offsets
#| tbl-cap: "BioSIM flight minus observed catch, over the comparable locality-years"
tibble::tibble(
  Statistic = c("p05 (start)", "p25", "p50 (median)", "p75", "p95 (end)",
                "peak day", "duration p05\u2013p95"),
  `Median offset, d` = c(median(cmp$d_p05),
                         median(cmp$sim_p25 - cmp$obs_p25),
                         median(cmp$d_p50),
                         median(cmp$sim_p75 - cmp$obs_p75),
                         median(cmp$d_p95),
                         median(cmp$d_peak),
                         median(cmp$d_duration))
) |>
  mutate(`Median offset, d` = sprintf("%+.1f", `Median offset, d`)) |>
  kable(align = c("l", "r"))
```

The model is **early, but only just**: half a day to four days, depending on
which part of the curve is compared. That is a much smaller bias than the one
to two weeks suggested by the three locality-years used to test the simulation
script, and it settles the open question left at the end of @sec-biosim.

The offset is largest at the start of the season and smallest at the end, which
is another way of saying that the modelled flight period is longer than the
observed one — 14.7 days against 11.0 between p05 and p95. Part of that is
real and part is an artefact of trapping: a trap that catches nothing on a night
when a few moths fly contributes nothing to the tails of the observed curve.

The average is not the whole story, though.

```{r}
#| label: fig-compare
#| fig-cap: "Two development models against the same observed catches, one point per comparable locality-year (n = 170). BioSIM is its modelled median flight date; bayessbw is its median pupation date plus the 12.9 d median pupation-to-catch interval. That shift is calibrated on these same data, so the bayessbw series sits on the 1:1 line by construction and only its scatter is informative. Left: predicted against observed median date. Right: the same offsets against latitude, with an ordinary least-squares fit per model."
#| fig-width: 9
#| fig-height: 4.4
library(ggplot2)
library(patchwork)

model_cols <- c(BioSIM = "#1f6f8b", bayessbw = "#d1670f")

## One row per locality-year and model.
cmp_long <- bind_rows(
  transmute(cmp, Location, Year, Latitude, obs = obs_p50,
            pred = sim_p50,        offset = d_p50,       model = "BioSIM"),
  transmute(cmp, Location, Year, Latitude, obs = obs_p50,
            pred = bayes_pred_p50, offset = d_p50_bayes, model = "bayessbw")
) |>
  mutate(model = factor(model, levels = names(model_cols)))

base <- theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(size = 10, face = "plain", hjust = 0),
        legend.title = element_blank(),
        legend.key.size = unit(0.9, "lines"))

lims <- range(c(cmp_long$obs, cmp_long$pred))

p_scatter <- ggplot(cmp_long, aes(obs, pred, colour = model)) +
  geom_abline(slope = 1, intercept = 0, linetype = "22", colour = "grey45") +
  geom_point(size = 1.7, alpha = 0.7) +
  annotate("text", x = lims[2], y = lims[2], hjust = 1, vjust = 1.8,
           size = 3, colour = "grey45", label = "1:1") +
  scale_colour_manual(values = model_cols) +
  coord_equal(xlim = lims, ylim = lims) +
  labs(title = "Site by site both models agree loosely",
       x = "Observed median catch date (DOY)",
       y = "Predicted date (DOY)") +
  theme(legend.position = "inside",
        legend.position.inside = c(0.02, 0.98),
        legend.justification = c(0, 1),
        legend.background = element_blank())

p_lat <- ggplot(cmp_long, aes(Latitude, offset, colour = model, fill = model)) +
  geom_hline(yintercept = 0, colour = "grey45", linetype = "22") +
  geom_point(size = 1.7, alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
              linewidth = 0.7, alpha = 0.18) +
  scale_colour_manual(values = model_cols) +
  scale_fill_manual(values = model_cols) +
  labs(title = "and both run early in the south",
       x = "Latitude (\u00b0N)",
       y = "Predicted minus observed (d)") +
  theme(legend.position = "none")

(p_scatter & base + theme(legend.position = "inside",
                          legend.position.inside = c(0.02, 0.98),
                          legend.justification = c(0, 1))) |
  (p_lat & base + theme(legend.position = "none"))
```

```{r}
#| label: tbl-compare-structure
#| tbl-cap: "Where the agreement is, and where it is not. Correlations over the 170 comparable locality-years"
## The same four correlations for either model, so the columns can be read
## against each other.
structure_r <- function(pred, offset) {
  d <- cmp |> mutate(pred = {{ pred }}, offset = {{ offset }})
  c(cor(d$pred, d$obs_p50),
    with(d |> group_by(Year) |> filter(n() >= 5) |>
           summarise(p = mean(pred), o = mean(obs_p50), .groups = "drop"),
         cor(p, o)),
    cor(residuals(lm(pred ~ factor(Year), d)),
        residuals(lm(obs_p50 ~ factor(Year), d))),
    cor(d$offset, d$Latitude),
    cor(d$offset, log10(d$season_total)))
}

tibble::tibble(
  Comparison = c("Predicted against observed, all locality-years",
                 "Year means (years with 5 or more localities)",
                 "Within years, across localities",
                 "Offset against latitude",
                 "Offset against log10 season total"),
  BioSIM   = structure_r(sim_p50, d_p50),
  bayessbw = structure_r(bayes_pred_p50, d_p50_bayes)
) |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  kable(align = c("l", "r", "r"))
```

This is the substantive result of the comparison, and it has three parts.

**Both models get the year right.** Averaged over the localities sampled in a
year, predicted and observed median dates correlate at 0.83 for BioSIM and
0.71 for bayessbw: warm early seasons and cold late ones are reproduced, and
the six-week span of median flight dates across the record is not an artefact
of which sites happened to be trapped.

**Neither gets the site right.** Once the year is removed, the correlation
across localities within a year collapses to 0.14 for BioSIM and 0.06 for
bayessbw. Whatever makes one Maine locality fly earlier than another in the
same summer, neither model captures it.

**The residual is not noise, and it is the same residual in both models.**
It has a clear spatial structure — the offset correlates with latitude at 0.49
(BioSIM) and 0.39 (bayessbw), so both run early in the south, and more weakly
with the size of the catch (0.28 and 0.17; the two are disentangled below) —
and the two models' offsets correlate with each other at 0.74, locality-year
by locality-year. That last number is the informative one. The models were fitted
to different data by different methods and, crucially, were driven by
different weather: BioSIM interpolates its own station network, bayessbw ran
on the CaSR grid extracted in @sec-weather. A spatial error common to both is
therefore unlikely to originate in either weather source. It points instead at
something both models share — temperature-driven development with no
site-level modifier — or at something neither of them contains, of which
immigration is the obvious candidate.

```{r}
#| label: tbl-compare-sites
#| tbl-cap: "Localities with at least three comparable years and a median offset beyond a week in either model"
cmp |>
  group_by(Location) |>
  filter(n() >= 3) |>
  summarise(years = n(),
            `latitude` = round(first(Latitude), 2),
            `median catch` = median(season_total),
            `BioSIM, d`   = round(median(d_p50), 1),
            `bayessbw, d` = round(median(d_p50_bayes), 1),
            .groups = "drop") |>
  filter(abs(`BioSIM, d`) > 7 | abs(`bayessbw, d`) > 7) |>
  arrange(`BioSIM, d`) |>
  kable(align = c("l", "r", "r", "r", "r", "r"))
```

The worst-fitting sites are southern and coastal — the same localities that the
weather section found to have the coolest Julys in the dataset despite their
latitude. A model that is too early there is a model whose development is
running too fast on those temperature series. They are also the localities
that catch the fewest moths, which is a second candidate explanation and is
separated from latitude below.

## From pupation to flight

The pupal stage is missing from `bayessbw` and the trap sees only adults, so the
interval between them can be measured rather than modelled.

```{r}
#| label: tbl-lag
#| tbl-cap: "Days from predicted pupation to the observed median catch"
tibble::tibble(
  Source = c("bayessbw pupation", "BioSIM pupation"),
  `Median, d` = c(median(cmp$lag_bayes), median(cmp$lag_biosim)),
  `Mean, d`   = c(mean(cmp$lag_bayes), mean(cmp$lag_biosim)),
  `SD, d`     = c(sd(cmp$lag_bayes), sd(cmp$lag_biosim))
) |>
  mutate(across(where(is.numeric), ~ round(.x, 1))) |>
  kable(align = c("l", "r", "r", "r"))
```

Both models put the median catch about a fortnight after pupation, which is
a plausible pupal period plus a few days of adult maturation, and the agreement
between two independent development models on that interval is reassuring. The
standard deviation of six days is the part worth pursuing: if it is structured
by temperature rather than random, it is a pupal development function waiting
to be fitted.

## Would a degree-day pupal stage help? {#sec-lag}

```{r}
#| label: lag-data
lag_test <- readRDS(file.path("data", "processed", "pupal_lag_test.rds"))
lag_rows <- lag_test$rows
lag_sum  <- lag_test$summary
```

A constant is a poor model of a poikilotherm's development, so the obvious
improvement on the fixed 12.9 d interval is to accumulate degree-days from
pupation and predict the catch median when a threshold is reached.
`R/06_pupal_lag_test.R` fits that and compares it with the constant.

```{r}
#| label: tbl-lag-models
#| tbl-cap: "Predicting the observed median catch from the predicted pupation date"
tibble::tibble(
  Model = c(sprintf("A  fixed lag, %.1f d", lag_sum$lag_A),
            sprintf("B  degree-days, base %.1f \u00b0C, K = %g",
                    lag_sum$fit_B$base, lag_sum$fit_B$K)),
  Parameters = c(1, 2),
  `RMSE in sample, d` = round(c(lag_sum$rmse_A, lag_sum$rmse_B), 2),
  `RMSE leave-one-year-out, d` = round(c(lag_sum$cv_rmse_A, lag_sum$cv_rmse_B), 2)
) |>
  kable(align = c("l", "r", "r", "r"))
```

The degree-day model wins by 0.08 days out of sample, on 170 locality-years.
The fitted base ran to the bottom of the grid, which is the diagnostic rather
than a detail: at a base of 0 °C the accumulation is nearly linear in mean
temperature, which is as close as the model can come to collapsing back into
the constant. Over the whole dataset it stretches the interval from 11 to 16
days, against a residual scatter of six.

## Why no development function would help

A development model can only use temperature, so two numbers bound what it
could ever achieve.

```{r}
#| label: tbl-lag-ceiling
#| tbl-cap: "The observed pupation-to-catch interval: what explains it"
tibble::tibble(
  Quantity = c(
    sprintf("Interval, mean \u00b1 sd"),
    "Mean temperature over the 14 d after pupation",
    "Correlation of the interval with that temperature",
    "Best fit to that temperature (5-df spline)",
    "Interval sd after a per-locality constant",
    "Interval sd after locality and year",
    "Correlation of the interval with latitude",
    "Correlation of the interval with July temperature"
  ),
  Value = c(
    sprintf("%.1f \u00b1 %.1f d", mean(lag_rows$lag_obs), lag_sum$sd_lag),
    sprintf("%.1f \u00b0C (%.1f\u2013%.1f)", mean(lag_rows$t_post),
            min(lag_rows$t_post), max(lag_rows$t_post)),
    sprintf("r = %+.2f", lag_sum$r_lag_tpost),
    sprintf("R\u00b2 = %.3f, residual sd %.2f d", lag_sum$spline_r2, lag_sum$spline_sd),
    sprintf("%.2f d", lag_sum$sd_site),
    sprintf("%.2f d", lag_sum$sd_site_year),
    sprintf("r = %+.2f", lag_sum$r_lag_lat),
    sprintf("r = %+.2f", lag_sum$r_lag_tjuly)
  )
) |>
  kable(align = c("l", "r"))
```

The interval does not respond to the temperature the pupae experienced: the
correlation is −0.04, where a development model needs a strong negative, and
the best smooth function of that temperature explains 6% of the variance. It
is instead a property of the **site**: `r sprintf("%.0f%%", lag_sum$var_share[1])`
of its variance sits between localities against
`r sprintf("%.0f%%", lag_sum$var_share[2])` between years, and a per-locality
constant cuts the scatter from 6.4 to 3.8 days.

Worse for the development explanation, the sign is backwards. The interval
grows with July temperature (+0.47) and falls with latitude (−0.39): the
warm southern sites, where development should be fastest, are the ones whose
catch arrives longest after predicted pupation. A pupal stage would shorten
the interval exactly where the data need it lengthened.

## What the long intervals look like

```{r}
#| label: tbl-lag-late
#| tbl-cap: "Catch falling after the modelled flight window"
tibble::tibble(
  Quantity = c("Share of the season's catch after BioSIM's flight p95, mean",
               "Correlation of that share with the interval",
               "Correlation of that share with latitude"),
  Value = c(sprintf("%.0f%%", 100 * mean(lag_rows$share_late)),
            sprintf("r = %+.2f", lag_sum$r_late_lag),
            sprintf("r = %+.2f", lag_sum$r_late_lat))
) |>
  kable(align = c("l", "r"))
```

A quarter of the catch, on average, falls after the modelled cohort has
finished flying, and that share is what distinguishes the long-interval sites
(r = 0.72). It is also a southern phenomenon (r = −0.46 with latitude). The
nightly curves show what that means.

```{r}
#| label: fig-catch-shape
#| fig-cap: "Nightly catches at the three localities with the longest and the three with the shortest pupation-to-catch interval, each in its best-sampled year. The shaded band is BioSIM's modelled flight window (p05 to p95) and the dashed line is the pupation date predicted by bayessbw. Vertical scales are free: the panels are about shape and position, not size. Top row, north to south: intervals of 3, 2 and 15 days; bottom row 22, 25 and 28 days."
#| fig-width: 9
#| fig-height: 5
library(ggplot2)

## Localities with at least three comparable years, ranked by their median
## interval; each shown in the year with the largest catch.
site_rank <- lag_rows |>
  group_by(Location) |>
  filter(n() >= 3) |>
  summarise(lag_med = median(lag_obs), .groups = "drop") |>
  arrange(lag_med)

chosen <- bind_rows(
  head(site_rank, 3) |> mutate(group = "Shortest interval"),
  tail(site_rank, 3) |> mutate(group = "Longest interval")
) |>
  left_join(lag_rows, by = "Location", relationship = "one-to-many") |>
  group_by(Location) |>
  slice_max(season_total, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(panel = sprintf("%s\n%d  (interval %.0f d)",
                         sub(", Maine$", "", Location), Year, lag_obs),
         panel = factor(panel, levels = panel[order(lag_med)]))

catch <- readRDS(file.path("data", "processed", "trap_daily.rds")) |>
  filter(status == "counted") |>
  inner_join(select(chosen, Location, Year, panel, group,
                    sim_p05, sim_p95, pup, obs_p50),
             by = c("locality" = "Location", "year" = "Year"))

ggplot(catch, aes(doy, moths)) +
  geom_rect(data = distinct(catch, panel, group, sim_p05, sim_p95),
            aes(xmin = sim_p05, xmax = sim_p95, ymin = -Inf, ymax = Inf),
            inherit.aes = FALSE, fill = "#1f6f8b", alpha = 0.13) +
  geom_vline(data = distinct(catch, panel, group, pup),
             aes(xintercept = pup), colour = "#d1670f",
             linetype = "22", linewidth = 0.5) +
  geom_col(aes(fill = group), width = 1) +
  scale_fill_manual(values = c("Longest interval" = "#b3541e",
                               "Shortest interval" = "#3b6e8f")) +
  facet_wrap(~ panel, scales = "free_y", ncol = 3) +
  labs(title = "Where the interval is long, the whole catch sits after the modelled flight window",
       subtitle = "Band: modelled flight, p05\u2013p95.  Dashed line: predicted pupation.",
       x = "Day of year", y = "Moths per night") +
  theme_minimal(base_size = 9) +
  theme(panel.grid.minor = element_blank(),
        strip.text = element_text(size = 8, lineheight = 1.1),
        plot.title = element_text(size = 10, face = "plain"),
        plot.subtitle = element_text(size = 8, colour = "grey35"),
        legend.position = "none")
```

```{r}
#| label: tbl-lag-shape
#| tbl-cap: "Shape of the observed catch curve, by the length of the interval"
lag_rows |>
  mutate(inside = 1 - share_late - share_early,
         group  = ifelse(lag_obs >= quantile(lag_obs, 0.75),
                         "Longest quarter of intervals", "The other three quarters")) |>
  group_by(group) |>
  summarise(`locality-years` = n(),
            `catch before the window` = sprintf("%.0f%%", 100 * mean(share_early)),
            `inside it` = sprintf("%.0f%%", 100 * mean(inside)),
            `after it` = sprintf("%.0f%%", 100 * mean(share_late)),
            `observed p05–p95, d` = round(mean(obs_p95 - obs_p05), 1),
            .groups = "drop") |>
  arrange(desc(`locality-years`)) |>
  kable(align = c("l", "r", "r", "r", "r"))
```

The short-interval sites catch their moths inside the modelled window and
then stop. At the long-interval sites the catch is **displaced**: 62% of it
falls after the window against 13% elsewhere, essentially none before it, and
at Brunswick Township in 1981 the entire season's catch — 99% — arrives after
the model says flight has ended.

The width, though, is unchanged: 11.4 days from p05 to p95 against 10.9 days
for the rest. A same-width curve arriving late is exactly what a slower pupal
stage would produce, which is why the result of @sec-lag is worth restating
carefully. The problem is not that the delay is implausible; it is that the
delay is not a function of temperature. It is a property of the site, and it
is largest at the warmest sites, where a development model must run faster
rather than slower.

Millinocket in 1980 is the instructive exception: catch through the modelled
window *and* a separate spike a fortnight after it. Two modes are what an
influx on top of a resident population should look like, and a systematic
search for them across the 170 curves has not been done.

## Latitude, or how many moths were caught?

The offset's spatial structure could be something simpler than geography: a
percentile estimated from a few hundred moths is a poorer statistic than one
estimated from twenty thousand, and the sites that fit worst are also sites
that catch little. Latitude and catch size turn out to be nearly independent
here (r = `r sprintf("%.2f", cor(cmp$Latitude, log10(cmp$season_total)))`
across locality-years), so the two can be separated.

```{r}
#| label: tbl-offset-covariates
#| tbl-cap: "The BioSIM offset against latitude and catch size"
cmp_cov <- cmp |> mutate(logn = log10(season_total))

f_lat  <- lm(d_p50 ~ Latitude, cmp_cov)
f_catch <- lm(d_p50 ~ logn, cmp_cov)
f_both <- lm(d_p50 ~ Latitude + logn, cmp_cov)
r2 <- function(f) summary(f)$r.squared

tibble::tibble(
  Term = c("Latitude, d per degree north",
           "Catch size, d per tenfold increase",
           "R\u00b2, latitude alone", "R\u00b2, catch alone", "R\u00b2, both",
           "Variance unique to latitude", "Variance unique to catch",
           "Variance shared"),
  Value = c(
    sprintf("%+.2f (p = %.3f)", coef(f_both)[["Latitude"]],
            summary(f_both)$coefficients["Latitude", 4]),
    sprintf("%+.2f (p = %.3f)", coef(f_both)[["logn"]],
            summary(f_both)$coefficients["logn", 4]),
    sprintf("%.3f", r2(f_lat)), sprintf("%.3f", r2(f_catch)),
    sprintf("%.3f", r2(f_both)),
    sprintf("%.3f", r2(f_both) - r2(f_catch)),
    sprintf("%.3f", r2(f_both) - r2(f_lat)),
    sprintf("%.3f", r2(f_lat) + r2(f_catch) - r2(f_both))
  )
) |>
  kable(align = c("l", "r"))
```

Both terms are significant and mostly independent: latitude carries four times
the unique variance of catch size. The next table asks the sharper question —
is this about *which localities*, or about *which years at a locality*?

```{r}
#| label: tbl-offset-within-between
#| tbl-cap: "Between localities and within them"
between <- cmp_cov |>
  group_by(Location) |>
  summarise(years = n(), lat = first(Latitude), logn = mean(logn),
            off = mean(d_p50), .groups = "drop")

## Within: each locality's own mean removed, so latitude cannot contribute.
within <- cmp_cov |> group_by(Location) |> filter(n() >= 3) |> ungroup()

tibble::tibble(
  Comparison = c(
    sprintf("Between localities (n = %d): mean offset vs latitude", nrow(between)),
    "Between localities: mean offset vs mean catch",
    sprintf("Within localities (%d rows, %d sites): offset vs catch",
            nrow(within), n_distinct(within$Location)),
    "Within localities, year means also removed"
  ),
  `r` = round(c(
    cor(between$off, between$lat),
    cor(between$off, between$logn),
    cor(residuals(lm(d_p50 ~ factor(Location), within)),
        residuals(lm(logn ~ factor(Location), within))),
    cor(residuals(lm(d_p50 ~ factor(Location) + factor(Year), within)),
        residuals(lm(logn ~ factor(Location) + factor(Year), within)))
  ), 2)
) |>
  kable(align = c("l", "r"))
```

A locality's poor years are not its low-catch years: within a site the offset
does not follow the size of the catch. The effect is between localities —
a property of *which sites*, not of how many moths a given season produced.

```{r}
#| label: tbl-offset-grid
#| tbl-cap: "Mean BioSIM offset by region and catch size"
cmp_cov |>
  mutate(region = ifelse(Latitude < 45.5, "South (below 45.5 \u00b0N)", "North"),
         band = ifelse(season_total < 1000, "under 1,000 moths", "1,000 or more")) |>
  group_by(region, band) |>
  summarise(`locality-years` = n(),
            `mean offset, d` = round(mean(d_p50), 1),
            `sd, d` = round(sd(d_p50), 1), .groups = "drop") |>
  arrange(desc(region), band) |>
  kable(align = c("l", "l", "r", "r", "r"))
```

The two effects add. Latitude is visible at either catch level and catch size
at either latitude, and the interaction is marginal
(p = `r sprintf("%.3f", summary(lm(d_p50 ~ Latitude * log10(season_total), cmp))$coefficients[4, 4])`):
the catch effect is somewhat stronger in the south. Restricted to catches of
1,000 moths or more, the latitude gradient survives at about
`r sprintf("%.1f", coef(lm(d_p50 ~ Latitude, filter(cmp, season_total >= 1000)))[2])`
d per degree against
`r sprintf("%.1f", coef(lm(d_p50 ~ Latitude, cmp))[2])`
d per degree overall, while the catch effect disappears
(r = `r sprintf("%.2f", with(filter(cmp, season_total >= 1000), cor(d_p50, log10(season_total))))`).

The worst-fitting sites are both southern and habitually small-catch, so their
offsets cannot be assigned to either cause alone. Southern localities that
caught in the thousands sit far closer to the model.

```{r}
#| label: tbl-offset-sites-catch
#| tbl-cap: "The four worst localities, and southern localities with large catches"
bind_rows(
  cmp |>
    group_by(Location) |>
    filter(n() >= 3) |>
    summarise(years = n(), lat = round(first(Latitude), 2),
              `median catch` = median(season_total),
              `median offset, d` = round(median(d_p50), 1), .groups = "drop") |>
    filter(`median offset, d` < -10) |>
    mutate(Group = "Worst fitting"),
  cmp |>
    filter(Latitude < 45.2, season_total > 5000) |>
    group_by(Location) |>
    summarise(years = n(), lat = round(first(Latitude), 2),
              `median catch` = median(season_total),
              `median offset, d` = round(median(d_p50), 1), .groups = "drop") |>
    mutate(Group = "Southern, large catch")
) |>
  relocate(Group) |>
  arrange(Group, `median offset, d`) |>
  kable(align = c("l", "l", "r", "r", "r"))
```

## Is the catch-size effect precision, or biology?

Part of it is precision. The spread of the offset is about twice as wide at
small catches as at large ones — sd
`r sprintf("%.1f", sd(filter(cmp, season_total < 1000)$d_p50))` d below a
thousand moths against
`r sprintf("%.1f", sd(filter(cmp, season_total >= 5000)$d_p50))` d above five
thousand — which is what estimating a percentile from few moths does.

Precision alone, though, produces scatter, not a shift, and a detection
threshold — the mechanism that would produce one — makes a specific
prediction: it should clip the sparse early nights, delaying the observed
start and *narrowing* the observed curve. The data go the other way.

```{r}
#| label: tbl-offset-shape-catch
#| tbl-cap: "Shape of the observed catch curve against catch size"
tibble::tibble(
  Quantity = c("Share of catch after the modelled flight window",
               "Share of catch before it",
               "Observed p05\u2013p95 width"),
  `r with log10 catch` = round(c(
    cor(lag_rows$share_late, log10(lag_rows$season_total)),
    cor(lag_rows$share_early, log10(lag_rows$season_total)),
    cor(lag_rows$obs_p95 - lag_rows$obs_p05, log10(lag_rows$season_total))
  ), 2)
) |>
  kable(align = c("l", "r"))
```

Small catches carry proportionally **more** of their catch after the modelled
window, not less before it, and their curves are slightly **wider**. That is
the opposite of clipping, and it is the same signature as the displaced
southern curves of @sec-lag.

It also sharpens the immigration hypothesis into something testable. If the
late weight came from moths arriving rather than from the local cohort, it
should matter most where the local population is small — which is what the
negative correlation with catch size says — and those late nights should
coincide with conditions that carry moths, which the extracted wind fields can
be asked directly. The alternative reading is that a small catch reflects a
site marginal for the insect, where the local cohort is both sparse and
genuinely late; distinguishing the two needs the nightly weather, not more
summary statistics.

## Open questions for the comparison

- **The spatial residual is the next problem, and neither the weather nor the
  missing pupal stage explains it.** Latitude is a proxy, not an explanation.
  The two models run on different weather and make the same spatial error, and
  @sec-lag rules out a temperature-driven pupal stage as the missing piece.
  What is left is what both models share — development with no site-level
  host, shading or microclimate modifier — and what neither contains:
  immigration. @sec-lag narrows it further: the southern catch curves are
  displaced late by a week or more at unchanged width, so whatever is missing
  acts like a site-fixed delay rather than a broadening. @sec-sync narrows it
  again, and changes its shape: the observed dates carry almost no
  latitudinal gradient while both models impose one of about four days per
  degree, so this is a gradient-slope error rather than a southern site
  effect.
- **Trap attraction is not modelled at all.** The observed curve is a catch
  curve. Nightly weather gates flight and gates trapping, and neither is in the
  comparison yet; the hourly CaSR variables extracted for every locality-year —
  temperature, wind, precipitation — were extracted for exactly this purpose.
- **The comparable set could be larger.** Relaxing the tail threshold from 5%
  to 10% would return 12 further records and the years 1987 and 1988; whether
  the added truncation is acceptable is a judgement the sensitivity table in
  @sec-traps supports.

# Is flight synchronous across Maine? {#sec-sync}

```{r}
#| label: sync-data
synchrony <- readRDS(file.path("data", "processed", "flight_synchrony.rds"))
sync_grad  <- synchrony$gradients
sync_gaps  <- synchrony$gaps
sync_pairs <- synchrony$pairs
sync_sum   <- synchrony$summary
```

@sec-compare leaves both models early in the south, and @sec-lag rules out a
missing pupal stage as the cause. That leaves a site-level effect neither
model contains, and immigration — a locality's catch need not be its own
cohort. `R/07_flight_synchrony.R` tests the second without using weather at
all, by asking whether southern catches happen on the region's dates or on
their own. Two predictions separate the explanations: a wave of moths arriving
from the north would flatten the latitudinal gradient in catch dates, and at
nightly resolution it would make distant localities spike on the same
calendar nights.

## The latitudinal gradient

```{r}
#| label: tbl-sync-gradient
#| tbl-cap: "Days per degree of latitude, with year fixed effects so that the interannual signal common to all localities is removed"
## Column names are set afterwards: R does not accept \u escapes inside
## backtick-quoted names.
sync_grad |>
  transmute(Series = series, n,
            per_degree = sprintf("%+.2f", d_per_degree),
            ci = sprintf("%+.2f to %+.2f", lo, hi),
            spread = sprintf("%.1f", d_per_degree * sync_sum$lat_span)) |>
  setNames(c("Series", "n", "d per \u00b0N", "95% CI", "implied spread, d")) |>
  kable(align = c("l", "r", "r", "r", "r"))
```

Over the `r sprintf("%.1f", sync_sum$lat_span)` degrees of latitude the trap
network spans, the observed median catch date moves by
**`r sprintf("%.1f", sync_sum$obs_spread)` days**; BioSIM moves it by
`r sprintf("%.1f", sync_sum$sim_spread)` and bayessbw by
`r sprintf("%.1f", sync_sum$bayes_spread)`. The observed gradient is not
distinguishable from zero, and it stays flat on the best-sampled records while
BioSIM's stays steep on the same rows.

```{r}
#| label: fig-sync-gradient
#| fig-cap: "Median catch date and modelled median flight date against latitude, both with the year mean removed so that localities from different years are comparable. Lines are ordinary least squares. The observed dates carry almost no latitudinal gradient; both models impose one of about four days per degree."
#| fig-width: 6.5
#| fig-height: 4.2
library(ggplot2)

cmp_anom <- cmp |>
  group_by(Year) |>
  mutate(obs_a = obs_p50 - mean(obs_p50),
         sim_a = sim_p50 - mean(sim_p50)) |>
  ungroup() |>
  select(Latitude, obs_a, sim_a) |>
  tidyr::pivot_longer(c(obs_a, sim_a), names_to = "which", values_to = "anom") |>
  mutate(which = factor(ifelse(which == "obs_a", "Observed catch",
                               "BioSIM flight"),
                        levels = c("BioSIM flight", "Observed catch")))

ggplot(cmp_anom, aes(Latitude, anom, colour = which, fill = which)) +
  geom_hline(yintercept = 0, colour = "grey45", linetype = "22") +
  geom_point(size = 1.6, alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x, linewidth = 0.8, alpha = 0.18) +
  scale_colour_manual(values = c("BioSIM flight" = "#1f6f8b",
                                 "Observed catch" = "#b3541e")) +
  scale_fill_manual(values = c("BioSIM flight" = "#1f6f8b",
                               "Observed catch" = "#b3541e")) +
  labs(title = "The catches carry no latitudinal gradient; the model imposes one",
       x = "Latitude (\u00b0N)",
       y = "Days from that year's mean",
       colour = NULL, fill = NULL) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(size = 10, face = "plain"),
        legend.position = "inside",
        legend.position.inside = c(0.02, 0.02),
        legend.justification = c(0, 0),
        legend.background = element_blank())
```

```{r}
#| label: tbl-sync-gaps
#| tbl-cap: "South minus north in each year with at least two localities in both halves of the network (split at 45.5 \u00b0N)"
sync_gaps |>
  transmute(Year = as.character(Year),
            `northern localities` = n_north, `southern` = n_south,
            `observed gap, d` = round(obs_gap, 1),
            `modelled gap, d` = round(sim_gap, 1)) |>
  kable(align = c("l", "r", "r", "r", "r"))
```

Averaged over those `r nrow(sync_gaps)` years the southern localities catch
their moths `r sprintf("%.1f", abs(sync_sum$obs_gap))` days before the
northern ones; the model puts them `r sprintf("%.1f", abs(sync_sum$sim_gap))`
days ahead. Seasonal timing across the network is, to within a couple of days,
the same everywhere.

This reframes the residual of @sec-compare. What looked like a southern site
effect is a **gradient-slope error**: both models spread flight across Maine
about seven times more than the catches do, and the offset we were measuring
is the difference between a steep modelled gradient and a flat observed one.
It also explains why both models show it despite running on different weather,
and why the worst sites are the extreme ones.

## Nightly synchrony

A flat seasonal gradient is consistent with immigration, but also with a
regionally synchronous emergence that the models exaggerate. The nightly
curves separate the two: arriving moths should make distant localities spike
together.

```{r}
#| label: tbl-sync-pairs
#| tbl-cap: "Correlation of nightly catch shares between localities trapped in the same year. The null pairs the same two localities in different years, averaged over 20 draws"
sync_pairs |>
  mutate(pair_type = recode(pair_type, NN = "North \u2013 north",
                            NS = "North \u2013 south", SS = "South \u2013 south")) |>
  group_by(`Pair` = pair_type) |>
  summarise(pairs = n(),
            `median r` = round(median(r0), 2),
            `% above 0.5` = round(100 * mean(r0 > 0.5)),
            `median distance, km` = round(median(dist_km)),
            `median best lag, d` = median(lag_best, na.rm = TRUE),
            .groups = "drop") |>
  kable(align = c("l", "r", "r", "r", "r", "r"))
```

Over all `r sync_sum$n_pairs` pairs the median correlation is
`r sprintf("%.2f", sync_sum$r0_median)` against a shuffled-year null of
`r sprintf("%.2f", sync_sum$null_median)`, and
`r sprintf("%.0f%%", 100 * sync_sum$r0_above_half)` of pairs exceed 0.5
against `r sprintf("%.0f%%", 100 * sync_sum$null_above_half)` in the null. So
same-night synchrony is real. It is also **local**:

```{r}
#| label: tbl-sync-distance
#| tbl-cap: "Nightly synchrony against separation"
sync_pairs |>
  mutate(band = cut(dist_km, c(0, 50, 100, 200, 400, Inf),
                    labels = c("under 50 km", "50\u2013100", "100\u2013200",
                               "200\u2013400", "over 400"))) |>
  group_by(`Separation` = band) |>
  summarise(pairs = n(), `median r` = round(median(r0), 2),
            `% above 0.5` = round(100 * mean(r0 > 0.5)), .groups = "drop") |>
  kable(align = c("l", "r", "r", "r"))
```

Synchrony decays with distance, and the north–south pairs are the weakest of
the three types despite having the same null. The best lag is zero days for
every pair type, so there is no systematic delay of the kind transport would
impose.

That is the opposite of what a northward wave delivering the southern catch
would produce. What the nightly data show is nearby traps sharing flight
nights — local weather gating flight and trappability — and southern traps
sharing few nights with northern ones.

## What this says about immigration

The hypothesis survives at the seasonal scale and fails at the nightly one.
The more parsimonious reading of both results is that real flight across
Maine is nearly synchronous and that both development models exaggerate how
strongly latitude and temperature spread it: the gradient is wrong, rather
than the southern catch being foreign.

One caveat keeps immigration alive: arrivals need not be simultaneous across
localities, because the wind fields differ from night to night, so weak
north–south nightly correlation is not proof of absence.

A second candidate — that the gradient is really topography, since BioSIM was
run with `elev = NaN` and the CaSR extraction carries no terrain — is tested
in @sec-elev below and ruled out.

That leads directly to a test of how much site-specific weather is buying us
at all, which is @sec-regional below: if flight is regionally synchronous, one
regional weather series should predict the catches about as well as a model
run at each trap's own coordinates.

## Is the gradient an elevation error? {#sec-elev}

```{r}
#| label: elev-data
elev_check <- readRDS(file.path("data", "processed", "elevation_check.rds"))
elev_sum   <- elev_check$summary
```

Elevation is the obvious suspect for a gradient that is too steep. `R/03`
leaves it unspecified, so BioSIM takes it from its own digital elevation
model, and elevation is itself a latitude proxy in this network: the southern
localities are coastal lowlands and the northern ones sit above 300 m
(median `r sprintf("%.0f", elev_sum$elev_south)` m against
`r sprintf("%.0f", elev_sum$elev_north)` m, r =
`r sprintf("%.2f", elev_sum$elev_lat_r)` with latitude).
`R/08_elevation_check.R` fetches an elevation for every locality from the USGS
3DEP point-query service — `r sprintf("%.0f to %.0f", elev_sum$elev_range[1], elev_sum$elev_range[2])`
m, median `r sprintf("%.0f", elev_sum$elev_median)` — and asks BioSIM the same
question twice, once with its own elevation model and once with the measured
value.

Elevation does move the model. Over the plausible range of 100 to 500 m the
flight date shifts by
`r paste(sprintf("%+.1f d per 100 m at %s", elev_sum$sens_per_100m$d, sub(",.*", "", elev_sum$sens_per_100m$Location)), collapse = " and ")`,
against a run-to-run noise of about a tenth of a day. The two signs are not a
typo: at the coastal locality a *higher* assumed elevation brings flight
*earlier*, which cannot be a lapse rate and suggests that elevation also
enters BioSIM's choice and weighting of source weather stations — coastal
stations at low elevation have cool summers. The mechanism is not established
here; what matters is that the model is demonstrably sensitive.

```{r}
#| label: tbl-elev-test
#| tbl-cap: "Modelled flight with BioSIM's own elevation model against modelled flight with the measured elevation, over a sample of locality-years spanning the latitude range"
tibble::tibble(
  Quantity = c("Locality-years compared",
               "Mean shift when the measured elevation is supplied",
               "Mean shift, southern localities",
               "Mean shift, northern localities",
               "Mean absolute offset against the catches",
               "Latitude gradient of modelled flight",
               "Latitude gradient of the observed catches"),
  Value = c(
    sprintf("%d", elev_sum$n_preview),
    sprintf("%+.1f d", elev_sum$mean_shift),
    sprintf("%+.1f d", elev_sum$shift_south),
    sprintf("%+.1f d", elev_sum$shift_north),
    sprintf("%.1f d \u2192 %.1f d", elev_sum$abs_off_dem, elev_sum$abs_off_usgs),
    sprintf("%+.2f \u2192 %+.2f d per \u00b0N", elev_sum$grad_dem, elev_sum$grad_usgs),
    sprintf("%+.2f d per \u00b0N", elev_sum$grad_obs)
  )
) |>
  kable(align = c("l", "r"))
```

BioSIM's elevation model is already right at these localities. Supplying the
measured value moves the flight date by a few tenths of a day — of the same
order as the replicate-to-replicate variation of the service itself — leaves
the mean offset against the catches where it was, and does not touch the
gradient, which stays near six days per degree against an observed tenth of a
day. A full re-run with elevation, one switch in `R/03`, would not change any
conclusion in this log, so it has not been made.

## How precisely does the location have to be right?

The elevation work turned up something that does matter. A modelled date at a
coastal locality is not a smooth function of where the point is: displacing
the same trap eastwards in steps, with BioSIM's own elevation model each time,
moves the answer by three days inside the first kilometre and then wanders
back.

```{r}
#| label: tbl-elev-coords
#| tbl-cap: "The coastal locality of the scan, displaced eastwards in steps of longitude"
elev_check$coords |>
  transmute(`displacement, km` = round(km, 2),
            `flight p50` = round(p50, 1),
            `shift, d` = sprintf("%+.1f", shift)) |>
  kable(align = c("r", "r", "r"))
```

That is not a lapse rate or a distance effect; it is what a discrete set of
source weather stations looks like when a point crosses from one
interpolation neighbourhood into another. Two consequences. A modelled date
cannot be attributed to a trap more finely than a few days at a coastal site,
which matters because the recorded coordinates in this dataset carry a stated
uncertainty of one to eight kilometres — though that uncertainty does not
predict the size of the offset (@sec-compare), so this behaves as noise rather
than as a bias. And a locality must always be queried at its recorded
coordinates: an earlier version of this work used a coordinate rounded to two
decimals from memory, landed 1.5 km away, and read that as evidence that
BioSIM's elevation model was wrong at the coast. It was not.

## Does site-specific weather help at all? {#sec-regional}

```{r}
#| label: regional-data
regional <- readRDS(file.path("data", "processed", "regional_weather_test.rds"))
reg_scores <- regional$scores
reg_rows   <- regional$rows
reg_sum    <- regional$summary
```

If flight is as synchronous as @sec-sync suggests, a model driven by one
regional weather series — a single prediction per year, the same for every
locality — should predict the catches about as well as the site-specific runs.
`R/09_regional_weather_test.R` builds that comparison. The regional BioSIM run
is one point at the centroid of the trap network
(`r sprintf("%.2f", reg_sum$centroid$Latitude)` °N,
`r sprintf("%.2f", abs(reg_sum$centroid$Longitude))` °W); the regional
bayessbw run is the same simulation as @sec-lag driven by the hourly
temperature averaged over all localities. Each predictor is given one
intercept, fitted on the other years, so a constant bias is not charged
against any of them and none is scored on its own year.

```{r}
#| label: tbl-regional
#| tbl-cap: "Predicting the observed median catch date over the 170 comparable locality-years, with a leave-one-year-out intercept"
reg_scores |>
  transmute(Predictor = predictor,
            `varies by site` = ifelse(varies_by_site, "yes", "no"),
            n,
            `RMSE, d` = round(rmse, 2),
            `MAE, d` = round(mae, 2),
            `r` = round(r, 2)) |>
  kable(align = c("l", "l", "r", "r", "r", "r"))
```

Site-specific weather does not merely fail to help. **It makes both models
worse.** Replacing the site-specific run with a single regional series changes
the error by `r sprintf("%+.2f", reg_sum$rmse_change)` days for BioSIM and
`r sprintf("%+.2f", reg_sum$bayes_change)` days for bayessbw, and raises the
correlation with the observed dates from 0.49 to 0.67 and from 0.40 to 0.52.
One number per year beats a model run at each trap's own coordinates.

The last line of the table is the floor: the mean observed date of the *other*
localities trapped in the same year, which is what a perfect regional
predictor would look like. At `r sprintf("%.2f", reg_scores$rmse[nrow(reg_scores)])`
days it is only
`r sprintf("%.2f", reg_scores$rmse[2] - reg_scores$rmse[nrow(reg_scores)])`
days better than regional BioSIM, so the regional run already captures most
of the predictable signal, and what the site-specific runs add on top of it is
error.

```{r}
#| label: tbl-regional-years
#| tbl-cap: "The regional predictions year by year, against the mean observed date. The intercepts of @tbl-regional are not applied here, so the columns differ by a constant"
reg_rows |>
  group_by(Year) |>
  summarise(localities = n(),
            `observed mean` = round(mean(obs_p50), 1),
            `BioSIM regional` = round(first(biosim_regional), 1),
            `BioSIM site mean` = round(mean(biosim_site), 1),
            `bayessbw regional` = round(first(bayes_regional), 1),
            .groups = "drop") |>
  mutate(Year = as.character(Year)) |>
  kable(align = c("l", "r", "r", "r", "r", "r"))
```

Two caveats on the comparison. The regional predictors cannot capture real
site-to-site variation, because they are constant within a year — the result
is not that no such variation exists, but that it is smaller than the error
the models make when they try to represent it. And averaging temperature
before running a development model is not the same as averaging the
development of each site, since the rate curves are non-linear; the regional
bayessbw run is a model of the average weather, not an average of the models.

Taken with @sec-sync this is the practical conclusion of the comparison. For
predicting when spruce budworm flies in Maine, the useful signal in these
models is the year, not the site; and a phenology model applied at a single
trap's coordinates will be wrong by about six days, mostly because it imposes
a latitudinal gradient the moths do not follow. @sec-count turns that finding
into a predictive model and measures what it can do.

# Predicting the nightly catch curve {#sec-count}

```{r}
#| label: count-data
count_model <- readRDS(file.path("data", "processed", "nightly_count_model.rds"))
cnt_scores  <- count_model$scores
cnt_cv      <- count_model$cv
cnt_nights  <- count_model$nights
cnt_sum     <- count_model$summary

## Headline numbers for the prose. Inline `r` cannot carry escaped backticks,
## so anything needing a backticked column name is computed here.
cnt_best <- cnt_cv |>
  filter(model == "reg+") |>
  summarise(mae    = mean(abs(err_p50)),
            rmse   = sqrt(mean(err_p50^2)),
            bias   = mean(err_p50),
            within = mean(abs(err_p50) <= 5))

cnt_best_all <- cnt_cv |>
  group_by(model) |>
  summarise(mae      = mean(abs(err_p50)),
            per_moth = sum(loglik * nights_n) / sum(total),
            .groups  = "drop")
```

Everything so far has compared summary dates. This section asks the
predictive question directly: given a development model, how well can the
**shape of a season** be predicted at a locality — the share of its catch
falling on each night?

Abundance is deliberately not predicted. Nothing in this project forecasts
population density, so every model below carries one free level per
locality-year that absorbs the season total and leaves only timing to
explain. A Poisson likelihood with such a level per group is a multinomial
over that group's nights, so the fitted values read directly as predicted
shares, and a held-out locality-year needs no level of its own.

`R/10_nightly_count_model.R` fits `r nrow(cnt_scores)` such models to
`r format(cnt_sum$n_nights, big.mark = ",")` trapped nights from
`r cnt_sum$n_ly` locality-years
(`r sprintf("%.0f%%", 100 * cnt_sum$zero_share)` of nights caught nothing) and
scores them leave-one-year-out.

| Model | What drives the shape |
|-------|-----------------------|
| `clim` | a seasonal curve only, `ns(day of year)` fitted from the other years — the climatological baseline any predictor must beat |
| `raw` | BioSIM's site curve used as is, no fitted parameters |
| `site` | the same curve with a free exponent |
| `site+` | plus a spline correction in days from the modelled median |
| `reg+` | the same, driven by one regional BioSIM curve per year |
| `reg+lat` | plus a latitude interaction |
| `pup+` | the same calibration applied to bayessbw's predicted pupation distribution instead of BioSIM's flight curve, locality by locality |
| `pupreg+` | and to one regional bayessbw run per year |
| `reg-med` | the regional curve stripped out entirely, leaving only the spline in days from its median: the development model contributes a date and nothing else |

: The nine specifications {#tbl-count-models}

The two bayessbw rows need no lag constant. The correction is a spline in days
from the driving curve's own median, so the interval between pupation and the
catch is fitted rather than assumed — which is what made the shifted series on
@fig-compare uninformative about level and is repaired here.

```{r}
#| label: tbl-count-scores
#| tbl-cap: "Leave-one-year-out skill. Log-likelihood is the multinomial log-likelihood of the observed counts given the predicted shares, reported per moth so that it reads as a mean log predicted share"
cnt_cv |>
  group_by(model) |>
  summarise(`log-lik per moth` = sum(loglik * nights_n) / sum(total),
            `mean |p50 error|, d` = mean(abs(err_p50)),
            `RMSE p50, d` = sqrt(mean(err_p50^2)),
            `bias, d` = mean(err_p50),
            `within 5 d` = sprintf("%.0f%%", 100 * mean(abs(err_p50) <= 5)),
            .groups = "drop") |>
  mutate(model = factor(model, levels = levels(cnt_scores$model))) |>
  arrange(model) |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  kable(align = c("l", "r", "r", "r", "r", "r"))
```

Four things come out of that table.


**A calibrated regional curve is the best predictor**, and it is a usable one:
`r sprintf("%.1f", cnt_best$mae)` days of mean absolute error on the median
catch date, a bias of `r sprintf("%+.1f", cnt_best$bias)` days, and
`r sprintf("%.0f%%", 100 * cnt_best$within)` of locality-years inside five
days — all out of sample, and with no information from the year being
predicted.

**BioSIM's raw site curve is worse than climatology.** Used as delivered it
loses to a seasonal average fitted from other years, on both the likelihood
and the median. Giving it a free exponent helps, a shape correction does not,
and replacing the site curve with the single regional one helps most. Latitude
makes things worse again, for the fourth time in this log.

**The pupation distribution does not do better than the flight curve.**
Applied locality by locality it lands where BioSIM's site curve does
(`r sprintf("%.2f", cnt_best_all$mae[cnt_best_all$model == "pup+"])` days
against `r sprintf("%.2f", cnt_best_all$mae[cnt_best_all$model == "site+"])`),
and the regional bayessbw run is the weaker of the two
(`r sprintf("%.2f", cnt_best_all$mae[cnt_best_all$model == "pupreg+"])` days)
— the opposite of BioSIM, where going regional helped most. Both bayessbw
specifications are almost unbiased, which is the spline absorbing the
pupation-to-catch interval as intended, but they are less sharp: a mean log
predicted share of
`r sprintf("%.2f", cnt_best_all$per_moth[cnt_best_all$model == "pup+"])`
and `r sprintf("%.2f", cnt_best_all$per_moth[cnt_best_all$model == "pupreg+"])`
against `r sprintf("%.2f", cnt_best_all$per_moth[cnt_best_all$model == "reg+"])`
for the best. Two independent development models place pupation within a day
of each other, so this is not a disagreement about the date; it is that
bayessbw's year-to-year placement carries less of the signal the catches
follow, consistent with its weaker year-mean correlation in
@tbl-compare-structure.

**Almost all of the shape is empirical.** The fitted exponent on the regional
curve is `r sprintf("%.2f", cnt_sum$exponent)`, so the modelled curve enters
the prediction heavily damped. `reg-med` settles how much it is worth by
removing it altogether and keeping only the spline in days from its median:
that costs nothing on timing
(`r sprintf("%.2f", cnt_best_all$mae[cnt_best_all$model == "reg-med"])` days
of mean absolute error against
`r sprintf("%.2f", cnt_best_all$mae[cnt_best_all$model == "reg+"])`, the same
`r sprintf("%.0f%%", 100 * mean(abs(cnt_cv$err_p50[cnt_cv$model == "reg-med"]) <= 5))`
inside five days) and a little on the nightly likelihood
(`r sprintf("%.2f", cnt_best_all$per_moth[cnt_best_all$model == "reg-med"])`
against `r sprintf("%.2f", cnt_best_all$per_moth[cnt_best_all$model == "reg+"])`
per moth). So what the development model contributes is *when* the season is
centred — which year runs early, which runs late — and the curve's own form
adds only a marginal sharpening of the nightly fit.

Measured on the trapped nights, where the fit is supported, the correction
moves the predicted median `r sprintf("%+.1f", cnt_sum$shift)` days relative
to BioSIM's curve as delivered, with a spread of
`r sprintf("%.1f", cnt_sum$shift_sd)` days across locality-years, and it
widens the predicted central 90% from
`r sprintf("%.1f", cnt_sum$raw_width)` to `r sprintf("%.1f", cnt_sum$cal_width)`
days. The observed central 90% is `r sprintf("%.1f", cnt_sum$obs_width)` days,
so the calibrated prediction is well centred but too broad: it is a good
predictor of the median date and an over-dispersed one of the season's spread.

```{r}
#| label: fig-count-skill
#| fig-cap: "Out-of-sample prediction of the median catch date. Left: the calibrated regional model against the observed median, one point per locality-year, with the 1:1 line. Right: the distribution of signed errors for each specification; the box spans the quartiles, the line is the median, whiskers reach 1.5 times the interquartile range."
#| fig-width: 9
#| fig-height: 4.2
library(ggplot2)
library(patchwork)

best <- filter(cnt_cv, model == "reg+")
lims <- range(c(best$obs_p50, best$pred_p50))

p_pred <- ggplot(best, aes(obs_p50, pred_p50)) +
  geom_abline(slope = 1, intercept = 0, linetype = "22", colour = "grey45") +
  geom_point(size = 1.8, alpha = 0.7, colour = "#1f6f8b") +
  annotate("text", x = lims[2], y = lims[2], hjust = 1, vjust = 1.8,
           size = 3, colour = "grey45", label = "1:1") +
  coord_equal(xlim = lims, ylim = lims) +
  labs(title = "Held-out years are predicted to about three days",
       x = "Observed median catch date (DOY)",
       y = "Predicted median (DOY)")

ord <- cnt_scores$model
p_err <- cnt_cv |>
  mutate(model = factor(model, levels = ord)) |>
  ggplot(aes(model, err_p50)) +
  geom_hline(yintercept = 0, colour = "grey45", linetype = "22") +
  geom_boxplot(aes(fill = model == "reg+"), width = 0.6, outlier.size = 0.7,
               linewidth = 0.4) +
  scale_fill_manual(values = c(`TRUE` = "#1f6f8b", `FALSE` = "grey85")) +
  labs(title = "and the regional curve beats the rest",
       x = NULL, y = "Predicted minus observed (d)") +
  theme(legend.position = "none")

base <- theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(size = 10, face = "plain", hjust = 0),
        legend.position = "none")

(p_pred & base) | (p_err & base + theme(axis.text.x = element_text(angle = 30, hjust = 1)))
```

```{r}
#| label: tbl-count-residual
#| tbl-cap: "Where the calibrated regional model still misses"
best_lat <- best |>
  left_join(distinct(cnt_nights, site_year, Latitude), by = "site_year")

tibble::tibble(
  Quantity = c("Mean |error|, southern localities (below 45.5 \u00b0N)",
               "Mean |error|, northern localities",
               "Error against latitude",
               "Error against log10 season total",
               "Worst locality-year"),
  Value = c(
    sprintf("%.2f d", mean(abs(best_lat$err_p50[best_lat$Latitude <  45.5]))),
    sprintf("%.2f d", mean(abs(best_lat$err_p50[best_lat$Latitude >= 45.5]))),
    sprintf("r = %+.2f", cor(best_lat$err_p50, best_lat$Latitude)),
    sprintf("r = %+.2f", cor(best_lat$err_p50, log10(best_lat$total))),
    with(best[which.max(abs(best$err_p50)), ],
         sprintf("%s %d, %+.1f d", sub(", Maine$", "", Location), Year, err_p50))
  )
) |>
  kable(align = c("l", "r"))
```

Calibration removes most of the spatial structure: the residual's correlation
with latitude falls to
`r sprintf("%+.2f", cor(best_lat$err_p50, best_lat$Latitude))`
from the +0.49 of @sec-compare. A southern excess survives — mean absolute
error `r sprintf("%.1f", mean(abs(best_lat$err_p50[best_lat$Latitude < 45.5])))`
days against `r sprintf("%.1f", mean(abs(best_lat$err_p50[best_lat$Latitude >= 45.5])))`
in the north — so the gradient is corrected on average rather than explained.

## What this model is and is not good for

It predicts **timing**, out of sample, to about three days on the median and
within five days four times out of five. That is enough to say when a flight
season will be centred at a Maine locality given the year's weather, which is
the operational question. Its predicted **spread** is too wide by about half
again, so a prediction interval taken straight from the fitted curve will be
conservative.

It does not predict **how many** moths will be caught: the season total is an
input, not an output. It also does not separate flight from trappability.
Nightly weather is absent by design, and the model's own diagnostics show
where that would pay: `r sprintf("%.0f%%", 100 * cnt_sum$zero_share)` of nights
caught nothing, and the count variance is several thousand times the mean, so
the nightly residual is dominated by night-to-night variation that a
development model cannot address. Temperature, wind and precipitation at each
locality-night are already extracted and would enter this model as covariates
without changing anything else about it.

One technical caveat for anyone refitting: the correction spline is a natural
spline in days from the modelled median, and its behaviour beyond about
\u00b115 days rests on few nights — the upturn at the far right of the fitted
correction is an edge artefact, not a second flight period.
