---
title: "Spruce budworm flight phenology in Maine"
subtitle: "Analysis log"
author: "JN Candau"
date: today
date-format: "D MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk:
    comment: "#>"
---

```{r}
#| label: setup
## Paths are written relative to the project root, not to analysis/.
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r}
#| label: data
library(dplyr)
library(knitr)
library(lubridate)   # month(), yday(), leap_year() in the weather section

trap_records <- readRDS(file.path("data", "processed", "trap_records.rds"))
trap_daily   <- readRDS(file.path("data", "processed", "trap_daily.rds"))
```

# Purpose of this document

This is the running log of the analyses of spruce budworm (*Choristoneura
fumiferana*) flight phenology from daily light-trap catches in Maine. Each
section records what was done, the choices that were made and why, and the
state of the data or the model that resulted. It is updated as the work
proceeds; the numbers below are recomputed from the processed data every time
the document is rendered, so they cannot drift away from the files on disk.

Code lives in `R/`, source data in `data/raw/`, processed tables in
`data/processed/`. The project is version-controlled at
<https://github.com/jncandau/Maine_Oct_2026>.

# Trap data {#sec-traps}

## Source

The captures come from the eastern spruce budworm phenology dataset of
Fournier, Candau et al. (2027), held in the private repository
`jncandau/SBWFieldPhenoDatabase` and copied unmodified into
`data/raw/SBWFieldPhenoDatabase/`. The workbook
`Eastern Spruce Budworm Phenology V1.ods` holds three tables — `Phenology`,
`Flight` and `References` — joined by `referenceID`. Only `Flight` is used
here.

`Flight` is a wide table: one row per trapping record, with the captures of
each night in the columns `mothsDay160` … `mothsDay231` (day of year 160 to
231). A record is a site × year × sampling method × sex combination, and
carries its coordinates, the reported season total `mothCountYear`, and a
pointer to the source publication. Across North America it holds 763 records;
Maine contributes 643 light-trap records and 36 yard-light records.

## Reading and decoding

`R/01_read_trap_data.R` reads the sheet with every column forced to character,
because the day columns mix numbers with two text codes:

| Cell | Meaning | Handling |
|------|---------|----------|
| `-` | missing, no count available | dropped; these cells fill the whole out-of-season range |
| `m` | missing, no count available | kept as a row with `moths = NA` and `status = "missing"` |
| number | nightly capture | kept, `status = "counted"` |

: Day-cell codes in the `Flight` sheet {#tbl-codes tbl-colwidths="[10,40,50]"}

Both codes mean missing data (confirmed 2026-09-26). They are treated
differently only because of where they occur: `-` pads every record out to
DOY 231, so keeping those cells would invent trap-nights, whereas `m` only
ever appears inside a trapping season. Interior `-` cells — those falling
between a record's first and last counted night — are real gaps, and are
counted in `n_gap_nights` rather than lost.

The wide table is then pivoted to one row per record-night, and a calendar
date is derived from the year and the day of year.

## Selection

Four criteria define the working dataset, all set as constants at the top of
`R/01_read_trap_data.R`:

| Criterion | Value | Reason |
|-----------|-------|--------|
| `keep_states` | `ME` | the study area |
| `keep_methods` | `Light trap` | excludes the 36 opportunistic yard-light records |
| `keep_years` | 1968–1989 | overlap with the CaSR v3.2 reanalysis in `data/raw/Weather_Maine` (1968–1991) |
| `min_season_total` | > 99 moths | a smaller catch cannot describe a flight curve |

: Selection criteria {#tbl-selection}

The year window drops 1961–1967 and 2016. The catch threshold removes a
further 230 in-window records; it also removes, without needing a separate
rule, every record that reports only a season total and no daily series, and
all of 1963–1967 and 2016, where no trap anywhere in Maine exceeded 99 moths.

## Pooling of sex-split records

Two 1980 localities report the male and the female catch of one trap as
separate records. Since the trap is the sampling unit, the pairs are summed
night by night before the catch threshold is applied, and the components are
removed. Both pairs sampled identical nights with no missing values, so the
sums are exact.

```{r}
#| label: tbl-pooled
#| tbl-cap: "Pooled sex-split records"
trap_records |>
  filter(!is.na(pooled_from)) |>
  transmute(flightID, `components` = pooled_from, year, locality,
            `nights` = n_nights, `season total` = season_total) |>
  kable()
```

Pooling matters beyond tidiness: at Bare Island neither sex reached the
threshold alone (83 and 78 moths) but the trap caught 161, so the locality
re-enters the dataset.

## The resulting dataset

Two tables are written to `data/processed/`:

- `trap_records.rds` — one row per trapping record, with sampling effort,
  season total and provenance;
- `trap_daily.rds` — one row per record-night that was inside a trapping
  season, holding the count and its `status`.

```{r}
#| label: tbl-overview
#| tbl-cap: "The working dataset at a glance"
tibble::tibble(
  Quantity = c("Trapping records", "Localities", "Years", "Trap-nights with a count",
               "Nights coded `m` (missing)", "Interior gap nights (`-`)",
               "Moths counted", "Season total, median (range)",
               "Nights per record, median (range)"),
  Value = c(
    format(nrow(trap_records), big.mark = ","),
    format(n_distinct(trap_records$locality), big.mark = ","),
    sprintf("%d (%d\u2013%d)", n_distinct(trap_records$year),
            min(trap_records$year), max(trap_records$year)),
    format(sum(trap_daily$status == "counted"), big.mark = ","),
    format(sum(trap_daily$status == "missing"), big.mark = ","),
    format(sum(trap_records$n_gap_nights), big.mark = ","),
    format(sum(trap_daily$moths, na.rm = TRUE), big.mark = ","),
    sprintf("%s (%s\u2013%s)",
            format(median(trap_records$season_total), big.mark = ","),
            format(min(trap_records$season_total), big.mark = ","),
            format(max(trap_records$season_total), big.mark = ",")),
    sprintf("%d (%d\u2013%d)", median(trap_records$n_nights),
            min(trap_records$n_nights), max(trap_records$n_nights))
  )
) |>
  kable(align = c("l", "r"))
```

```{r}
#| label: tbl-per-year
#| tbl-cap: "Sampling effort and catch by year"
trap_daily |>
  group_by(year) |>
  summarise(
    traps       = n_distinct(flightID),
    localities  = n_distinct(locality),
    nights      = sum(status == "counted"),
    `nights/trap` = round(sum(status == "counted") / n_distinct(flightID), 1),
    `first DOY` = min(doy),
    `last DOY`  = max(doy),
    moths       = sum(moths, na.rm = TRUE),
    .groups     = "drop"
  ) |>
  mutate(year = as.character(year)) |>   # a year is a label, not a count
  kable(format.args = list(big.mark = ","))
```

The network is a rotating panel rather than a fixed set of stations: most
localities contribute a handful of years, and the annual number of traps
varies with the outbreak and with the Maine Forest Service's programme.

```{r}
#| label: tbl-site-years
#| tbl-cap: "Localities by the number of years they contribute"
trap_records |>
  group_by(locality) |>
  summarise(years = n_distinct(year), .groups = "drop") |>
  count(years, name = "localities") |>
  arrange(years) |>
  kable()
```

### Data quality

The following checks are run by `R/01_read_trap_data.R` at every read. They
report problems; they never repair them.

```{r}
#| label: tbl-quality
#| tbl-cap: "Quality checks on the retained records"
gaps <- filter(trap_records, n_gap_nights > 0)
miss <- trap_daily |> filter(status == "missing") |> distinct(flightID)
tibble::tibble(
  Check = c("Season total disagrees with the sum of the daily counts",
            "Locality-year with more than one record",
            "Records with nights coded `m`",
            "Records with interior gaps (`-` inside the season)"),
  Records = c(0L, 0L, nrow(miss), nrow(gaps))
) |>
  kable(align = c("l", "r"))
```

```{r}
#| label: tbl-gaps
#| tbl-cap: "Records with unrecorded nights inside their trapping season"
trap_records |>
  filter(n_gap_nights > 0) |>
  transmute(flightID, year = as.character(year), locality,
            season = sprintf("%d\u2013%d", first_doy, last_doy),
            `nights counted` = n_nights, `gap nights` = n_gap_nights,
            `season total` = season_total) |>
  arrange(desc(`gap nights`)) |>
  kable(format.args = list(big.mark = ","))
```

Georgetown Township 1969 is the one record where the gaps dominate: 23
unrecorded nights against 21 counted, over a nominal season of 44 days. The
1980 pooled records were operated on an irregular schedule rather than
interrupted. All other retained records are continuous between their first and
last night.

## Minimum effort for phenology

A season total above 99 moths says nothing about whether the trap was running
long enough, or at the right time, to describe a flight curve. The rule below
matches the requirement to the statistic: for a percentile to be identifiable,
the sampling window has to bracket it. If more than 5% of a season's catch
falls on the first night, the 5th percentile sits at or before that night and
cannot be located; the same argument applies to p95 and the last night.

| Criterion | Threshold | What it protects against |
|-----------|-----------|--------------------------|
| `min_nights` | ≥ 15 counted nights | percentiles interpolated from too few points |
| `min_coverage` | counted nights / season span ≥ 0.7 | holes inside the season displacing the cumulative curve |
| `max_tail_share` | ≤ 0.05 of the catch on the first *and* on the last night | a flight period truncated at either end |

: Effort criteria, set as constants in `R/01_read_trap_data.R` {#tbl-effort-rule}

The criteria are recorded, not enforced: `trap_records` carries `coverage`,
`tail_first`, `tail_last` and the logical `usable_phenology`, so an analysis
that needs only the median can relax them.

```{r}
#| label: tbl-usable
#| tbl-cap: "Records excluded from percentile-based phenology, by reason"
trap_records |>
  mutate(reason = case_when(
    usable_phenology                       ~ "usable",
    n_nights < 15                          ~ "too few nights (< 15)",
    coverage < 0.7                         ~ "too gappy (coverage < 0.7)",
    TRUE                                   ~ "flight period truncated (tail > 5%)"
  )) |>
  count(reason, name = "records") |>
  arrange(desc(records)) |>
  kable(align = c("l", "r"))
```

Effort itself is rarely the problem: coverage is exactly 1 for most records
and the median series runs 25 nights. Truncation is what binds, and it is not
a marginal call in the worst cases — East Allagash 1986 took 77% of its season
on the final night, Round Mountain 1972 71% on the first.

```{r}
#| label: tbl-usable-year
#| tbl-cap: "Usable records per year"
trap_records |>
  group_by(year) |>
  summarise(records = n(), usable = sum(usable_phenology),
            localities = n_distinct(locality[usable_phenology]),
            nights = sum(n_nights[usable_phenology]), .groups = "drop") |>
  mutate(year = as.character(year)) |>
  kable(format.args = list(big.mark = ","))
```

The cost is concentrated in the sparse years. Sensitivity of the tail
threshold, holding the other two criteria fixed:

```{r}
#| label: tbl-tail-sensitivity
#| tbl-cap: "Effect of the tail threshold"
bind_rows(lapply(c(0.05, 0.10, 0.15), function(tt) {
  k <- with(trap_records, n_nights >= 15 & coverage >= 0.7 &
              tail_first <= tt & tail_last <= tt)
  tibble::tibble(`tail threshold` = tt,
                 records    = sum(k),
                 localities = n_distinct(trap_records$locality[k]),
                 years      = n_distinct(trap_records$year[k]),
                 `years lost` = paste(setdiff(unique(trap_records$year),
                                              unique(trap_records$year[k])),
                                      collapse = ", "))
})) |>
  kable(align = c("r", "r", "r", "r", "l"))
```

1987 and 1988 have a single trap each, so one failing record erases the year.
Whether that is acceptable depends on the analysis: a pooled site-year model
loses little, a year-by-year trend loses the two terminal years of the
outbreak collapse.

## Open questions

- **1987 and 1988 have no usable record.** Under the current tail threshold
  both years disappear from the phenology set. Relaxing the threshold to 0.10
  restores 1987 at the cost of admitting series whose flight period is
  probably truncated; the decision belongs with the analysis, not the reader.
- **Weather resolution.** The CaSR reanalysis covers every retained locality
  (all within 6.3 km of a grid cell) and every retained year, but air
  temperature is stored 3-hourly, not hourly, unlike the other four variables.
- **Records excluded by the year window.** 1961–1967 and 2016 are outside the
  reanalysis, and are set aside rather than lost: widening `keep_years` and
  setting `min_season_total` to 0 in `R/01_read_trap_data.R` restores the full
  643-record Maine light-trap set.

# Weather data {#sec-weather}

```{r}
#| label: weather-data
weather_hourly <- readRDS(file.path("data", "processed", "weather_hourly.rds"))

## One summary pass over the nested hourly tables; every weather figure below
## is taken from it.
wsum <- weather_hourly |>
  rowwise() |>
  summarise(
    Location, Year, Latitude, tile,
    hours       = nrow(Weather),
    na_temp     = sum(is.na(Weather$Temp)),
    tt_measured = sum(!is.na(Weather$TT)),
    max_dev     = max(abs(Weather$Temp - Weather$TT), na.rm = TRUE),
    tmin        = min(Weather$Temp),
    tmax        = max(Weather$Temp),
    july_mean   = mean(Weather$Temp[month(Weather$datetime) == 7]),
    season_mean = mean(Weather$Temp[yday(Weather$datetime) %in% 160:230]),
    rain_mm     = sum(Weather$PR0) * 1000,
    wind_mean   = mean(Weather$UVC),
    .groups     = "drop"
  )
```

## Source

Weather comes from the ECCC Canadian Surface Reanalysis (**CaSR v3.2**), a
~10 km rotated-pole reanalysis. The two tiles that cover Maine were downloaded
from the [CaSR download
page](https://hpfx.collab.science.gc.ca/~scar700/rcas-casr/download_CaSR_regions_var_period_fr.html)
into `data/raw/Weather_Maine/` — 60 NetCDF files, 4.1 GB, five variables in
six four-year periods spanning 1968–1991.

| Variable | Kind | Meaning | Units | Stored |
|----------|------|---------|-------|--------|
| `TT` | analysis | air temperature at 1.5 m | °C | every 3 h |
| `PR0` | analysis | quantity of precipitation | m | hourly |
| `FB` | forecast | downward solar flux at surface | W m⁻² | hourly |
| `P0` | forecast | surface pressure | mb | hourly |
| `UVC` | forecast | wind modulus at 10 m | kts | hourly |

: CaSR variables extracted {#tbl-casr-vars}

The reanalysis window is what fixes `keep_years` in the trap selection: 1968 is
the first year with weather, and no trap in 1990 or 1991 was sampled.

## Extraction

`R/02_read_weather_data.R` takes the locality-years from
`data/processed/trap_records.rds`, projects the trap coordinates onto the
rotated-pole grid, assigns each locality to the tile that contains it, and
extracts the grid cell holding the trap for every hour of the calendar year.

```{r}
#| label: tbl-weather-grid
#| tbl-cap: "Trap localities on the CaSR grid"
tibble::tibble(
  Quantity = c("Locality-years extracted", "Localities", "Years",
               "Hourly rows", "Southern tile / northern tile",
               "Distance to the grid cell used (km)"),
  Value = c(
    format(nrow(wsum), big.mark = ","),
    n_distinct(wsum$Location),
    sprintf("%d (%d\u2013%d)", n_distinct(wsum$Year), min(wsum$Year), max(wsum$Year)),
    format(sum(wsum$hours), big.mark = ","),
    ## tiles are named by rlat band: 351-385 is the southern one, 386-420 the northern
    paste(as.integer(table(wsum$tile[!duplicated(wsum$Location)])), collapse = " / "),
    "0.7 to 6.3, median 3.9"
  )
) |>
  kable(align = c("l", "r"))
```

Two details of the file layout needed care.

*The period files are not aligned on calendar years.* Each spans 31 December
13:00 UTC of the year before its period to 31 December 12:00 UTC of its last
year, so the final 11 hours of 1971, 1975, 1979, 1983 and 1987 sit in the
*next* period file. Selecting layers by year within a single file silently
returns 8,749-hour years; the script reads both files for those years.

*The tiles are read separately.* Each locality is extracted from the one tile
that covers it, rather than merging two 35,064-layer rasters per variable and
period.

## The resulting dataset

`data/processed/weather_hourly.rds` (51 MB) holds one row per locality-year
with a nested `Weather` tibble: `datetime` (UTC), `datetime_est` (local
standard time, UTC−5 all year), the interpolated `Temp`, the raw `TT`, and
`PR0`, `FB`, `P0`, `UVC`.

```{r}
#| label: tbl-weather-complete
#| tbl-cap: "Completeness of the extracted series"
tibble::tibble(
  Check = c("Locality-years with a complete calendar year of hours",
            "Hours with no interpolated temperature",
            "Hours with a measured (3-hourly) temperature",
            "Hours missing any of PR0, FB, P0, UVC"),
  Value = c(
    sprintf("%d of %d", sum(wsum$hours %in% c(8760L, 8784L)), nrow(wsum)),
    format(sum(wsum$na_temp), big.mark = ","),
    sprintf("%s (%.0f%% of all hours)", format(sum(wsum$tt_measured), big.mark = ","),
            100 * sum(wsum$tt_measured) / sum(wsum$hours)),
    "0"
  )
) |>
  kable(align = c("l", "r"))
```

### Interpolation of air temperature

Temperature is stored every three hours, so two hours in three are filled with
`chillR::interpolate_gaps_hourly()`, which fits an idealised daily temperature
curve to the daily minima and maxima — hence the latitude argument — and
interpolates the residuals. The measured values are left untouched, which is
worth verifying rather than assuming:

```{r}
#| label: tbl-weather-interp
#| tbl-cap: "Interpolation audit"
tibble::tibble(
  Quantity = c("Largest |interpolated - measured| at a measured hour (°C)",
               "Share of hours interpolated",
               "Temperature range over all locality-years (°C)"),
  Value = c(
    format(max(wsum$max_dev), digits = 3, scientific = TRUE),
    sprintf("%.0f%%", 100 * (1 - sum(wsum$tt_measured) / sum(wsum$hours))),
    sprintf("%.1f to %.1f", min(wsum$tmin), max(wsum$tmax))
  )
) |>
  kable(align = c("l", "r"))
```

The deviation at measured hours is floating-point noise, so `Temp` reproduces
`TT` exactly where `TT` exists and only fills the gaps between.

### Does the weather look like Maine?

```{r}
#| label: tbl-weather-climate
#| tbl-cap: "Climate of the trap localities, as extracted"
tibble::tibble(
  Quantity = c("Mean July temperature (°C)",
               "Mean temperature over the flight window, DOY 160-230 (°C)",
               "Correlation of mean July temperature with latitude",
               "Annual precipitation (mm)",
               "Mean wind speed at 10 m (kts)"),
  Value = c(
    sprintf("%.1f to %.1f", min(wsum$july_mean), max(wsum$july_mean)),
    sprintf("%.1f to %.1f", min(wsum$season_mean), max(wsum$season_mean)),
    sprintf("%.2f", cor(wsum$july_mean, wsum$Latitude)),
    sprintf("%d to %d, median %d", round(min(wsum$rain_mm)), round(max(wsum$rain_mm)),
            round(median(wsum$rain_mm))),
    sprintf("%.1f to %.1f", min(wsum$wind_mean), max(wsum$wind_mean))
  )
) |>
  kable(align = c("l", "r"))
```

Summer temperature falls towards the north, but latitude alone orders the
localities poorly. The coolest Julys belong to coastal and island sites —
Bare Island 17.2 °C at 44.6° N, Georgetown 17.4 °C at 43.8° N, Steuben
17.6 °C at 44.5° N — all cooler than Estcourt, 21.0 °C on the Québec border at
47.4° N. Any temperature-driven model of flight timing therefore has to carry
the maritime gradient, not just latitude.

## Open questions for the weather

- **Temperature is interpolated, not measured, for two hours in three.** Any
  night-level analysis that depends on a specific hour — the temperature at
  dusk, or hours above a flight threshold — inherits the idealised diurnal
  curve, not the reanalysis. Aggregates over a whole night are much less
  exposed than instantaneous values.
- **Local time is a constant offset.** `datetime_est` is UTC−5 all year, so it
  is one hour behind civil time in summer. A series nested by UTC year starts
  at 19:00 on 31 December of the preceding year in local time.
- **Reanalysis, not stations.** CaSR is a model product on a 10 km grid: the
  cell holding a trap is 0.7–6.3 km from it and represents ~100 km², so the
  extracted series describes the neighbourhood, not the trap clearing.

# Modelled phenology from BioSIM {#sec-biosim}

```{r}
#| label: biosim-data
biosim_phenology <- readRDS(file.path("data", "processed", "biosim_phenology.rds"))
biosim_flight    <- readRDS(file.path("data", "processed", "biosim_flight.rds"))
```

Alongside the observed catches, spruce budworm seasonal biology was simulated
at every trap locality-year with **BioSIM**, so that observed and modelled
flight phenology can be compared on the same dates and the same statistic.

## Model and weather

`R/03_biosim_phenology.R` calls the BioSIM Web API through the `BioSIM` R
package, which drives a Java client. For each year one request covers the
localities sampled that year: BioSIM assembles the daily weather for those
coordinates from its own historical station network and runs
`Spruce_Budworm_Biology` on it — Régnière's process-oriented model of spruce
budworm seasonal biology (Régnière 1982, 1983, 1984, 1987, 1990; Régnière &
You 1991), driven by daily minimum, mean and maximum temperature.

This is a second, independent weather source: the CaSR reanalysis of the
previous section is not what BioSIM uses. Every simulated day in this run came
back marked as resting on observed rather than generated weather.

## Parameters

The model takes eight parameters. Seven are left at the server defaults; the
two mortality switches are turned off.

| Parameter | Default | Used here | Meaning |
|-----------|---------|-----------|---------|
| `ApplyMortality` | 1 | **0** | apply natural mortality over the population |
| `ApplyAdultMortality` | 1 | **0** | (undocumented in the model help) |
| `CreateEggs` | 0 | 0 | allow more than one generation |
| `TreeKind` | 0 | 0 | host tree on which the insect develops |
| `LastSampleDate` | *(empty)* | *(unset)* | date of the last field observation |
| `ObservedAI` | 0 | 0 | last observed average instar |
| `Defoliation` | 0 | 0 | (undocumented) |
| `AdultLongevity` | 0 | 0 | (undocumented) |

: BioSIM parameters for `Spruce_Budworm_Biology` {#tbl-biosim-parms}

Mortality is off because the question here is *when* the stages occur, not how
many survive. With mortality on, only a few percent of the cohort reaches the
adult stage, so the flight curve would be estimated from a handful of
individuals: checked at one Maine locality-year, the season totals are
`MaleEmergence + FemaleEmergence` = 100% and `MaleFlight + FemaleFlight` = 97%
with mortality off, against 10% and 6% under the defaults.

`LastSampleDate` and `ObservedAI` are left unset, so the simulations are driven
by weather alone and are not anchored on any field observation of development.

## What the model returns, and what is kept

The output is daily and everything is a **percentage of the simulated cohort**,
quantised in steps of 0.25%. Besides the eight stages (`L2o` … `Adults`) it
reports emergence, flight, oviposition and death, separately for each sex.
Three quantities are kept, each the sum of the two sexes:

| Series | Definition | Kind |
|--------|------------|------|
| `Pupae_mf` | `MalePupae` + `FemalePupae` | stock: percentage present that day |
| `Adult_mf` | `MaleAdult` + `FemaleAdult` | stock: percentage present that day |
| `Flight_mf` | `MaleFlight` + `FemaleFlight` | activity: flight that day |

: Series retained from the model output {#tbl-biosim-series}

Only flight is a daily flux, so only its cumulative curve is a genuine
distribution over the season — and it is the series to compare with trap
catches. The two stocks are weighted by how long individuals remain in the
stage and are kept for context. The sex-specific columns reproduce the
both-sexes columns the model also reports exactly, which is checked at every
run.

The simulation is repeated ten times per locality-year and the runs are
averaged, which smooths the 0.25% quantisation of a single run; percentiles are
unaffected by the choice of averaging over summing.

## The resulting dataset

```{r}
#| label: tbl-biosim-overview
#| tbl-cap: "The BioSIM run at a glance"
tibble::tibble(
  Quantity = c("Locality-years simulated", "Localities", "Years",
               "Runs per locality-year", "Rows of daily output",
               "Days resting on observed weather"),
  Value = c(
    format(nrow(biosim_flight), big.mark = ","),
    n_distinct(biosim_flight$Location),
    sprintf("%d (%d\u2013%d)", n_distinct(biosim_flight$Year),
            min(biosim_flight$Year), max(biosim_flight$Year)),
    sprintf("%d", max(biosim_flight$n_runs)),
    "829,060 raw / 82,906 averaged",
    "all"
  )
) |>
  kable(align = c("l", "r"))
```

```{r}
#| label: tbl-biosim-series-summary
#| tbl-cap: "Modelled phenology, mean over the locality-years"
biosim_phenology |>
  mutate(series = factor(series, levels = c("Pupae_mf", "Adult_mf", "Flight_mf"))) |>
  group_by(series) |>
  ## The range must be taken before `p50` is replaced by its mean: summarise()
  ## evaluates its arguments in order, in the same data mask.
  summarise(`p50 range` = sprintf("%.0f\u2013%.0f", min(p50), max(p50)),
            `p05` = mean(p05), `p25` = mean(p25), `p50` = mean(p50),
            `p75` = mean(p75), `p95` = mean(p95),
            `duration (d)` = mean(duration),
            .groups = "drop") |>
  relocate(`p50 range`, .after = `duration (d)`) |>
  mutate(across(where(is.numeric), ~ round(.x, 1))) |>
  kable(align = c("l", rep("r", 7)))
```

Modelled flight is centred on day of year 190 on average — 9 July — and the
central 90% of it spans a fortnight. Pupation runs about ten days ahead of
flight, and the adult stock peaks slightly after flight because adults persist
once mortality is switched off.

The locality-year medians are spread over six weeks (DOY 169 to 210) and all
227 values differ, so the simulations respond to site and year rather than
returning a regional average: 1974 alone spans DOY 182 to 210 across its 21
localities.

## Cohorts that do not finish

With mortality off, a locality-year should end with essentially the whole
cohort having flown. Not all of them do: part of the cohort fails to complete
development and is dropped, which shows up as a flight season total well below
100%. `flight_complete` marks the locality-years that reach at least 90%.

```{r}
#| label: tbl-biosim-complete
#| tbl-cap: "Completeness of the simulated cohorts"
biosim_flight |>
  mutate(band = case_when(season_pct >= 90 ~ "90% or more (flight_complete)",
                          season_pct >= 50 ~ "50 to 90%",
                          TRUE             ~ "below 50%")) |>
  group_by(band) |>
  summarise(`locality-years` = n(),
            `mean p50` = round(mean(p50), 1),
            `season total, %` = sprintf("%.1f\u2013%.1f", min(season_pct), max(season_pct)),
            .groups = "drop") |>
  arrange(desc(`locality-years`)) |>
  kable(align = c("l", "r", "r", "r"))
```

```{r}
#| label: tbl-biosim-incomplete
#| tbl-cap: "Locality-years whose cohort did not complete flight"
biosim_flight |>
  filter(!flight_complete) |>
  arrange(season_pct) |>
  transmute(Location, Year = as.character(Year),
            `season total, %` = round(season_pct, 1),
            p05 = round(p05, 1), p50 = round(p50, 1), p95 = round(p95, 1)) |>
  kable(align = c("l", "l", "r", "r", "r", "r"))
```

Two features of that list matter. It is concentrated at particular sites rather
than scattered — Clayton Lake is flagged in five separate years — which points
to a site effect rather than to bad years. And the flagged locality-years are
*later*, not earlier, than the complete ones, which is the opposite of what one
would expect if the surviving fraction were the fast developers; it is
consistent instead with incompleteness arising in cold site-years where
development runs late. The flag is carried on all three series and nothing is
dropped: what to do with these locality-years is a decision for the analysis.

## Open questions for the modelled phenology

- **The comparison with the catches is still to be done.** On the three
  locality-years used to test the script, modelled flight ran one to two weeks
  ahead of the observed catch window; whether that holds over the whole run is
  the next question, and the two tables join on locality-year.
- **Three parameters are undocumented.** `ApplyAdultMortality`, `Defoliation`
  and `AdultLongevity` have no description in the model help and no stated
  units. `AdultLongevity = 0` in particular may be what makes the adult stock
  persist to the end of the year.
- **Nothing anchors the simulation to observed development.** The model offers
  `LastSampleDate` and `ObservedAI` for that, and the `Phenology` sheet of the
  source workbook holds average-instar observations — none of them in Maine,
  but the option exists for a validation elsewhere.


# Modelled pupation from bayessbw

```{r}
#| label: pupal-data
pupal_phenology <- readRDS(file.path("data", "processed", "pupal_phenology.rds"))
```

A second, independent model of development was run over the same
locality-years: the Bayesian development model of Studens et al.
([kdis19/bayessbw](https://github.com/kdis19/bayessbw)), driven by the CaSR
hourly temperatures of @sec-weather rather than by BioSIM's own weather.

## What the model is, and what "pupal phenology" means here

`bayessbw` fits Sharpe-DeMichele / Schoolfield development-rate curves to
laboratory rearing data, in a Bayesian hierarchical form: every posterior draw
carries a rate curve for each larval stage plus a lognormal individual-variation
term. The fitted posterior is distributed as the R package
[`sbwFieldPheno`](https://github.com/studensk/sbwFieldPheno), whose `dev_days()`
accumulates development hour by hour through a temperature series and returns
the day of year on which each simulated individual reaches a given stage.

The laboratory data stop at L6, so the model describes the five larval stages
**L2 to L6 only**. Pupation is the end of L6 — the last transition the model
knows about — and the pupal stage itself is not modelled. *Pupal phenology*
here therefore means the **date of pupation**, not the duration of the pupal
stage. This is also how `bayessbw`'s own weather simulations use the package.

`R/04_pupal_phenology.R` runs it with the settings below.

| Setting | Value | Why |
|---------|-------|-----|
| colony | `NB` | six are available (`AB`, `IPQL`, `NB`, `NWT`, `ON`, `QC`); New Brunswick is the nearest field colony to Maine |
| window | 1 March – 31 August | the window used in `bayessbw`'s own simulations |
| posterior draws | 100 | every draw the package holds |
| individuals per draw | 100 | the simulated population |

: Settings for the pupation simulation {#tbl-pupal-settings}

Two things about that table matter more than the rest. `dev_days()` starts
accumulating development on the **first row of weather it is given**, so 1 March
is an assumption about the date of post-diapause L2 emergence, and every
predicted date moves with it. And because all 100 posterior draws are used, the
posterior spread reported below is the model's own uncertainty, not a sampling
error.

## The resulting dataset

```{r}
#| label: tbl-pupal-overview
#| tbl-cap: "Predicted pupation across the 227 locality-years"
tibble::tibble(
  Quantity = c("Locality-years", "Population median (p50), mean",
               "Population median, range",
               "Within-population spread (p05\u2013p95), mean",
               "Posterior interval on the median, mean width"),
  Value = c(
    format(nrow(pupal_phenology), big.mark = ","),
    sprintf("DOY %.1f", mean(pupal_phenology$p50)),
    sprintf("DOY %.0f\u2013%.0f", min(pupal_phenology$p50), max(pupal_phenology$p50)),
    sprintf("%.1f d", mean(pupal_phenology$spread)),
    sprintf("%.1f d", mean(pupal_phenology$post_hi - pupal_phenology$post_lo))
  )
) |>
  kable(align = c("l", "r"))
```

Each simulation returns a matrix of individuals by posterior draws. Sorting a
draw's column turns it into that draw's population distribution, and taking the
median rank by rank across draws gives the **median simulated population** —
the summary `bayessbw` itself uses, written out in full as
`pupal_population.rds`. Percentiles in the table above are percentiles of that
population; the posterior interval is built the other way round, from the
median of each draw, so the two columns separate individual variation from
parameter uncertainty.

They differ by a factor of four: the fitted curves place the population median
to within about two days, while the population itself takes nine days to pupate
from p05 to p95. Individual variation, not parameter uncertainty, dominates.

```{r}
#| label: tbl-pupal-vs-biosim
#| tbl-cap: "Predicted pupation, bayessbw against BioSIM"
biosim_pupae <- biosim_phenology |>
  filter(series == "Pupae_mf") |>
  select(Location, Year, biosim_p50 = p50)

pupal_phenology |>
  select(Location, Year, bayes_p50 = p50) |>
  inner_join(biosim_pupae, by = c("Location", "Year")) |>
  summarise(
    `locality-years` = n(),
    `bayessbw p50, mean` = round(mean(bayes_p50), 1),
    `BioSIM p50, mean`   = round(mean(biosim_p50), 1),
    `difference, median` = round(median(bayes_p50 - biosim_p50), 1),
    `difference, range`  = sprintf("%+.1f to %+.1f", min(bayes_p50 - biosim_p50),
                                   max(bayes_p50 - biosim_p50)),
    `correlation`        = round(cor(bayes_p50, biosim_p50), 2)
  ) |>
  kable(align = c("r", "r", "r", "r", "r", "r"))
```

The two models were built from different data, fitted by different methods and
driven by different weather, and they place pupation within a day of each
other on average. Their disagreement at individual locality-years reaches two
weeks in both directions, which is the more useful number: it is the scale at
which either model's site-level prediction should be trusted.

## Open questions for the pupation model

- **The 1 March start is an assumption, not a measurement.** Anchoring the
  simulation on BioSIM's own L2 emergence date instead would make the two
  models comparable on equal terms and remove one arbitrary choice.
- **One colony of six.** `NB` is used throughout. The other five are one
  configuration change away, and comparing them would show how much the source
  population matters against site and year.
- **The pupal stage is not modelled.** Getting from pupation to flight needs
  either BioSIM or a separate pupal development function; the section below
  measures that interval empirically instead.


# Model versus observed {#sec-compare}

```{r}
#| label: compare-data
model_vs_observed <- readRDS(file.path("data", "processed", "model_vs_observed.rds"))
cmp <- filter(model_vs_observed, comparable)
```

`R/05_compare_model_observed.R` brings the three sides together, one row per
locality-year: the percentiles of the nightly catch curve, the percentiles of
BioSIM's daily flight curve, and the predicted date of pupation from both
models. All of them are percentiles of a curve over the season computed by the
same helper (`R/functions/phenology.R`), so `p50` always means *half of the
season's total has happened by this day*.

## What can be compared

A caught moth is not the same thing as a flying moth: the trap samples adults
that are active, close enough and attracted to light, whereas BioSIM reports the
percentage of the cohort in the flight state. The comparison below is about
**timing**, never about numbers.

Two flags decide which rows carry a meaningful comparison, and both are kept in
the output rather than filtered away.

```{r}
#| label: tbl-comparable
#| tbl-cap: "Locality-years available for comparison"
model_vs_observed |>
  mutate(group = case_when(
    comparable ~ "Comparable: usable season and complete cohort",
    usable_phenology & !flight_complete ~ "Usable season, incomplete modelled cohort",
    !usable_phenology & flight_complete ~ "Truncated trapping season, complete cohort",
    TRUE ~ "Neither"
  )) |>
  group_by(group) |>
  summarise(`locality-years` = n(),
            `median offset, d` = round(median(d_p50), 1),
            .groups = "drop") |>
  arrange(desc(`locality-years`)) |>
  kable(align = c("l", "r", "r"))
```

Only the first group is used for the statistics that follow. The truncated
seasons are shown in the table because their offset is the diagnostic one would
expect: a season that starts too late or ends too early pushes the observed
median towards the middle of the sampled window, and those rows sit further
from the model than the comparable ones.

## Flight

```{r}
#| label: tbl-offsets
#| tbl-cap: "BioSIM flight minus observed catch, over the comparable locality-years"
tibble::tibble(
  Statistic = c("p05 (start)", "p25", "p50 (median)", "p75", "p95 (end)",
                "peak day", "duration p05\u2013p95"),
  `Median offset, d` = c(median(cmp$d_p05),
                         median(cmp$sim_p25 - cmp$obs_p25),
                         median(cmp$d_p50),
                         median(cmp$sim_p75 - cmp$obs_p75),
                         median(cmp$d_p95),
                         median(cmp$d_peak),
                         median(cmp$d_duration))
) |>
  mutate(`Median offset, d` = sprintf("%+.1f", `Median offset, d`)) |>
  kable(align = c("l", "r"))
```

The model is **early, but only just**: half a day to four days, depending on
which part of the curve is compared. That is a much smaller bias than the one
to two weeks suggested by the three locality-years used to test the simulation
script, and it settles the open question left at the end of @sec-biosim.

The offset is largest at the start of the season and smallest at the end, which
is another way of saying that the modelled flight period is longer than the
observed one — 14.7 days against 11.0 between p05 and p95. Part of that is
real and part is an artefact of trapping: a trap that catches nothing on a night
when a few moths fly contributes nothing to the tails of the observed curve.

The average is not the whole story, though.

```{r}
#| label: fig-compare
#| fig-cap: "Modelled and observed median flight date. Left: one point per comparable locality-year, against the 1:1 line. Right: the same offsets against latitude, with an ordinary least-squares fit. Points are shaded by year."
#| fig-width: 9
#| fig-height: 4.2
library(ggplot2)
library(patchwork)

base <- theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(size = 10, face = "plain", hjust = 0),
        legend.position = "none")

lims <- range(c(cmp$obs_p50, cmp$sim_p50))

p_scatter <- ggplot(cmp, aes(obs_p50, sim_p50, colour = Year)) +
  geom_abline(slope = 1, intercept = 0, linetype = "22", colour = "grey45") +
  geom_point(size = 1.9, alpha = 0.85) +
  annotate("text", x = lims[2], y = lims[2], hjust = 1, vjust = 1.8,
           size = 3, colour = "grey45", label = "1:1") +
  scale_colour_viridis_c(option = "mako", end = 0.85, direction = -1) +
  coord_equal(xlim = lims, ylim = lims) +
  labs(title = "Site by site the agreement is loose",
       x = "Observed median catch date (DOY)",
       y = "Modelled median flight date (DOY)")

p_lat <- ggplot(cmp, aes(Latitude, d_p50, colour = Year)) +
  geom_hline(yintercept = 0, colour = "grey45", linetype = "22") +
  geom_point(size = 1.9, alpha = 0.85) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
              colour = "grey20", fill = "grey80", linewidth = 0.6) +
  scale_colour_viridis_c(option = "mako", end = 0.85, direction = -1,
                         guide = guide_colourbar(barheight = 8)) +
  labs(title = "and the model runs early in the south",
       x = "Latitude (\u00b0N)",
       y = "Model minus observed (d)") +
  theme(legend.position = "right")

(p_scatter & base) + (p_lat & base + theme(legend.position = "right"))
```

```{r}
#| label: tbl-compare-structure
#| tbl-cap: "Where the agreement is, and where it is not"
tibble::tibble(
  Comparison = c("All comparable locality-years",
                 "Year means (years with 5 or more localities)",
                 "Within years, across localities",
                 "Offset against latitude"),
  `r` = c(
    cor(cmp$sim_p50, cmp$obs_p50),
    with(cmp |> group_by(Year) |> filter(n() >= 5) |>
           summarise(s = mean(sim_p50), o = mean(obs_p50), .groups = "drop"),
         cor(s, o)),
    cor(residuals(lm(sim_p50 ~ factor(Year), cmp)),
        residuals(lm(obs_p50 ~ factor(Year), cmp))),
    cor(cmp$d_p50, cmp$Latitude)
  )
) |>
  mutate(r = round(r, 2)) |>
  kable(align = c("l", "r"))
```

This is the substantive result of the comparison, and it has two halves.

**The model gets the year right.** Averaged over the localities sampled in a
year, modelled and observed median dates correlate at 0.83: warm early seasons
and cold late ones are reproduced, and the six-week span of median flight dates
across the record is not an artefact of which sites happened to be trapped.

**It does not get the site right.** Once the year is removed, the correlation
across localities within a year collapses to 0.14. Whatever makes one Maine
locality fly earlier than another in the same summer, the model is not
capturing it — and the residual is not noise, because it has a clear spatial
structure: the offset correlates with latitude at 0.49, so the model runs
early in the south and closer to observed in the north.

```{r}
#| label: tbl-compare-sites
#| tbl-cap: "Localities with at least three comparable years and a median offset beyond a week"
cmp |>
  group_by(Location) |>
  filter(n() >= 3) |>
  summarise(years = n(),
            `latitude` = round(first(Latitude), 2),
            `median offset, d` = round(median(d_p50), 1),
            .groups = "drop") |>
  filter(abs(`median offset, d`) > 7) |>
  arrange(`median offset, d`) |>
  kable(align = c("l", "r", "r", "r"))
```

The worst-fitting sites are southern and coastal — the same localities that the
weather section found to have the coolest Julys in the dataset despite their
latitude. A model that is too early there is a model whose development is
running too fast on those temperature series.

## From pupation to flight

The pupal stage is missing from `bayessbw` and the trap sees only adults, so the
interval between them can be measured rather than modelled.

```{r}
#| label: tbl-lag
#| tbl-cap: "Days from predicted pupation to the observed median catch"
tibble::tibble(
  Source = c("bayessbw pupation", "BioSIM pupation"),
  `Median, d` = c(median(cmp$lag_bayes), median(cmp$lag_biosim)),
  `Mean, d`   = c(mean(cmp$lag_bayes), mean(cmp$lag_biosim)),
  `SD, d`     = c(sd(cmp$lag_bayes), sd(cmp$lag_biosim))
) |>
  mutate(across(where(is.numeric), ~ round(.x, 1))) |>
  kable(align = c("l", "r", "r", "r"))
```

Both models put the median catch about a fortnight after pupation, which is
a plausible pupal period plus a few days of adult maturation, and the agreement
between two independent development models on that interval is reassuring. The
standard deviation of six days is the part worth pursuing: if it is structured
by temperature rather than random, it is a pupal development function waiting
to be fitted.

## Open questions for the comparison

- **The spatial residual is the next problem.** Latitude is a proxy, not an
  explanation. The candidates are the weather driving the models (BioSIM's
  station interpolation against the CaSR grid, which the weather section
  already showed disagree on which sites are coolest), the absence of any
  site-level host or shading effect in the models, and immigration — Maine
  catches in these years include moths that flew in from elsewhere, and a
  locality's catch is not necessarily its own cohort.
- **Trap attraction is not modelled at all.** The observed curve is a catch
  curve. Nightly weather gates flight and gates trapping, and neither is in the
  comparison yet; the hourly CaSR variables extracted for every locality-year —
  temperature, wind, precipitation — were extracted for exactly this purpose.
- **The comparable set could be larger.** Relaxing the tail threshold from 5%
  to 10% would return 12 further records and the years 1987 and 1988; whether
  the added truncation is acceptable is a judgement the sensitivity table in
  @sec-traps supports.
